---
title: 'OHI 2016: Prepping Primary Productivity data for Commerical Fishing Pressure Layers '
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohiprep/src/templates/ohi_hdr.html'
pdf_document:
  toc: true
---

#Summary

This script calculates mean annual net primary production values globally from 2003 to 2015. This data is used to standardize pressures of commercial fishing.


#Updates from previous assessment

We were able to incorporate 2 more years of data, 2014 and 2015.

***

#Data Source

**Reference**: Behrenfeld, M.J. and Falkowski, P.G., 1997. Photosynthetic rates derived from satellite‐based chlorophyll concentration. Limnology and oceanography, 42(1), pp.1-20. 

**Downloaded**: July 30, 2016

**Description**:  Monthly Net Primary Production  (mg C / m2 / day)

**Native data resolution**: 0.083 x 0.083 degree grid   

**Time range**: 2002 - 2015, monthly. Only partial data provided for 2002. OHI uses 2003 - 2015.

**Format**:  XYZ format

***
  
# Setup

```{r setup, eval = F, message = F,warning = F}

library(fields)
library(raster)
library(doParallel)
library(foreach)
library(parallel)

source("~/github/ohiprep/src/R/common.R")
source('vgpm_func.R')

#paths

dir_git <- file.path(dir_M,'git-annex/globalprep/prs_fish/v2016/VGPM_primary_productivity')

#unzip all files
tars <- list.files(file.path(dir_M,'git-annex/globalprep/_raw_data/VGPM_primary_productivity/d2016'),full.names=T, pattern = '.tar')

lapply(tars,untar, exdir = file.path(dir_M,'git-annex/globalprep/_raw_data/VGPM_primary_productivity/d2016'))

```

# Create monthly rasters

This code sources a function, `vgpm.raster` that transforms the raw .xyz files into GeoTIFFs.

```{r runFunction, eval = F}

files = list.files(file.path(dir_M,'git-annex/globalprep/_raw_data/VGPM_primary_productivity/d2016'), full.names=TRUE, pattern = '.gz')

registerDoParallel(12)

foreach (file = files) %dopar%{
  
  print(file)
  vgpm.raster(file, w.lon, e.lon, n.lat, s.lat, log = TRUE, 
              color = tim.colors(30))
}


```

# Get mean annual npp

Calculate the mean annual Net Primary Production per year and save as rasters.

```{r, eval = F}

#global ocean raster at 1km

ocean = raster(file.path(dir_M,'model/GL-NCEAS-Halpern2008/tmp/ocean.tif'))


# set mollweide projection
moll_crs = CRS("+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +units=m +no_defs")

registerDoParallel(7)

foreach (i = 2004:2015) %dopar%{
  
  print(i)
  
  l    = list.files(file.path(dir_git,'int/rasterized_rawdata'),pattern=as.character(i),full.names=T)%>%
          stack()%>%
          calc(fun=function(x){mean(x,na.rm=T)})%>%
            projectRaster(.,crs=moll_crs, over = T,filename = paste0(file.path(dir_git,'int/annual_npp/annual_mean_npp_'),i,'.tif',sep=''),overwrite=T)
  
  
  out  = raster::resample(l,ocean,method='ngb',
                 filename=paste0(file.path(dir_git,'int/annual_npp/annual_mean_npp_moll_1km_'),i,'.tif',sep=''),overwrite=T)
  
}

```
***

##Citation information  
Behrenfeld, M.J. and Falkowski, P.G., 1997. Photosynthetic rates derived from satellite‐based chlorophyll concentration. Limnology and oceanography, 42(1), pp.1-20.

Downloaded from http://www.science.oregonstate.edu/ocean.productivity/standard.product.php