---
title: 'OHI: Social Progress Index'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---


#Summary
This script gapfills the social progress index data and formats it for the OHI global assessment.


#Updates from previous assessment
This is the first year of data.

***

#Data Source 
**Reference**:      
 http://www.socialprogressimperative.org/

Stern, S., A. Wares and T. Hellman. 2016. Social Progress Index: 2016 Methodological Report.

**Downloaded**: 9/28/2016

**Description**: Social Progress Index scores and components for countries.

**Time range**: 2016 

***
  
#Methods
```{r}
# load libraries, set directories
library(ohicore)  #devtools::install_github('ohi-science/ohicore@dev')
library(dplyr)
library(stringr)
library(tidyr)

## comment out when knitting
#setwd("globalprep/prs_res_spi/v2016")


### Load FAO-specific user-defined functions
source('../../../src/R/common.R') # directory locations

```

# Social Progress Index data

Organize data and gapfill missing countries that have incomplete data.  This is the first gapfilling step.  The second step will involve gapfilling countries that have no data.  This first level of gapfilling probably has very little error.

```{r}
cats <- read.csv(file.path(dir_M, 'git-annex/globalprep/_raw_data/SocialProgressIndex/d2016/spi_categories.csv')) %>%
  mutate(category = gsub(" ", "", category))
  
spi <- read.csv(file.path(dir_M, 'git-annex/globalprep/_raw_data/SocialProgressIndex/d2016/2016-Results_spi.csv'), check.names=FALSE, stringsAsFactors=FALSE) 

names(spi) <- gsub(" ", "", names(spi))

spi <- spi %>%
  dplyr::select(-CountryCode) %>%
  gather('category', 'score', -Country) %>%
  left_join(cats, by = "category")

```

## Gapfill bhn component
[NOTE: There seems like there must be a better way to do this.]

```{r}

bhn_tmp <- spi %>%
  filter(subcategory == "bhn") %>%
  spread(category, score)

## gapfill Watershed and Sanitation data  
mod <- lm(WaterandSanitation ~ Shelter, data=bhn_tmp, na.action=na.exclude)
summary(mod)
bhn_tmp$WaterandSanitation_pred <-  predict(mod, newdata = select(bhn_tmp, Shelter))

bhn_tmp <- bhn_tmp %>%
  mutate(gapfill = ifelse(is.na(WaterandSanitation) & !is.na(WaterandSanitation_pred), "WaterandSanitation_gf", NA)) %>%
  mutate(WaterandSanitation = ifelse(is.na(WaterandSanitation), WaterandSanitation_pred, WaterandSanitation))

## gapfill Shelter data
mod <- lm(Shelter ~ WaterandSanitation, data=bhn_tmp, na.action=na.exclude)
summary(mod)
bhn_tmp$Shelter_pred <-  predict(mod, newdata = select(bhn_tmp, WaterandSanitation))

bhn_tmp <- bhn_tmp %>%
  mutate(gapfill = ifelse(is.na(Shelter) & !is.na(Shelter_pred), paste(bhn_tmp$gapfill, "Shelter_gf"), gapfill)) %>%
  mutate(Shelter = ifelse(is.na(Shelter), Shelter_pred, Shelter))

## gapfill Nutrition and Basic Medical Care
mod <- lm(NutritionandBasicMedicalCare ~ WaterandSanitation + Shelter, data=bhn_tmp, na.action=na.exclude)
summary(mod)
bhn_tmp$NutritionandBasicMedicalCare_pred <-  predict(mod, newdata = select(bhn_tmp, WaterandSanitation, Shelter))

bhn_tmp <- bhn_tmp %>%
  mutate(gapfill = ifelse(is.na(NutritionandBasicMedicalCare) & !is.na(NutritionandBasicMedicalCare_pred), paste(bhn_tmp$gapfill, "NutritionandBasicMedicalCare_gf"), gapfill)) %>%
  mutate(NutritionandBasicMedicalCare = ifelse(is.na(NutritionandBasicMedicalCare), NutritionandBasicMedicalCare_pred, NutritionandBasicMedicalCare))

## gapfill personal safety
mod <- lm(PersonalSafety ~ WaterandSanitation + Shelter + NutritionandBasicMedicalCare, data=bhn_tmp, na.action=na.exclude)
summary(mod)
bhn_tmp$PersonalSafety_pred <-  predict(mod, newdata = select(bhn_tmp, WaterandSanitation, Shelter, NutritionandBasicMedicalCare))

bhn_tmp <- bhn_tmp %>%
  mutate(gapfill = ifelse(is.na(PersonalSafety) & !is.na(PersonalSafety_pred), paste(bhn_tmp$gapfill, "PersonalSafety_gf"), gapfill)) %>%
  mutate(PersonalSafety = ifelse(is.na(PersonalSafety), PersonalSafety_pred, PersonalSafety))

### 
bhn_gf <- bhn_tmp %>%
  mutate(gapfill = gsub("NA ", "", gapfill)) %>%
  select(-ends_with("pred")) %>%
  gather('category', 'score', -Country, -subcategory, -gapfill)

```


## Gapfill fw component


```{r}

fw_tmp <- spi %>%
  filter(subcategory == "fw") %>%
  spread(category, score)

## gapfill EnvironmentalQuality  
mod <- lm(EnvironmentalQuality ~ AccesstoInformationandCommunications + HealthandWellness, data=fw_tmp, na.action=na.exclude)
summary(mod)
fw_tmp$EnvironmentalQuality_pred <-  predict(mod, newdata = select(fw_tmp, AccesstoInformationandCommunications, HealthandWellness))

fw_tmp <- fw_tmp %>%
  mutate(gapfill = ifelse(is.na(EnvironmentalQuality) & !is.na(EnvironmentalQuality_pred), "EnvironmentalQuality_gf", NA)) %>%
  mutate(EnvironmentalQuality = ifelse(is.na(EnvironmentalQuality), EnvironmentalQuality_pred, EnvironmentalQuality))

## gapfill Access to Basic Knowledge  
mod <- lm(AccesstoBasicKnowledge ~ AccesstoInformationandCommunications + HealthandWellness + EnvironmentalQuality, 
          data=fw_tmp, na.action=na.exclude)
summary(mod)
fw_tmp$AccesstoBasicKnowledge_pred <-  predict(mod, newdata = select(fw_tmp, AccesstoInformationandCommunications, HealthandWellness, EnvironmentalQuality))

fw_tmp <- fw_tmp %>%
  mutate(gapfill = ifelse(is.na(AccesstoBasicKnowledge) & !is.na(AccesstoBasicKnowledge_pred), paste(gapfill, "AccesstoBasicKnowledge_gf", sep=", "), gapfill)) %>%
  mutate(AccesstoBasicKnowledge = ifelse(is.na(AccesstoBasicKnowledge), AccesstoBasicKnowledge_pred, AccesstoBasicKnowledge))

### 
fw_gf <- fw_tmp %>%
  mutate(gapfill = gsub("NA, ", "", gapfill)) %>%
  select(-ends_with("pred")) %>%
  gather('category', 'score', -Country, -subcategory, -gapfill)


```


## Gapfill op component

```{r}

op_tmp <- spi %>%
  filter(subcategory == "op") %>%
  spread(category, score)

## gapfill AccesstoAdvancedEducation  
mod <- lm(AccesstoAdvancedEducation ~ PersonalRights + PersonalFreedomandChoice, data=op_tmp, na.action=na.exclude)
summary(mod)
op_tmp$AccesstoAdvancedEducation_pred <-  predict(mod, newdata = select(op_tmp, PersonalRights, PersonalFreedomandChoice))

op_tmp <- op_tmp %>%
  mutate(gapfill = ifelse(is.na(AccesstoAdvancedEducation) & !is.na(AccesstoAdvancedEducation_pred), "AccesstoAdvancedEducation_gf", NA)) %>%
  mutate(AccesstoAdvancedEducation = ifelse(is.na(AccesstoAdvancedEducation), AccesstoAdvancedEducation_pred, AccesstoAdvancedEducation))

## gapfill PersonalFreedomandChoice  
mod <- lm(PersonalFreedomandChoice ~ AccesstoAdvancedEducation + PersonalRights, data=op_tmp, na.action=na.exclude)
summary(mod)
op_tmp$PersonalFreedomandChoice_pred <-  predict(mod, newdata = select(op_tmp, AccesstoAdvancedEducation, PersonalRights))

op_tmp <- op_tmp %>%
  mutate(gapfill = ifelse(is.na(PersonalFreedomandChoice) & !is.na(PersonalFreedomandChoice_pred), paste(gapfill, "PersonalFreedomandChoice_gf", sep=", "), gapfill)) %>%
  mutate(PersonalFreedomandChoice = ifelse(is.na(PersonalFreedomandChoice), PersonalFreedomandChoice_pred, PersonalFreedomandChoice))

## gapfill ToleranceandInclusion 
mod <- lm(ToleranceandInclusion ~ AccesstoAdvancedEducation + PersonalFreedomandChoice + PersonalRights, data=op_tmp, na.action=na.exclude)
summary(mod)
op_tmp$ToleranceandInclusion_pred <-  predict(mod, newdata = select(op_tmp, AccesstoAdvancedEducation, PersonalFreedomandChoice, PersonalRights))

op_tmp <- op_tmp %>%
  mutate(gapfill = ifelse(is.na(ToleranceandInclusion) & !is.na(ToleranceandInclusion_pred), paste(gapfill, "ToleranceandInclusion_gf", sep=", "), gapfill)) %>%
  mutate(ToleranceandInclusion = ifelse(is.na(ToleranceandInclusion), ToleranceandInclusion_pred, ToleranceandInclusion))

### 
op_gf <- op_tmp %>%
  mutate(gapfill = gsub("NA, ", "", gapfill)) %>%
  select(-ends_with("pred")) %>%
  gather('category', 'score', -Country, -subcategory, -gapfill)


```


## Combine data

```{r}

data_subcategory <- rbind(bhn_gf, op_gf, fw_gf) %>%
  group_by(Country, subcategory) %>%
  summarize(score = mean(score),
            gapfill = paste(unique(gapfill), collapse=", ")) %>%
  mutate(gapfill = gsub("NA, ", "", gapfill)) %>%
  mutate(gapfill = gsub("NA", NA, gapfill)) %>%
  mutate(score = ifelse(score > 100, 100, score),
         score = ifelse(score < 0, 0, score)) %>%
  rename(score_gf = score) %>%
  ungroup()


data_subcategory$subcategory <- mapvalues(data_subcategory$subcategory, 
                              from=c("bhn", "fw", "op"), 
                              to=c("bhn_score", "fw_score", "op_score"))

spi_subcategory <- spi %>%
  filter(subcategory %in% c("bhn_score", "fw_score", "op_score")) %>%
left_join(data_subcategory, by=c("Country", "subcategory")) %>%
  mutate(score = ifelse(is.na(score), score_gf, score)) %>%
  select(Country, score, subcategory, gapfill)

## Average subcategories to get spi score
spi_gf <- spi_subcategory %>%
  group_by(Country) %>%
  summarize(score_gf = mean(score),
            gapfill = paste(unique(gapfill), collapse = ", ")) %>%
  mutate(gapfill = gsub(", NA", "", gapfill)) %>%
    mutate(gapfill = gsub("NA, ", "", gapfill)) %>%
  mutate(gapfill = gsub("NA", NA, gapfill)) %>%
  mutate(subcategory = "spi")

spi_only <- spi %>%
  filter(subcategory == "spi") %>%
  left_join(spi_gf, by = c("Country", "subcategory")) %>%
  mutate(score = ifelse(is.na(score), score_gf, score)) %>%
  select(Country, score, gapfill)

write.csv(spi_only, "int/Country_spi.csv", row.names=FALSE)


```

# Assign countries to OHI regions

```{r}

spi_only <- spi_only %>%
  mutate(Country = ifelse(Country=="C\xf4te d'Ivoire", "Cote d'Ivoire", Country)) 

spi_rgn <- name_2_rgn(df_in = spi_only, 
                       fld_name='Country')

#122 ohi regions with data

```

# Compare to WGI

```{r}

wgi <- read.csv('../../../../ohi-global/eez2016/layers/wgi_all.csv') 

wgi_spi <- wgi %>%
  left_join(spi_rgn)

plot(wgi_spi$resilience_score*100, wgi_spi$score, ylab="Social Progress Index, score", xlab="Worldwide Governance Score")
abline(0,1, col="red")

mod <- lm(score ~ resilience_score, data=wgi_spi)
summary(mod)

```

# Second round of gapfilling 
In this case, UN geopolitical regions and WGI scores are used to estimate regions with no data.

```{r}

rgns_gf_un <- georegions %>%
  left_join(spi_rgn) %>%
  left_join(wgi)

## Compare models to select a gapfilling method
mod1 <- lm(score ~ r2, data = rgns_gf_un)
mod2 <- lm(score ~ r2 + resilience_score, data = rgns_gf_un)

mod3 <- lm(score ~ r1, data = rgns_gf_un)
mod4 <- lm(score ~ r1 + resilience_score, data = rgns_gf_un, na.action="na.exclude")

mod5 <- lm(score ~ resilience_score, data = rgns_gf_un)

mod6 <- lm(score ~ r1 + poly(resilience_score, 2), data = rgns_gf_un, na.action = "na.exclude")
mod7 <- lm(score ~ r1 + poly(resilience_score, 3), data = rgns_gf_un, na.action = "na.exclude")

plot(predict(mod7), rgns_gf_un$score)
abline(0,1)
plot(predict(mod4), rgns_gf_un$score)
abline(0,1)
plot(0:100, predict(mod7, newdata = data.frame(r1=1, resilience_score=seq(0, 1, by=0.01))), type="l")
points(rgns_gf_un$resilience_score*100, rgns_gf_un$score)
plot(0:100, predict(mod6, newdata = data.frame(r1=1, resilience_score=seq(0, 1, by=0.01))), type="l")
points(rgns_gf_un$resilience_score*100, rgns_gf_un$score)


AIC(mod1, mod2, mod3, mod4, mod5, mod6, mod7)

## Estimate missing data and gapfill
mod_gf <- lm(score ~ r1 + poly(resilience_score, 3), data = rgns_gf_un, na.action = na.exclude)
rgns_gf_un$score_pred <- predict(mod_gf, newdata = rgns_gf_un[, c("r1", "resilience_score")])

rgns_gf <- rgns_gf_un %>%
  mutate(gapfill = ifelse(is.na(score) & !is.na(score_pred), "UN georgn & WGI", gapfill)) %>%
  mutate(score = ifelse(is.na(score), score_pred, score)) %>%
  select(rgn_id, score, gapfill) %>%
  mutate(score = ifelse(score > 100, 100, score),
         score = ifelse(score < 0, 0, score))


```

# Uninhabited regions

These regions will receive an NA for their score (when established population is < 100 people). 

```{r}
uninhab <- read.csv('../../../src/LookupTables/rgn_uninhabited_islands.csv') %>%
  filter(is.na(est_population) | est_population < 100)

rgns_gf_uninhab <- rgns_gf %>%
  mutate(score = ifelse(rgn_id %in% uninhab$rgn_id, NA, score)) %>%
  mutate(gapfill = ifelse(rgn_id %in% uninhab$rgn_id, NA, gapfill))

```

# Save final data

```{r}

gf_data <- rgns_gf_uninhab %>%
  select(rgn_id, method = gapfill) %>%
  mutate(gapfill = ifelse(is.na(method), 0, 1))
write.csv(gf_data, "output/spi_gf.csv", row.names=FALSE)


res_data <- rgns_gf_uninhab %>%
  select(rgn_id, resilience_score=score) %>%
  mutate(resilience_score = resilience_score/100)
write.csv(res_data, "output/spi_res.csv", row.names=FALSE)

prs_data <- rgns_gf_uninhab %>%
  select(rgn_id, pressure_score=score) %>%
  mutate(pressure_score = 1 - (pressure_score/100))
write.csv(prs_data, "output/spi_prs.csv", row.names=FALSE)
```

***

