---
title: 'OHI 2017: Sea Level Rise Pressure Layer'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../src/templates/ohi_hdr.html'
pdf_document:
  toc: true
---

# Summary

This layer preparation script does the following for newly available SLR data.
  
- Clips all monthly rasters to the coast using a 3 nautical mile offshore buffer
- Calculates annual mean sea level anomaly rasters from monthly data
- Rescales values from 0 to 1 using the reference point
- Sets to zero all negative values, indicating decreases in mean sea level
- Resamples raster to ~ 1km^2^ and reproject to Molleweide

This process is completed entirely within this script. The raw data is downloaded externally and held on a server at NCEAS. Although the raw data is not provided, this script can be used on the data downloaded from Aviso [here](http://www.aviso.altimetry.fr/en/data/products/sea-surface-height-products/global/msla-mean-climatology.html). You will need to register with Aviso in order to get a username and password for data access.

# Updates from previous assessment

One additional year of data, 2016, was added. In addition we are using a better mask to eliminate cells farther than 3nm offshore. This change requires a rerun of all year within the dataset.

***

# Data

The source data are monthly mean sea level anomalies, in meters. These anomalies are calculated by subtracting the current absolute sea level for each month from the average sea level for that month calculated from 1993 - 2012.

**Reference**: The altimeter products were produced and distributed by Aviso (http://www.aviso.altimetry.fr/), as part of the Ssalto ground processing segment. [AVISO MSLA heights, monthly means](http://www.aviso.altimetry.fr/en/data/products/sea-surface-height-products/global/msla-mean-climatology.html)  
**Downloaded**: August 1, 2017 (for 2016 data)   
**Description**: Monthly mean sea level anomaly (meters above mean sea level)   
**Native data resolution**: 0.25 degree grid cells   
**Time range**: January 1993 - December 2016  
**Format**: NetCDF   

***

# Methods

## Setup

``` {r setup, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/',message = FALSE, warning = FALSE)

#setting up provenance
# devtools::install_github('oharac/provRmd')
# library(provRmd)
# prov_setup()

library(tidyverse)
library(raster)
library(RColorBrewer)
library(maps)
library(parallel)
library(foreach)
library(doParallel)
library(stringr)
library(sf)



#define github file path
 dir_git <- '~/github/ohiprep'

 #setwd("globalprep/prs_slr/v2017")
source('../../../src/R/common.R')

#define server file path
dir_anx <- file.path(dir_M, 'git-annex/globalprep')

#define raw data file path (non-NCEAS folks will not have access to this data)
dir_anx_aviso <- file.path(dir_M, 'git-annex/globalprep/_raw_data/AVISO_slr')

### set up proj4string options: WGS84
p4s_wgs84 <- '+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'

### Define spectral color scheme for plotting maps
cols      = rev(colorRampPalette(brewer.pal(9, 'Spectral'))(255)) # rainbow color scheme

# define mollweide projection
mollCRS <- CRS('+proj=moll')

# Read in ocean raster with cells at 1km. Use this as a template for resampling. This is held on an NCEAS server and folks without access will not be able to use this mask.
ocean = raster(file.path(dir_M, 'git-annex/globalprep/spatial/v2017/ocean.tif'))
```


## Data Prep

### Clip data to coastal cells

All NetCDF files for each month are rasterized.

```{r clip}

#list all netCDF files
nc_files <- c(list.files(file.path(dir_anx_aviso, 'd2017'),
                       full.names = TRUE, pattern = '.nc'),
              list.files(file.path(dir_anx_aviso, 'd2016/msla_monthly_mean'),
                      full.names = TRUE, pattern = '.nc'))
```

The raw monthly data looks like this:

```{r plot_raw}

plot(raster(nc_files[1]), col=cols, axes=F, main = "Sea Level Rise (m) January 2016")

```

The following code is used to:
1. Rasterize each monthly NetCDF file
2. Rotate each raster so that the Atlantic Ocean is centered in the raster

The output is saved in the folder `int/msla_monthly`

```{r, eval=F}

registerDoParallel(10)

## parallel forloop function that rotates each monthly file, sets the long/lat projection, and keeps only coastal cells - saved to GitHub

foreach(file = nc_files) %dopar% {
  
  m_yr <- substr(file,nchar(file)-10,nchar(file)-3)
    
  #read in month raster
   r <- raster(file)%>%
          rotate()

    #define projection of the raster before reprojecting
  projection(r) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
  
writeRaster(r, filename = file.path(dir_M, sprintf('git-annex/globalprep/prs_slr/v2017/int/msla_monthly/msla_monthly_%s.tif', m_yr)), overwrite=TRUE)

}

```


## Annual mean sea level anomalies

Annual mean sea level anomaly rasters are calculated from the monthly data. 

``` {r msla_monthly_to_annual, eval=F}

msla_files <- list.files(file.path(dir_M, 'git-annex/globalprep/prs_slr/v2017/int/msla_monthly'), full.names=TRUE)

## stack all rasters for this year, and calc annual mean, then write as raster

registerDoParallel(6)

foreach(yr = c(1993:2016)) %dopar%{ #yr=2003
  
  files = msla_files[str_detect(msla_files, as.character(yr))]
  
  rast_annual_mean <- stack(files)%>%
                      calc(mean, na.rm = TRUE)%>%
                      writeRaster(filename = file.path(dir_M, 
  sprintf('git-annex/globalprep/prs_slr/v2017/int/msla_annual_mean/msla_annual_%s.tif', yr)), overwrite = TRUE)
}



```

## Changing the projection and masking
Since we are only interested in the increase in sea level near the coasts, we apply a mask to the raster layers that removes all cells farther than 3nm offshore. This mask was created previously for the OHI global 2016 assessment.

```{r plot_mask, eval = FALSE}

# ## 3nm offshore raster to select only nearshore cells.
# poly_3nm <- read_sf(file.path(dir_M, "git-annex/Global/NCEAS-Regions_v2014/data"), "rgn_offshore3nm_mol")
# poly_3nm[duplicated(poly_3nm$rgn_id), ]
# s <- fasterize(poly_3nm, ocean, field="rgn_id")
# s <- calc(s, fun=function(x){ifelse(x>0, 1, NA)}) 
# writeRaster(s, file.path(dir_M, "git-annex/globalprep/prs_slr/v2017/int/ocean_mask.tif"))
s <- raster(file.path(dir_M, "git-annex/globalprep/prs_slr/v2017/int/ocean_mask.tif"))

#s <- raster(file.path('../v2016/int/rast_3nm_mask_fasterize.tif'))
plot(s, col='red')


annual_means <- list.files(file.path(dir_M, 'git-annex/globalprep/prs_slr/v2017/int/msla_annual_mean'), full=TRUE)

foreach(file = annual_means) %dopar%{ # file = annual_means[1]
  
  yr <- str_sub(file,-8,-5)

      rast_data <- raster(file) %>%
      projectRaster(crs = mollCRS, over=TRUE) %>%
      resample(ocean, method = 'ngb', filename = file.path(dir_M, sprintf('git-annex/globalprep/prs_slr/v2017/int/msla_annual_mol/mlsa_annual_mol_%s.tif', yr)), overwrite=TRUE)
      
}


annual_mol <- list.files(file.path(dir_M, 'git-annex/globalprep/prs_slr/v2017/int/msla_annual_mol'), full=TRUE)

foreach(file = annual_mol) %dopar%{ # file=annual_mol[2]

  yr <- str_sub(file,-8,-5)
  
    rast <- raster(file)

    mask(raster(file), s, filename = file.path(dir_M, sprintf('git-annex/globalprep/prs_slr/v2017/int/msla_annual_mol_coastal/msla_annual_mol_coastal_%s.tif', yr)), overwrite=TRUE)

}

plot(raster(file.path(dir_M, 'git-annex/globalprep/prs_slr/v2017/int/msla_annual_mol_coastal/msla_annual_mol_coastal_2010.tif')))
```



## Reference Point

The reference point is the 99.99th quantile of the entire data distribution from 1993 - 2015. (This value has been updated due to changes in the source data, previously was 0.246225 m, currently is 0.3359385 m).

```{r ref, eval=F}

coastal_rasts <- list.files(file.path(dir_M, 'git-annex/globalprep/prs_slr/v2017/int/msla_annual_mol_coastal'), pattern="tif", full.names = TRUE)

#get data across all years to 2015

vals <- c()

for(i in 1993:2015){ # i=1993
  print(i)
  m <- coastal_rasts[which(str_sub(coastal_rasts, -8, -5) == i)] %>%
    raster() %>%
    getValues() %>%
    na.omit()
  
  vals <- c(vals,m)
  
}

ref_point_slr <- quantile(vals, 0.9999)


ref <- read_csv(file.path('../../supplementary_information/v2016/reference_points_pressures.csv')) #grab reference value 
ref$ref_point[ref$pressure == "Sea Level Rise"] <- ref_point_slr
write.csv(ref, "../../supplementary_information/v2016/reference_points_pressures.csv", row.names = FALSE)
  
ref <- read_csv(file.path('../../supplementary_information/v2016/reference_points_pressures.csv')) %>% #grab reference value  from the supp_info csv
       filter(pressure == "Sea Level Rise") %>%
       .$ref_point %>%
        as.numeric()

```     

## Rescale 

Each annual raster is recaled from 0 to 1 using the reference point. If a value is greater than the reference point, it is automatically given a value of 1.

```{r rescale, eval=F}

foreach(file = coastal_rasts) %dopar%{ # file = coastal_rasts[10]
  yr <- str_sub(file, -8,-5)
    raster(file)%>%
      calc(fun=function(x){ifelse(x<0, 0, x)}) %>% #set all negative values to 0
      calc(fun=function(x){ifelse(x>ref, 1, x/ref)}, 
           filename = file.path(dir_M, sprintf('git-annex/globalprep/prs_slr/v2017/output/slr_%s.tif', yr)), overwrite=TRUE) }

```

# Results

```{r plot_results}

r <- raster(file.path(dir_M,'git-annex/globalprep/prs_slr/v2017/output/slr_2016.tif'))
plot(ocean, col='cornsilk2', axes=FALSE, box=FALSE, main='Sea Level Rise Pressure 2016', legend=FALSE)		
plot(r, col=cols, axes=FALSE, box=FALSE, add=TRUE)


r_new <- raster(file.path(dir_anx,'prs_slr/v2017/output/slr_2015.tif'))
r_old <- raster(file.path(dir_anx,'prs_slr/v2016/output/slr_2015.tif'))

rasterVis::histogram(r_old, main ='Sea Level Pressure 2015 old data')
rasterVis::histogram(r_new, main ='Sea Level Pressure 2015 new data')


```


```{r extract regions, eval=FALSE}

# raster/zonal data
slr_loc <- file.path(dir_M, "git-annex/globalprep/prs_slr/v2017/output")

zones <- raster(file.path(dir_M, "git-annex/globalprep/spatial/v2017/regions_eez_with_fao_ant.tif"))  # raster data

rgn_data <- read_sf(file.path(dir_M, 'git-annex/globalprep/spatial/v2017'), 'regions_2017_update') %>%
  st_set_geometry(NULL) %>%
  dplyr::filter(rgn_type == "eez") %>%
  dplyr::select(rgn_id = rgn_ant_id, rgn_name)

# read in raster files
rasts <- list.files(slr_loc, full.names = TRUE)
rasts <- rasts[!grepl(".aux", rasts)]

stack_slr <- stack(rasts)

# extract data for each region:
regions_stats <- zonal(stack_slr,  zones, fun="mean", na.rm=TRUE, progress="text")


regions_stats2 <- data.frame(regions_stats)
setdiff(regions_stats2$zone, rgn_data$rgn_id) # High Seas regions are in there, makes sense....no land
setdiff(rgn_data$rgn_id, regions_stats2$zone)

regions_stats2 <- regions_stats2 %>%
  rename(rgn_id = zone) %>%
  filter(rgn_id <=250) %>%
  gather("year", "pressure_score", -1) %>%
  mutate(year = as.numeric(as.character(substring(year, 5, 8))))

write.csv(regions_stats2, "output/slr.csv", row.names = FALSE)

## visualize data
library(googleVis)

plotData <- regions_stats2%>%
  left_join(rgn_data)%>%
  dplyr::select(rgn_name, year, pressure_score) %>%
  dplyr::arrange(rgn_name, year) %>%
  data.frame()

Motion=gvisMotionChart(plotData, 
                       idvar="rgn_name", 
                       timevar="year")
plot(Motion)

print(Motion, file='slr.html')

```



```{r compare}

new_data <- read.csv("output/slr.csv") %>%
  dplyr::select(rgn_id, year, new_pressure_score = pressure_score)
old <- read.csv("../v2016/output/slr_updated.csv") %>%
  mutate(year = year -1) %>%
  left_join(new_data, by=c("year", "rgn_id"))

plot(old$pressure_score, old$new_pressure_score, ylab="new score", xlab = "old score")
abline(0,1, col="red")


```


### Citation information  

The altimeter products were produced and distributed by Aviso (http://www.aviso.altimetry.fr/), as part of the Ssalto ground processing segment. [AVISO MSLA heights, monthly means](http://www.aviso.altimetry.fr/en/data/products/sea-surface-height-products/global/msla-mean-climatology.html)
