-----

# Provenance

``` {r provenance_info, echo = FALSE, message = FALSE, warning = FALSE}

script_prov_out_df <- script_prov(.prov_parent_script_file) 
```

- _Run ID: `r script_prov_out_df$run_id` (`r script_prov_out_df$run_hash %>% str_sub(1, 7)`); run tag: "`r script_prov_out_df$run_tag`"_
- _Run elapsed time: `r script_prov_out_df$elapsed_time` seconds; run memory usage: `r script_prov_out_df$memory_use` MB_
- _System info: `r script_prov_out_df$msg_sys`_
- _Session info_
    - _`r script_prov_out_df$msg_ses`_
    - _`r script_prov_out_df$msg_base_pkgs`_
    - _`r script_prov_out_df$msg_att_pkgs`_

``` {r provenance_plot,  echo = FALSE, message = FALSE, warning = FALSE}

prov_plot_out <- plot_prov(.script_track)

DiagrammeR::render_graph(prov_plot_out)

```

``` {r cleanup,  echo = FALSE, message = FALSE, warning = FALSE}
  ### delete footers from local prov/ directory
  ftr_list <- list.files(system.file('footer', package = 'provRmd'))
  invisible(
    lapply(ftr_list, function(x) {
      unlink(x)
    })
  )
```

