-----

# Provenance

``` {r provenance_info, echo = FALSE, message = FALSE, warning = FALSE}

script_prov_out_df <- script_prov(.prov_parent_script_file) 
```

- _Run ID: `r script_prov_out_df$run_id` (`r script_prov_out_df$run_hash %>% str_sub(1, 7)`); run tag: "`r script_prov_out_df$run_tag`"_
- _Run elapsed time: `r script_prov_out_df$elapsed_time` seconds; run memory usage: `r script_prov_out_df$memory_use` MB_
- _System info: `r script_prov_out_df$msg_sys`_
- _Session info_
    - _`r script_prov_out_df$msg_ses`_
    - _`r script_prov_out_df$msg_base_pkgs`_
    - _`r script_prov_out_df$msg_att_pkgs`_

``` {r provenance_plot,  echo = FALSE, message = FALSE, warning = FALSE}

prov_plot_out <- plot_prov(.script_track)

DiagrammeR::render_graph(prov_plot_out)

```

``` {r provenance_table,  echo = FALSE, message = FALSE, warning = FALSE}

prov_tbl <- .script_track %>% #[ , 1:10] %>%
  mutate(commit_url  = str_replace(commit_url, 'Previous commit: ', ''),
         commit_url  = ifelse(commit_url == 'no version control info found', NA, commit_url),
         commit_hash = ifelse(is.na(commit_url), NA,
                             sprintf('[%s](%s)', str_sub(commit_url, -40, -34), commit_url)),
         commit_hash_short = ifelse(is.na(commit_url), NA,
                             sprintf('%s', str_sub(commit_url, -40, -34))))

run_id   <- first(prov_tbl$run_id)
run_hash <- first(prov_tbl$run_hash) %>% str_sub(1, 7)
run_tag  <- first(prov_tbl$run_tag)
run_date <- first(prov_tbl$run_date)

prov_tbl1 <- prov_tbl %>%
  mutate(file_name = basename(file_loc),
         file_dir  = dirname(file_loc) %>% str_replace(path.expand('~'), '~')) %>%
  dplyr::select(sequence, file_name, parent_chunk,
                file_dir, filetype, 
                uncomm_chgs = uncommitted_changes, commit_hash_short)

DT::datatable(prov_tbl1 %>% 
                arrange(filetype, file_name),
                caption = sprintf('Provenance summary for run %s (%s): %s (%s)', 
                                  run_id, run_hash, run_tag, run_date),             
              rownames = FALSE,
              class = 'stripe hover compact',
              options  = list(dom = 'tp'))

### DT::datatable creates interactive tables with inactive links; perhaps cut the commit_hash down?
### knitr::kable creates active links, but no interactivity.  Which?
### 
# prov_tbl2 <- prov_tbl %>%
#   mutate(file_name = basename(file_loc),
#          file_dir  = dirname(file_loc) %>% str_replace(path.expand('~'), '~')) %>%
#   dplyr::select(sequence, file_name, parent_chunk,
#                 file_dir, filetype, 
#                 uncomm_chgs = uncommitted_changes, commit_hash)
# 
# knitr::kable(prov_tbl2 %>% 
#                arrange(filetype, file_name),
#              caption = sprintf('Provenance summary for run %s (%s): %s (%s)', run_id, run_hash, run_tag, run_date),             
#              row.names = FALSE)

```

``` {r cleanup,  echo = FALSE, message = FALSE, warning = FALSE}
  ### delete footers from local prov/ directory
  ftr_list <- list.files(system.file('footer', package = 'provRmd'))
  invisible(
    lapply(ftr_list, function(x) {
      unlink(x)
    })
  )
```

