---
title: "OHI 2016 Pressure: Ultraviolet Radiation"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohiprep/src/templates/ohi_hdr.html'
pdf_document:
  toc: true
---

[REFERENCE RMD FILE: https://rawgit.com/OHI-Science/ohiprep/master/globalprep/Pressures_OceanAcidification/v2015/oa_create_layer_2015.html]

#Summary
[general description: What data are being generated? Why (what project, etc.)? Upstream/downstream processing information that might be helpful?  Other information?]

#Updates from previous assessment
[Any significant changes in methods from previous analyses?]

***

#Data Source [NOTE: can be copied from README.md in rawdata file]
**Reference**: [citation for source data; website, literature, contact information. Version of data (if relevant). Screenshots if a series of menus had to be navigated to obtain the data.]

**Downloaded**: [date downloaded or received]

**Description**:  [e.g., surface aragonite state]

**Native data resolution**: [e.g., 1 degree, 30 m, etc.]   

**Time range**: [e.g., 1880-1899, monthly data provided for each year] 

**Format**:  [e.g. NetCDF]

***
  
#Methods

## Setup

```{r setup}

source('~/github/ohiprep/src/R/common.R')
library(raster)
library(ncdf4)
library(dplyr)
library(rhdf5)
library(ggplot2)
library(RColorBrewer)

raw_data_dir = file.path(dir_M,'git-annex/globalprep/_raw_data/NASA_OMI_AURA_UV/d2016')

#years of data we are using for this data layer
yrs <- c(2005:2015)
mths <- c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12")
days <- seq(1,358,7)



#ocean raster at 1km
  ocean = raster(file.path(dir_M, 'model/GL-NCEAS-Halpern2008/tmp/ocean.tif'))
  
#set mollweide projection CRS
  mollCRS=crs('+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +units=m +no_defs')
  
cols      = rev(colorRampPalette(brewer.pal(9, 'Spectral'))(255)) # rainbow color scheme


```

## Create rasters from .nc files and get the mean and standard deviation per month

```{r}


#list all files from raw data folder

files = list.files(file.path(raw_data_dir,'test'),pattern='*.he5$',full.names=T)

attribute <- "/HDFEOS/GRIDS/OMI UVB PRODUCT/Data Fields/ErythemalDailyDose"

he_to_ras <- function(x){
  
  
h <- h5read(x,attribute)
r <- raster(h,xmn=-90,ymn=-180,xmx = 90, ymx = 180, crs = "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")%>%t()%>%flip(direction='y')

r[r<0]<-NA

return(r)
  
}


for (i in yrs){
  
  print(i)
  
    l <- files[substr(files[],96,99)==i]
    
    #now grab weeks
    for (j in 1:length(days)){
      
      print(j)
    
      dys       <- l[days[j]:(days[j]+6)]     #get all files from the year for these 7 days
      rasters   <- lapply(dys[!is.na(dys)],he_to_ras)      #rasterize all of these .nc files. This function will assign NA to the empty ones
    
      uv_week   <- stack(rasters)
  
      #calculate the mean
      week_mean    <-  calc(uv_week, fun=function(x){mean(x,na.rm=T)},
                         filename=paste0(file.path(dir_M),'/git-annex/globalprep/prs_uv/v2016/int/weekly_means/weekly_means_',i,'_',j,'.tif'),
                         overwrite=T)
      
      
      week_sd      <-  calc(uv_week, fun=function(x){sd(x,na.rm=T)},
                         filename=paste0(file.path(dir_M),'/git-annex/globalprep/prs_uv/v2016/int/weekly_sd/weekly_sd_',i,'_',j,'.tif'),
                         overwrite=T)
      
      week_mean_sd <-  overlay(week_mean,week_sd,fun=function(x,y){x+y},
                         filename=paste0(file.path(dir_M),'/git-annex/globalprep/prs_uv/v2016/int/weekly_mean_sd/weekly_mean_sd_',i,'_',j,'.tif'),
                         overwrite=T)
      
    }
}

#get weekly climatologies

names_weekly <- list.files(file.path(dir_M,'git-annex/globalprep/prs_uv/v2016/int/weekly_means'),full.names=T)

for(i in 1:52){
  
  print(i)
  
  #get all rasters for week i
    if(i %in% c(1:9)){
       w = names_weekly[(substr(names_weekly,87,91)==paste0(i,'.tif'))]%>%stack()
    }else{
      w  = names_weekly[(substr(names_weekly,87,88)==i)]%>%stack()
    }
  #mean
  m  = calc(w,fun=function(x){mean(x,na.rm=T)},filename=paste0(file.path(dir_M),
           '/git-annex/globalprep/prs_uv/v2016/int/weekly_climatologies/mean_week_',i,'.tif'),overwrite=T)
  #sd
  sd = calc(w,fun=function(x){sd(x,na.rm=T)},progress='text',filename=paste0(file.path(dir_M),
           '/git-annex/globalprep/prs_uv/v2016/int/weekly_climatologies/sd_week_',i,'.tif'),overwrite=T)
  
  #mean plus sd
  m_sd = overlay(m,sd,fun=function(x,y){x+y},filename = paste0(file.path(dir_M),
           '/git-annex/globalprep/prs_uv/v2016/int/weekly_climatologies/mean_sd_week_',i,'.tif'),overwrite=T)
}
```



Compare each week to the climatology for week (mean + sd)

```{r}

# Second Loop to calculate annual positive anomalies


for (i in 2005:2015){
  
  print(i)
  
  s = stack()
  
  for (j in 1:52){
    
    print(j)

    w_mean = raster(paste0(file.path(dir_M),'/git-annex/globalprep/prs_uv/v2016/int/weekly_means/weekly_means_',i,'_',j,'.tif')) #mean UV for week i

    w_anom = raster(paste0(file.path(dir_M),'/git-annex/globalprep/prs_uv/v2016/int/weekly_climatologies/mean_sd_week_',j,'.tif')) #get the week climatology
    
    count = overlay(w_mean,w_anom,fun=function(x,y){ifelse(x>y,1,0)},progress='text') #compare to average anomaly for that week 
    
    s = stack(s,count)
    
  }
  
  year = calc(s,fun=function(x){sum(x,na.rm=T)},progress='text',filename=paste0('int/annual_pos_anomalies_',i,'.tif'),overwrite=T)
}

```


Look at difference in total number of anomalys in recent 5 years (2011 - 2015) compared to first 5 years (2005-2009)

```{r}

l   <- list.files('int',full.names=T)

ref <- list.files('int',full.names=T)[1:5]%>%stack()%>%calc(.,fun=function(x){sum(x,na.rm=T)})

library(parallel)

registerDoParallel(5)

foreach(i = c(2007:2011)) %dopar% { #i=2005

  print(i)

  yrs <- c(i:(i+4))
  
  s   <- stack(l[substr(l,26,29)%in%yrs])%>%sum(.,na.rm=T)

  diff = s - ref #calculate difference between total number of anomalies in recent and historical (2005-2009) time periods

  projection(diff) <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
  
  out = projectRaster(diff,crs=mollCRS,progress='text',over=T)%>%
         resample(.,ocean,method='ngb',progress='text')%>%
          mask(.,ocean,filename=paste0(dir_M,'/git-annex/globalprep/prs_uv/v2016/output/uv_',min(yrs),'_',max(yrs),'-2005-2009_moll_1km.tif',sep=""),overwrite=T)
  
   resc_num         <- quantile(out,prob=0.9999) # calculate the 99.99th quantile  

  sprintf('Rescaling')

  out_rescale = calc(out,fun=function(x){ifelse(x>0,ifelse(x>resc_num,1,x/resc_num),0)},
                     filename=paste0(dir_M,'/git-annex/globalprep/prs_uv/v2016/output/uv_',min(yrs),'_',max(yrs),'-2005-2009_rescaled_moll_1km.tif',sep=""),overwrite=T)
}

```


#Check out the output

```{r output}

out <- list.files(file.path(dir_M,'git-annex/globalprep/prs_uv/v2016/output'),pattern = '_rescaled_moll_1km.tif',full.names=T)%>%stack()

plot(out,box=F,col=cols,axes=F)

```

***

#Citation information  
Jari Hovila, Antii Arola, and Johanna Tamminen (Oct ), OMI/Aura Surface UVB Irradiance and Erythemal Dose Daily L3 Global 1.0x1.0 deg Grid, version 003, NASA Goddard Space Flight Center, , Accessed [Enter User Data Access Date at] 10.5067/Aura/OMI/DATA3009