---
title: 'OHI: Social Progress Index: Exploring gapfilling'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---


#Summary
This script gapfills the social progress index data and formats it for the OHI global assessment.

# Updates
Previously, only one year of data were available, but now there are 4 years of data. Updated method of gapfilling.  Now use Hmisc's functions to gapfill missing data.  This method is used to gapfill subcategories with 1 or more subsubcategories of data.  We use the mean of 50 permutations using the "regression" method.

***

**Citation**: http://www.socialprogressimperative.org/

Stern, S., A. Wares and T. Epner. 2017. Social Progress Index: 2017 Methodology Report.

**Source information**: 
http://www.socialprogressimperative.org/global-index/ --> Export

**Date**: 8/10/2017

**Time range**: 2014-2017

**Native data resolution**: country scores

**Format**:  Excel file

**Description**: Social Progress Index scores and components for countries.

```{r}
# load libraries, set directories
library(ohicore)  #devtools::install_github('ohi-science/ohicore@dev')
library(dplyr)
library(stringr)
library(tidyr)
library(Hmisc)

## comment out when knitting
#setwd("globalprep/prs_res_spi/v2017")


### Load FAO-specific user-defined functions
source('../../../src/R/common.R') # directory locations

set.seed(227)

```

# Social Progress Index data

Organize data and gapfill missing countries that have incomplete data.  This index is comprised of 3 indicators, which are each comprised of 4 subindicators, which are comprised of several datasets.  If one of the subindicators are missing, the SPI is not calculated.  The first round of gapfilling involves using relationships between the the subindicators to gapfill missing data.  If a region is missing all subindicator data, then a second round of gapfilling is performed using relationships between UN geopolitical regions and the World Governance Indicator to gapfill the SPI score.

The following gets all years of data (currently in separate files).

```{r}
cats <- read.csv(file.path(dir_M, 'git-annex/globalprep/_raw_data/SocialProgressIndex/d2016/spi_categories.csv')) %>%
  mutate(category = gsub(" ", "", category))

files <- list.files(file.path(dir_M, 
          "git-annex/globalprep/_raw_data/SocialProgressIndex/d2017"),
          pattern = "spi_", full = TRUE)

all_spi <- data.frame()

for(spi in files){ #spi = files[1]

spi_data <- read.csv(spi, check.names=FALSE, stringsAsFactors=FALSE) 

names(spi_data) <- gsub(" ", "", names(spi_data))

yr <- substring(basename(spi), 5, 8)

spi_data <- spi_data %>%
  dplyr::select(-CountryCode) %>%
  mutate(year = yr) %>%
  gather('category', 'score', -Country, -year) %>%
  filter(category %in% cats$category) %>%
  left_join(cats, by = "category")

all_spi <- rbind(all_spi, spi_data)

}

```


## Gapfilling: Step 1
In this case, we use relationships between the subindicators to estimate missing data.


### Gapfill bhn indicator 

```{r}

set.seed(227)
bhn_subs <- all_spi %>%
  dplyr::filter(subcategory %in% c("bhn")) %>%
  spread(category, score)

bhn_tmp <- all_spi %>%
  dplyr::filter(subcategory %in% c("bhn_score")) %>%
  select(Country, year, bhn_score=score) %>%
  left_join(bhn_subs, by=c("Country", "year"))


bhn_tmp <- bhn_tmp %>%
rowwise() %>%
  mutate(NA_tot = sum(is.na(NutritionandBasicMedicalCare), is.na(NutritionandBasicMedicalCare), is.na(Shelter),
             is.na(WaterandSanitation)))
hist(bhn_tmp$NA_tot)
table(bhn_tmp$NA_tot)

#md.pattern(select(bhn_tmp, -(1:4))) # mice package

imputes <- 50
bhn_gf <- aregImpute(~ WaterandSanitation + Shelter + NutritionandBasicMedicalCare + PersonalSafety, 
                         data = bhn_tmp, type = "regression", n.impute = imputes)


bhn_gf

# to get mean and sd of all imputations
impute_scores_all <- data.frame()

for (imp in 1:imputes){ #imp = 1
  
  imputed <- impute.transcan(bhn_gf, imputation=imp, data=bhn_tmp, list.out=TRUE,
                           pr=FALSE, check=FALSE)
  
      subcat_data <- data.frame(imputed)
  
      impute_scores <- data.frame(Country = bhn_tmp$Country, 
                                  year = bhn_tmp$year,
                                  imputation = imp)
      
      impute_scores <- cbind(impute_scores, subcat_data)
    
      impute_scores_all <- rbind(impute_scores_all, impute_scores)
    }

impute_scores_summary <- impute_scores_all %>%
  gather("subcategory", "score", -(1:3)) %>%
  group_by(Country, year, subcategory) %>%
  dplyr::summarize(score_predict = mean(score),
            sd_score_predict = sd(score))


bhn_tmp_long <- bhn_tmp %>%
  select(Country, year, bhn_score, NA_tot, NutritionandBasicMedicalCare, 
         PersonalSafety, Shelter, WaterandSanitation) %>%
  gather("subcategory", "score", -(1:4)) %>%
  left_join(impute_scores_summary, by=c("Country", "year", "subcategory"))


summary(bhn_tmp_long)

bhn_data <- bhn_tmp_long %>%
  mutate(score = ifelse(is.na(score) & NA_tot < 4, score_predict, score)) %>%
  group_by(Country, year) %>%
  dplyr::summarize( #bhn_score_old = mean(bhn_score), # used this to test to make sure all is well
            score = mean(score),
            NA_tot = mean(NA_tot)) %>%
  mutate(subcategory = "bhn")


```

### Gapfill op indicator 

```{r}
set.seed(227)
op_subs <- all_spi %>%
  dplyr::filter(subcategory %in% c("op")) %>%
  spread(category, score)

op_tmp <- all_spi %>%
  dplyr::filter(subcategory %in% c("op_score")) %>%
  select(Country, year, op_score=score) %>%
  left_join(op_subs, by=c("Country", "year"))


op_tmp <- op_tmp %>%
rowwise() %>%
  mutate(NA_tot = sum(is.na(AccesstoAdvancedEducation), is.na(PersonalFreedomandChoice), is.na(PersonalRights),
             is.na(ToleranceandInclusion)))
hist(op_tmp$NA_tot)
table(op_tmp$NA_tot)

#md.pattern(select(op_tmp, -(1:4))) # mice package

imputes <- 50
op_gf <- aregImpute(~ AccesstoAdvancedEducation + PersonalFreedomandChoice + PersonalRights + ToleranceandInclusion, 
                         data = op_tmp, type = "regression", n.impute = imputes)


op_gf

# to get mean and sd of all imputations
impute_scores_all <- data.frame()

for (imp in 1:imputes){ #imp = 1
  
  imputed <- impute.transcan(op_gf, imputation=imp, data=op_tmp, list.out=TRUE,
                           pr=FALSE, check=FALSE)
  
      subcat_data <- data.frame(imputed)
  
      impute_scores <- data.frame(Country = op_tmp$Country, 
                                  year = op_tmp$year,
                                  imputation = imp)
      
      impute_scores <- cbind(impute_scores, subcat_data)
    
      impute_scores_all <- rbind(impute_scores_all, impute_scores)
    }

impute_scores_summary <- impute_scores_all %>%
  gather("subcategory", "score", -(1:3)) %>%
  group_by(Country, year, subcategory) %>%
  dplyr::summarize(score_predict = mean(score),
            sd_score_predict = sd(score))


op_tmp_long <- op_tmp %>%
  select(Country, year, op_score, NA_tot, AccesstoAdvancedEducation, 
         PersonalFreedomandChoice, PersonalRights, ToleranceandInclusion) %>%
  gather("subcategory", "score", -(1:4)) %>%
  left_join(impute_scores_summary, by=c("Country", "year", "subcategory"))


summary(op_tmp_long)

op_data <- op_tmp_long %>%
  mutate(score = ifelse(is.na(score) & NA_tot < 4, score_predict, score)) %>%
  group_by(Country, year) %>%
  dplyr::summarize( #op_score_old = mean(op_score), # used this to test to make sure all is well
            score = mean(score),
            NA_tot = mean(NA_tot)) %>%
  mutate(subcategory = "op")


```


### Gapfill fw indicator 

```{r}
set.seed(227)
fw_subs <- all_spi %>%
  dplyr::filter(subcategory %in% c("fw")) %>%
  spread(category, score)

fw_tmp <- all_spi %>%
  dplyr::filter(subcategory %in% c("fw_score")) %>%
  select(Country, year, fw_score=score) %>%
  left_join(fw_subs, by=c("Country", "year"))


fw_tmp <- fw_tmp %>%
rowwise() %>%
  mutate(NA_tot = sum(is.na(AccesstoBasicKnowledge), is.na(AccesstoInformationandCommunications),
                      is.na(EnvironmentalQuality),is.na(HealthandWellness)))
hist(fw_tmp$NA_tot)
table(fw_tmp$NA_tot)

#md.pattern(select(fw_tmp, -(1:4))) # mice package

imputes <- 50
fw_gf <- aregImpute(~ AccesstoBasicKnowledge + AccesstoInformationandCommunications + EnvironmentalQuality + HealthandWellness, 
                         data = fw_tmp, type = "regression", n.impute = imputes)


fw_gf

# to get mean and sd of all imputations
impute_scores_all <- data.frame()

for (imp in 1:imputes){ #imp = 1
  
  imputed <- impute.transcan(fw_gf, imputation=imp, data=fw_tmp, list.out=TRUE,
                           pr=FALSE, check=FALSE)
  
      subcat_data <- data.frame(imputed)
  
      impute_scores <- data.frame(Country = fw_tmp$Country, 
                                  year = fw_tmp$year,
                                  imputation = imp)
      
      impute_scores <- cbind(impute_scores, subcat_data)
    
      impute_scores_all <- rbind(impute_scores_all, impute_scores)
    }

impute_scores_summary <- impute_scores_all %>%
  gather("subcategory", "score", -(1:3)) %>%
  group_by(Country, year, subcategory) %>%
  dplyr::summarize(score_predict = mean(score),
            sd_score_predict = sd(score))


fw_tmp_long <- fw_tmp %>%
  select(Country, year, fw_score, NA_tot, AccesstoBasicKnowledge, 
         AccesstoInformationandCommunications, EnvironmentalQuality, HealthandWellness) %>%
  gather("subcategory", "score", -(1:4)) %>%
  left_join(impute_scores_summary, by=c("Country", "year", "subcategory"))


summary(fw_tmp_long)

fw_data <- fw_tmp_long %>%
  mutate(score = ifelse(is.na(score) & NA_tot < 4, score_predict, score)) %>%
  group_by(Country, year) %>%
  dplyr::summarize( #fw_score_old = mean(fw_score), # used this to test to make sure all is well
            score = mean(score),
            NA_tot = mean(NA_tot)) %>%
  mutate(subcategory = "fw")


```



## Combine data to get Social Progress Index
The next step averages the 3 indicators and identifies gapfilling.  

```{r}

spi_calc <- rbind(bhn_data, op_data, fw_data) %>%
  mutate(gapfill = paste(subcategory, NA_tot, sep = "_")) %>%
  mutate(score = ifelse(score > 100, 100, score),
         score = ifelse(score < 0, 0, score)) %>%
  group_by(Country, year) %>%
  dplyr::summarize(score = mean(score),
            method = paste(unique(gapfill), collapse=", "),
            gapfill = sum(NA_tot)) %>%
  mutate(method = ifelse(gapfill == 0, NA, method),
         gapfill = ifelse(gapfill >= 1, 1, 0)) %>%
  data.frame()

# check:
check <- spi_data %>%
  filter(category=="SocialProgressIndex") %>%
  dplyr::select(Country, year, true_score=score) %>%
  left_join(spi_calc)
plot(check$true_score, check$score)
abline(0,1, col="red")


write.csv(spi_calc, "int/Country_spi.csv", row.names=FALSE)


```

## Assign countries to OHI regions

```{r}

spi_calc <- read.csv("int/Country_spi.csv", stringsAsFactors = FALSE)
spi_calc <- spi_calc %>%
  mutate(Country = ifelse(Country=="C<f4>te d'Ivoire", "Cote d'Ivoire", Country),
         Country = ifelse(Country=="R<e9>union", "Reunion", Country),
         Country = ifelse(Country=="Cura<e7>ao", "Curacao", Country),
         Country = ifelse(Country=="Korea, Democratic Republic of", "North Korea", Country),
         Country = ifelse(Country=="St. Helena", "Saint Helena", Country),
         Country = ifelse(Country=="St. Martin", "Northern Saint-Martin", Country))

# Channel Islands are Jersey and Guernsey, but these are already in the data

        
spi_rgn <- name_2_rgn(df_in = spi_calc, 
                       fld_name='Country',
                      flds_unique = c("year"))

# Weight the following duplicates by population
# Northern Mariana Islands and Guam
# China, Macao, Hong Kong
# Puerto Rico and Virgin Islands (U.S.)
# Guadeloupe and Martinique

pop_weights <- data.frame(Country = c("China", "Hong Kong", "Macao",
                 "Guadeloupe", "Martinique",
                 "Guam", "Northern Mariana Islands",
                 "Puerto Rico", "Virgin Islands (U.S.)"),
                 pop = c(1389315824, 7184000, 614500,
                         404394, 399637,
                         159358, 53833,
                         3411000, 102951))
spi_rgn <- spi_rgn %>%
  left_join(pop_weights, by = "Country") %>%
  mutate(pop = ifelse(is.na(pop), 1, pop)) %>%
  group_by(rgn_id, rgn_name, year) %>%
  dplyr::summarize(score = weighted.mean(score, pop, na.rm=TRUE),
            method = paste(unique(method), collapse=" "),
            gapfill = weighted.mean(gapfill, pop, na.rm=TRUE))%>%
  dplyr::mutate(year = as.numeric(as.character(year)))

```

# Compare to WGI

WGI is a couple years behind the SPI, so we will use 2015 WGI for the 2016 and 2017 SPI.  There is a strong correlation betwen the WGI and SPI indicators.
```{r}

wgi <- read.csv('../../prs_res_wgi/v2017/output/wgi_res.csv') %>%
  select(rgn_id, year, wgi_score =resilience_score)

length(unique(wgi$rgn_id))

wgi_2016 <- wgi %>%
  filter(year == 2015) %>%
  mutate(year = 2016)

wgi_2017 <- wgi %>%
  filter(year == 2015) %>%
  mutate(year = 2017)

wgi <- rbind(wgi, wgi_2016, wgi_2017)

wgi_spi <- wgi %>%
  left_join(spi_rgn, by=c("rgn_id", "year"))

plot(wgi_spi$wgi_score*100, wgi_spi$score, ylab="Social Progress Index, score", xlab="Worldwide Governance Score")
abline(0,1, col="red")

mod <- lm(score ~ wgi_score, data=wgi_spi)
summary(mod)

```

# Second round of gapfilling 
In this case, UN geopolitical regions and WGI scores are used to estimate regions with no data.  Based on this analysis, a model that includes third degree polynomial of WGI data and r1 UN geopolitical regions is the best model to predict missing SPI values. 

```{r}

years <- data.frame(year = min(spi_rgn$year):max(spi_rgn$year))

rgns_gf_un <-  georegions %>%
  merge(years) %>%
  left_join(spi_rgn) %>%
  left_join(wgi) %>%
  mutate(r2 = as.factor(r2)) %>%
  mutate(r1 = as.factor(r1))


## Compare models to select a gapfilling method
mod1 <- lm(score ~ r2, data = rgns_gf_un, na.action="na.exclude")
mod2 <- lm(score ~ r2 + wgi_score, data = rgns_gf_un, na.action="na.exclude")

mod3 <- lm(score ~ r1, data = rgns_gf_un)
mod4 <- lm(score ~ r1 + wgi_score, data = rgns_gf_un, na.action="na.exclude")

mod5 <- lm(score ~ wgi_score, data = rgns_gf_un)

mod6 <- lm(score ~ r1 + poly(wgi_score, 2), data = rgns_gf_un, na.action = "na.exclude")
mod7 <- lm(score ~ r1 + poly(wgi_score, 3), data = rgns_gf_un, na.action = "na.exclude")

plot(predict(mod2), rgns_gf_un$score)
abline(0,1)
plot(predict(mod4), rgns_gf_un$score)
abline(0,1)


AIC(mod1, mod2, mod3, mod4, mod5, mod6, mod7)

## Estimate missing data and gapfill
## need to add this because some R2 regions have no data
r2_regions <- unique(rgns_gf_un$r2[!is.na(rgns_gf_un$score)])
rgns_gf_un$r2 <- ifelse(rgns_gf_un$r2 %in% r2_regions, rgns_gf_un$r2, NA)

r1_regions <- unique(rgns_gf_un$r1[!is.na(rgns_gf_un$score)])
rgns_gf_un$r1 <- ifelse(rgns_gf_un$r1 %in% r1_regions, rgns_gf_un$r1, NA)


mod_gf_r2 <- lm(score ~ r2 + wgi_score, data = rgns_gf_un, na.action = na.exclude)
rgns_gf_un$score_pred_r2 <- predict(mod_gf_r2, newdata = rgns_gf_un[, c("r2", "wgi_score")])

mod_gf_r1 <- lm(score ~ r1 + wgi_score, data = rgns_gf_un, na.action = na.exclude)
rgns_gf_un$score_pred_r1 <- predict(mod_gf_r1, newdata = rgns_gf_un[, c("r1", "wgi_score")])

mod_gf_wgi <- lm(score ~ wgi_score, data = rgns_gf_un, na.action = na.exclude)
rgns_gf_un$score_pred_wgi <- predict(mod_gf_wgi, newdata = data.frame(wgi_score = rgns_gf_un$wgi_score))


rgns_gf <- rgns_gf_un %>%
  mutate(method = ifelse(is.na(score), "UN georgn & WGI", method)) %>%
  mutate(gapfill = ifelse(is.na(score), "1", gapfill)) %>%
  mutate(score = ifelse(is.na(score), score_pred_r2, score)) %>%
  mutate(score = ifelse(is.na(score), score_pred_r1, score)) %>%
  mutate(score = ifelse(is.na(score), score_pred_wgi, score)) %>%
  select(rgn_id, year, score, method, gapfill) %>%
  mutate(score = ifelse(score > 100, 100, score),
         score = ifelse(score < 0, 0, score))

summary(rgns_gf)  # should be no NA values
length(unique(rgns_gf$rgn_id)) # should be 220 regions

```

# Uninhabited regions

These regions will receive an NA for their score (when established population is < 100 people). 

```{r}
uninhab <- read.csv('../../../src/LookupTables/rgn_uninhabited_islands.csv') %>%
  filter(is.na(est_population) | est_population < 100)

rgns_gf_uninhab <- rgns_gf %>%
  mutate(score = ifelse(rgn_id %in% uninhab$rgn_id, NA, score)) %>%
  mutate(gapfill = ifelse(rgn_id %in% uninhab$rgn_id, NA, gapfill))

```

# Save final data

```{r}

gf_data <- rgns_gf_uninhab %>%
  select(rgn_id, year, gapfill, method) %>%
  mutate(gapfill = ifelse(is.na(method), 0, 1))
write.csv(gf_data, "output/spi_gf.csv", row.names=FALSE)


res_data <- rgns_gf_uninhab %>%
  select(rgn_id, year, resilience_score=score) %>%
  mutate(resilience_score = resilience_score/100)
write.csv(res_data, "output/spi_res.csv", row.names=FALSE)

prs_data <- rgns_gf_uninhab %>%
  select(rgn_id, year, pressure_score=score) %>%
  mutate(pressure_score = 1 - (pressure_score/100))
write.csv(prs_data, "output/spi_prs.csv", row.names=FALSE)

```

***

