---
title: 'OHI: Clean Waters/Trash trend'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

[REFERENCE RMD FILE: https://cdn.rawgit.com/OHI-Science/ohiprep/master/globalprep/CW_trend_trash/v2016/cw_trend_trash_dataprep.html]

#Summary
We describe how we calculated global trash trends for the clean water goal of the OHI 2016 assessment.


#Updates from previous assessment
Previously, we used trends in coastal population as a proxy for trash trends.  This new data describes trends in mismanaged plastic waste for 2010 and projected for 2025.

***

#Data Source 
**Reference**: Jambeck et al. 2016. Plastic waste inputs from land into the ocean. Science 347:768-771.

**Description**:  Describes (Supplementary data) mismanaged plastic waste in 2010 and estimates mismanaged plastic waste in 2025. Using these data, we estimate trash trends during a 5 year period.

**Native data resolution**: Country scale   

**Format**:  Excel table in Supplementary data.

***
  
#Methods

```{r}
## load necessary packages and set up path directories
library(dplyr)
library(tidyr)
# library(devtools)
# devtools::install_github("ohi-science/ohicore@dev") 
library(ohicore)

# setwd('globalprep/cw_trend_trash/v2016') #comment out when knitting

source('../../../src/R/common.R')


```

Raw data is on Mazu (NCEAS private server):
```{r}

## get and format data:
trash <- read.csv(file.path(dir_M, 'git-annex/globalprep/_raw_data/MarinePlastics_Jambeck/d2016/1260352_SupportingFile_Suppl_modified.csv')) %>%
  select(Country, mpw_2010 = Mismanaged_plastic_waste_in_2010, mpw_2025=Mismanaged_plastic_waste_in_2025) %>%
  mutate(mpw_2010 = as.character(mpw_2010)) %>%
  mutate(mpw_2010 = gsub(",", "", mpw_2010)) %>%
  mutate(mpw_2010 = as.numeric(mpw_2010)) %>%
  mutate(mpw_2025 = as.character(mpw_2025)) %>%
  mutate(mpw_2025 = gsub(",", "", mpw_2025)) %>%
  mutate(mpw_2025 = as.numeric(mpw_2025)) %>%
  gather("mpw", "value", starts_with("mpw"))
  
```

Covert country names to OHI regions:

```{r}
## fix a few regions that are reported as one (these will be given the same trend)
# Dhekelia: Small British overseas territory on Cyprus
# 

# Channel Islands: Jersey and Guernsey (also reported individually...delete)
trash <- trash %>%
  filter(!(Country %in% c("Channel Islands")))

antilles <- data.frame(Country = "Netherlands Antilles", country2 = c("Bonaire", "Sint Eustatius", "Saba"))%>% # already included: Curacao, Sint Maarten)
  mutate(country2 = as.character(country2)) %>%
  mutate(Country = as.character(Country))

trash_country_mod <- trash %>%
  left_join(antilles, by="Country") %>%
  mutate(country2 = ifelse(is.na(country2), Country, country2)) %>%
  select(Country=country2, mpw, value)

trash_rgn <- name_2_rgn(df_in = trash_country_mod, 
                       fld_name = 'Country',
                       flds_unique = 'mpw')


### For duplicate regions, weight by region area
weights <- data.frame(Country = c("Puerto Rico", "USVI", 
                         "Northern Mariana Islands", "Guam",
                         "China", "Hong Kong", "Macao",
                         "Guadeloupe", "Martinique"),
             rgn_name =c("Puerto Rico and Virgin Islands of the United States", "Puerto Rico and Virgin Islands of the United States",
                         "Northern Mariana Islands and Guam", "Northern Mariana Islands and Guam",
                         "China", "China", "China",
                         "Guadeloupe and Martinique", "Guadeloupe and Martinique"),
             weight = c(3515, 134, 179, 210, 3705000, 426, 11, 629, 436))

trash_rgn <- trash_rgn %>%
  left_join(weights, by=c("Country", "rgn_name")) %>%
  mutate(weight = ifelse(is.na(weight), 1, weight)) %>%
  group_by(rgn_id, rgn_name, mpw) %>%
  summarize(value = weighted.mean(value, weight)) %>%
  data.frame()

```

Gapfilling missing data using a linear model with population data as a predictor:

```{r}
population <- read.csv('../../le_world_bank/v2016/output/population_2016.csv') %>%
  select(rgn_id, count)

trash_gf <- spread(trash_rgn, mpw, value)

trash_gf <- population %>%
  left_join(trash_gf, by="rgn_id")

### Predicting the 2010 data:
plot(log(mpw_2010+1) ~ log(count+1), data=trash_gf)
mod_2010 <- lm(log(mpw_2010+1) ~ log(count+1), data=trash_gf)
summary(mod_2010)

trash_gf$predict_ln_2010 <- predict(mod_2010, newdata = trash_gf)
trash_gf$predict_2010 <- exp(predict(mod_2010, newdata = trash_gf))


### Predicting the 2025 data:
plot(log(mpw_2025+1) ~ log(count+1), data=trash_gf)
mod_2025 <- lm(log(mpw_2025+1) ~ log(count+1), data=trash_gf)
summary(mod_2025)

trash_gf$predict_ln_2025 <- predict(mod_2025, newdata = trash_gf)
trash_gf$predict_2025 <- exp(predict(mod_2025, newdata = trash_gf))


## save corresponding gapfilling dataset:
gf_trash_data <- trash_gf %>%
  mutate(gapfilled = ifelse(is.na(mpw_2010), 1, 0)) %>% 
  mutate(method = ifelse(gapfilled == 1, "lm_population", NA)) %>%
  mutate(error = NA) %>%
  select(rgn_id, gapfilled, method, error)

write.csv(gf_trash_data, 'output/cw_trash_trend_gf.csv', row.names=FALSE)


## get gapfilled values for 2010 and 2025 estimates:
trash_trend_data <- trash_gf %>%
  mutate(mpw_2010 = ifelse(is.na(mpw_2010), predict_2010, mpw_2010)) %>% 
  mutate(mpw_2025 = ifelse(is.na(mpw_2025), predict_2025, mpw_2025)) %>% 
  mutate(mpw_2010 = ifelse(mpw_2010<1, 0, mpw_2010)) %>% 
  mutate(mpw_2025 = ifelse(mpw_2025<1, 0, mpw_2025)) %>%
  select(rgn_id, mpw_2010, mpw_2025)

```

Final calculations.  The data are presented for 2010 and 2025.  We calculate the proportional change over this 15 year period as:

Pchange_15year = (trash_2025 - trash_2010)/trash_2010

To determine the change per year we divide the Pchange_15year by 15 years.  To calculate trend over a 5 year period, this value is multiplied by 5.
```{r}
## Final calculations:
trash_trend <-  trash_trend_data %>%
  mutate(trend_15yr = (mpw_2025 - mpw_2010)/mpw_2010) %>%
  mutate(trend_5yr = trend_15yr/15*5) %>%
  mutate(trend_5yr = ifelse(trend_5yr > 1, 1, trend_5yr)) %>%
  mutate(trend_5yr = ifelse(trend_5yr < -1, -1, trend_5yr)) %>%
  select(rgn_id, trend = trend_5yr)
write.csv(trash_trend, 'output/cw_trash_trend.csv', row.names=FALSE)

```
