---
title: "OHI 2017 - Food Provision: Calculating fisheries stock status"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohiprep/src/templates/ohi_hdr.html'
pdf_document:
  toc: true
---

To calculate the status of global fish stocks from the Sea Around Us catch data we use catch-MSY (CMSY) developed by [Martell and Froese (2012)](http://onlinelibrary.wiley.com/doi/10.1111/j.1467-2979.2012.00485.x/full). This model requires catch and resilience information for each stock. Stocks are defined by FAO areas and are limited to only those records where catch is reported at the species level.

The [`datalimited`](https://github.com/datalimited/datalimited) R package, developed by Sean Anderson, is used to run CMSY.

# Setup 

To install the `datalimited` package:

```{r install_datalimited, eval=F}
# install.packages("devtools")
# devtools::install_github("datalimited/datalimited")
```

```{r setup, message=F, warning=F}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/',message = FALSE, warning = FALSE)

library(datalimited) #has the 4 catch only models
library(tidyverse)
library(doParallel)
registerDoParallel(cores = 8)

source('~/github/ohiprep/src/R/common.R')

```

# Load catch data

Read in catch data aggregated from half degree cell to OHI region [catch_data_prep.Rmd](https://cdn.rawgit.com/OHI-Science/ohiprep/master/globalprep/fis/v2017/catch_data_prep.html).

```{r catch_data}

catch <- read_csv('data/stock_catch.csv') %>%
          rename(common = taxon_common_name) %>%
          select(-X1)

```

# Catch-MSY

```{r cmsy, eval=F}

if (!file.exists(file.path(dir_M,"git-annex/globalprep/fis/v2017/int/cmsy-fits.rds"))) {
  
cmsy_fits <- plyr::dlply(catch, c("stock_id", "common"), function(x) {
  
    #make sure the data is ordered from 1950 to 2010
    x <- arrange(x,year)
    out <- cmsy(ct = x$tons, yr = x$year,  start_r = resilience(x$Resilience[1]), 
      reps = 2e4)
    out$year <- x$year
    out
  }, .parallel = TRUE)
saveRDS(cmsy_fits, file = file.path(dir_M,"git-annex/globalprep/fis/v2017/int/cmsy-fits.rds"))
} else {
  cmsy_fits <- readRDS(file.path(dir_M,"git-annex/globalprep/fis/v2017/int/cmsy-fits.rds"))
}

fake_data <- data.frame(bbmsy_q2.5 = NA, bbmsy_q25 = NA, bbmsy_q50 = NA, 
  bbmsy_q75 = NA, bbmsy_q97.5 = NA)

cmsy_bbmsy <- plyr::ldply(cmsy_fits, function(x) {
  bbmsy_cmsy <- x$biomass[, -1] / x$bmsy
  bbmsy_out <- tryCatch({
    bbmsy_out <- summarize_bbmsy(bbmsy_cmsy)
    bbmsy_out$year <- x$year
    bbmsy_out}, error = function(e) fake_data)
})
cmsy_bbmsy$model <- "CMSY"

write.csv(cmsy_bbmsy,file='data/cmsy_bbmsy.csv')

```

Explore why there are some NAs (I think non convergance)

```{r nas, eval=F}

nas <- cmsy_bbmsy%>%
  group_by(stock_id)%>%
  summarize(m = mean(bbmsy_mean))%>%
  filter(is.na(m))

nrow(nas)

```

## Results

Looking at mean bbmsy for 2014

```{r}

results <- read_csv('data/cmsy_bbmsy.csv') %>%
            filter(year == 2014)
hist(results$bbmsy_mean)



```

Why are so many around 0.4?

```{r}

stocks_0.4 <- results %>%
              filter(bbmsy_mean < 0.5,
                     bbmsy_mean > 0.3)

```

[Anderson, S.C., Cooper, A.B., Jensen, O.P., et al. (2017) Improving estimates of population status and trend with superensemble models. Fish and Fisheries.](http://onlinelibrary.wiley.com/doi/10.1111/faf.12200/full)







