---
title: "OHI 2017 - Food Provision: Preparing catch data for BBmsy calculations and mean catch (weights in FIS model) and FP weights"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohiprep/src/templates/ohi_hdr.html'
pdf_document:
  toc: true
---

# Summary

This script takes the Sea Around Us Project (SAUP) catch data, provided at a resolution of half-degree cells globally, and aggregates catch to **stock levels**. For the Ocean Health Index, we assume a stock is represented by the FAO region in which the species is caught.

In order to aggregate to FAO regions, we associate each cell to the FAO region and the OHI region in which it is located.

An example of our aggregation proces: New Zealand is located entirely in FAO region 81. All catch reported by New Zealand will be aggregated by species to the FAO region. If a species was reported as caught in both New Zealand waters and in the High Seas of area 81, these two records will be combined into one by summing the catch.

## Updates from previous assessment
Catch data now goes from 1950 - 2014. Previously catch data was only available through 2010.

***

# Data
The Sea Around Us Project shared the spatialized catch data with OHI on joined to a lookup table that links SAUP region names and ids to the FAO region they are located in. The proportional area of each EEZ within the FAO region was also calculated for overlapping EEZs.


Reference: 

* Downloaded: June 22, 2017
* Description: Tons per half degree cell with information on sector type, industry type, fishing entity, reporting status and taxonomic information
* Native data resolution: 0.5 degree grid cells
* Time range: 1950 - 2014
* Format: R data files (.rds)

***

## Setup

``` {r setup, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/',message = FALSE, warning = FALSE)

## Libraries
library(readr)
library(dplyr)
library(raster)
library(parallel)
library(purrr)
library(tidyr)

source('~/github/ohiprep/src/R/common.R')

## Paths for data
path_data = file.path(dir_M, "git-annex/globalprep/_raw_data/SAUP/d2017/annual_data")
```

## Load Data

These files are large so using the data.table package is recommended due to R memory limitations.

```{r load_data, eval=F}

## information about half degree cells

cells <- read_csv('cells.csv') %>%
  select(-X1)
```

***

# Aggregate catch

Aggregate catch per OHI region and FAO area. This catch will be used twice. 

(1) The catch is used to weight scores per region. For this we need to use all catch records, including those not reported at the species level.

(2) The catch data at species level is used to calculate stock status (BBmsy) per stock (remember that our definition of a stock is a species caught within a single FAO area).

```{r, eval=F}

## list all data files

data_files <- list.files(file.path(path_data), full.names = T)

## function to wrangle data into what we need (total catch per OHI region per stock)

stock_rgn_catch <- function(df) {

output_df <- readRDS(df) %>%
        select(year, taxon_scientific_name, taxon_common_name, cell_id, catch_sum, taxon_key) %>%
        rename(CellID = cell_id) %>% #change cellid name to match what is in cells.csv
        left_join(cells) %>%
        mutate(catch_prop = catch_sum * area) %>%
        group_by(year, rgn_id,fao_id, taxon_scientific_name, taxon_common_name, taxon_key) %>%
        summarise(catch = sum(catch_prop)) %>%
        ungroup() %>%
        mutate(stock_id = gsub(" ", "_", paste(taxon_scientific_name, fao_id, sep='-'), fixed=TRUE))%>%
        rename(fao_rgn  = fao_id,
               tons     = catch)

return(output_df)

}

## use purrr::map to apply the function to all yearly datasets

a <- map_df(data_files, stock_rgn_catch)

#this is a large file (150 MB) so it is saved on the server
write.csv(a, file =  file.path(dir_M,'git-annex/globalprep/fis/v2017/int/stock_catch_by_rgn.csv'),row.names=FALSE)

```

***

# Prep data for B/Bmsy calculations

Catch-MSY is the model we use to estimate stock status for all global stocks. This model requires information about the resilience of each species in addition to the catch data.

Load taxonomic resilience information

```{r resilience,eval=F}

# add the taxon_resilence data to catch for b/bmsy calculations
taxon_res = read_csv('../v2017/data/taxon_resilience_lookup.csv') %>%
              #mutate(common = ifelse(common %in% "Silver croaker", paste(common, sciname, sep=" "), common)) %>%
              dplyr::select(taxon_common_name=common, Resilience)
```

Filter out all stocks that don't meet our conditions:

(1) Keep all stocks that have at least 1000 tons mean annual harvest
(2) Keep all stocks with time series of 20 years or more

```{r prep_data_bbmsy, eval=F}

#set variables to filter by
min_yrs = 20
min_tons = 1000

#read in catch data created above
df <- read_csv(file.path(dir_M,'git-annex/globalprep/fis/v2017/int/stock_catch_by_rgn.csv'))

#create dataset ready to run through catch only models

    stks <- df%>%
              filter(taxon_key >= 600000,               #remove all records of catch reported at higher taxonomic levels than species
                     tons     > 0)%>%                  #remove records of 0 catch
              select(-rgn_id)%>%                       #remove rgn_id since we aggregate stocks to the FAO level   
              group_by(stock_id,year,fao_rgn,taxon_scientific_name,taxon_common_name,taxon_key)%>%
              summarise(tons = sum(tons))%>%           #calculate total tons per stock
              ungroup()%>%
              group_by(stock_id)%>%
              mutate(nyrs = n(),                       #get the total number of years the stock has records for   
                     avg_ann_catch = mean(tons))%>%    #calculate the mean catch over all catch years
              ungroup()%>%
              filter(avg_ann_catch >= min_tons,        #keep only those stocks that meet our conditions
                              nyrs >= min_yrs)%>%
              left_join(taxon_res)%>%                  #add resilience information
              dplyr::select(year,taxon_scientific_name,taxon_common_name,fao_rgn,stock_id,taxon_key,Resilience,tons)

# check on stocks that don't have a resilience
no_res <- filter(stks, is.na(Resilience)) %>%
          select(taxon_scientific_name, taxon_common_name) %>%
          distinct()
    
nrow(no_res) #148 species do not have a Resilience. These will get assigned a Medium Resilience by default by the CMSY model.


write.csv(stks, file = 'data/stock_catch.csv')
```


```{r}

stks = read.csv('data/stock_catch.csv')

DT::datatable(head(stks,n=100))

```


# Prep data for mean catch 

Mean catch data is used to weight the B/Bmsy values in the fishery subgoal.  

```{r mean catch, eval=F}

catch <- read_csv(file.path(dir_M,'git-annex/globalprep/fis/v2017/int/stock_catch_by_rgn.csv')) %>%
  rename(common = taxon_common_name, fao_id = fao_rgn, species=taxon_scientific_name)
summary(catch)

## filter out non ohi eez regions
catch <- catch %>%
  filter(!is.na(rgn_id)) %>%
  filter(!is.na(fao_id)) %>%
  filter(rgn_id <= 250) %>%
  filter(rgn_id != 213)

catch <- catch %>%
  dplyr::select(year, rgn_id, fao_id, stock_id, taxon_key, tons) %>%
  group_by(rgn_id, fao_id, taxon_key, stock_id, year) %>%
  summarize(catch = sum(tons)) %>%
  ungroup()

data.frame(filter(catch, stock_id == "Elasmobranchii-57" & rgn_id==1))
data.frame(filter(catch, stock_id == "Carcharhinidae-57" & rgn_id==1))

#---------------------------------------------
# for years with no reported catch, add zero values
# (after first reported catch)
# --------------------------------------------

## these data have no zero catch values, so this is added here:
catch_zeros <- catch %>%
  spread(year, catch) %>%
  data.frame() %>%
  gather("year", "catch", num_range("X", min(catch$year):max(catch$year))) %>%
  mutate(year = as.numeric(gsub("X", "", year))) %>%
  mutate(catch = ifelse(is.na(catch), 0, catch))

## this part eliminates the zero catch values prior to the first reported non-zero catch   
catch_zeros <- catch_zeros %>%
  group_by(fao_id, taxon_key, stock_id, rgn_id) %>%
  arrange(year) %>%
  mutate(cum_catch = cumsum(catch)) %>%
  filter(cum_catch > 0) %>%
  dplyr::select(-cum_catch) %>%
  ungroup()


#---------------------------------------------
### Calculate mean catch for ohi regions (using data from 1980 onward)
### These data are used to weight the RAM b/bmys values 
# --------------------------------------------

mean_catch <- catch_zeros %>%
  filter(year >= 1980) %>%
  group_by(rgn_id, fao_id, taxon_key, stock_id) %>%
  mutate(mean_catch=mean(catch, na.rm=TRUE))%>%
  filter(mean_catch != 0)  %>%      ## some stocks have no reported catch for time period
  ungroup()
filter(mean_catch, stock_id == "Elasmobranchii-57" & rgn_id==1)

data.frame(filter(mean_catch, stock_id == "Carcharhinidae-57" & rgn_id==1))

#---------------------------------------------
# Toolbox formatting and save
# --------------------------------------------
mean_catch_toolbox <- mean_catch %>%
  mutate(stock_id_taxonkey = paste(stock_id, taxon_key, sep="_")) %>%
  dplyr::select(rgn_id, stock_id_taxonkey, year, mean_catch) %>%
  filter(year>=2001) %>%  # filter to include only analysis years
  data.frame()


write.csv(mean_catch_toolbox, "data/mean_catch.csv", row.names=FALSE)



```

# Prep data for food provision weights 

These data determine the tonnes of food provided by fisheries.  Ultimately, the proportion of food from fisheries relative to mariculture will be calculated to weight the contributions of fishery and mariculture scores to final food provision scores.  

```{r tonnes fp, eval=F}

total_catch_FP <- mean_catch %>%
  group_by(rgn_id, year) %>%
  summarize(fis_catch = sum(catch)) %>%
  dplyr::select(rgn_id, year, fis_catch) %>%
  filter(year >= 2005) # filter to include only the relevant analysis years

write.csv(total_catch_FP, "data/FP_fis_catch.csv", row.names=FALSE)

```

***

# Citation information

Pauly D. and Zeller D. (Editors), 2015. Sea Around Us Concepts, Design and Data (seaaroundus.org)