---
title: "OHI 2017 - Food Provision: Combining RAM and CMSY B/Bmsy data"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../src/templates/ohi_hdr.html'
pdf_document:
  toc: true
---

# Summary

This script prepares the final B/Bmsy data:
1. Calculates the 5 year running average of B/Bmsy data generated using the CMSY method
2. Obtains a B/Bmsy value for each catch record (each FAO/OHI/year/species combination), prioritizing RAM data


## Updates from previous assessment
Same as previous years.

***

# Data

B/Bmsy values from the RAM Legacy Stock Assessment data are generated in RAM_data_prep.Rmd

B/Bmsy values from the CMSY method are generated in calculate_bbmsy.Rmd

Mean catch data created in "catch_data_prep.Rmd"

***

## Setup

``` {r setup, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/',message = FALSE, warning = FALSE)

## highlight out when knitting
# setwd("globalprep/fis/v2017")

library(dplyr)
library(tidyr)
library(zoo)
library(stringr)

source('../../../src/R/common.R')

```

## Prepare B/Bmsy values from CMSY model

For the CMSY generated B/Bmsy values we use the five year running mean of the values to smooth the data and to account for model uncertainty.

```{r CMSY mean}

cmsy <- read.csv('data/cmsy_bbmsy.csv') %>%
  filter(!is.na(bbmsy_mean)) %>%
    dplyr::select(stock_id, year, bbmsy_q2.5,bbmsy_q97.5,bbmsy_sd, bbmsy_mean, model) %>%
    arrange(stock_id, year) %>%
    group_by(stock_id) %>%
    mutate(mean_5year        = rollmean(bbmsy_mean, 5, align="right", fill=NA))
  write.csv(cmsy, "int/cmsy_b_bmsy_mean5yrs.csv", row.names=FALSE)

```

## Combine RAM and CMSY B/Bmsy values and SAUP catch data

A few regions have multiple RAM stocks of the same species. In these cases, we will average the B/Bmsy values of the species, weighted by the area of the RAM stock.

```{r combine}

cmsy <- read.csv('int/cmsy_b_bmsy_mean5yrs.csv') %>%
  select(stock_id, year, cmsy_bbmsy=mean_5year)

ram <- read.csv("int/ram_bmsy.csv")

mean_catch <- read.csv("data/mean_catch.csv") %>%
  mutate(stock_id_taxonkey = as.character(stock_id_taxonkey)) %>%
  mutate(taxon_key = str_sub(stock_id_taxonkey, -6, -1)) %>%
  mutate(stock_id = substr(stock_id_taxonkey, 1, nchar(stock_id_taxonkey)-7)) 


## combine data
setdiff(cmsy$stock_id, mean_catch$stock_id)
setdiff(mean_catch$stock_id, cmsy$stock_id)
intersect(mean_catch$stock_id, cmsy$stock_id) #920 stocks with CMSY-B/Bmsy data

setdiff(ram$stock_id, mean_catch$stock_id)
setdiff(mean_catch$stock_id, ram$stock_id)
intersect(ram$stock_id, mean_catch$stock_id) #256 stocks with RAM-B/Bmsy data (although RAM is matched by fao and rgn ids)

data <- mean_catch %>%
  left_join(ram, by=c('rgn_id', 'stock_id', "year")) # 27554 catch records (catch from specific fao and ohi regions) 
                                                     # have RAM data (~ 5% of catch records)

sum(duplicated(paste(data$rgn_id, data$stock_id, data$year, sep="_")))   # 2000 regions with multiple stocks of the same species
tmp <- data[duplicated(paste(data$rgn_id, data$stock_id, data$year, sep="_")), ]

## Example of a region with multiple RAM stocks of the same species
filter(data, rgn_id == 9, year == 2001, stock_id == "Thunnus_alalunga-71")

## regions with multiple stocks of the same species will have B/Bmsy values averaged, weighted by the area of the RAM stock
## within the region

data <- data %>%
  group_by(rgn_id, taxon_key, stock_id, year, mean_catch) %>%   
  summarize(ram_bmsy = weighted.mean(ram_bmsy, RAM_area_m2, na.rm=TRUE),
            gapfilled = ifelse(all(is.na(gapfilled)), NA, max(gapfilled, na.rm=TRUE)),
            method = paste(method, collapse = ", ")) %>%
  ungroup()

# check that averaging went ok
filter(data, rgn_id == 9, year == 2001, stock_id == "Thunnus_alalunga-71")

# add in the B/Bmsy values from the CMSY approach
data <- data %>%
  left_join(cmsy, by=c("stock_id", "year"))

```


### Formatting and saving final data

B/Bmsy values for each catch record are generated (for the species where this is possible) and saved. A corresponding gapfilling dataset is also saved.
```{r save}

 data <- data %>%
   mutate(bmsy_data_source = ifelse(!is.na(ram_bmsy), "RAM", NA)) %>%
   mutate(bmsy_data_source = ifelse(is.na(bmsy_data_source) & !is.na(cmsy_bbmsy), "CMSY", bmsy_data_source)) %>%
   mutate(bbmsy = ifelse(is.na(ram_bmsy), cmsy_bbmsy, ram_bmsy)) %>%
   select(rgn_id, stock_id, taxon_key, year, bbmsy, bmsy_data_source, RAM_gapfilled=method, mean_catch) %>%
   filter(year >= 2001)


write.csv(data, "data/fis_bbmsy_gf.csv", row.names=FALSE) 
 
bbmsy <- data %>%
  select(rgn_id, stock_id, year, bbmsy) %>%
  filter(!is.na(bbmsy))

write.csv(bbmsy, "data/fis_bbmsy.csv", row.names=FALSE) 

```

