---
title: 'OHI 2016 - Commerical Fishing Pressure Layers '
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohiprep/src/templates/ohi_hdr.html'
pdf_document:
  toc: true
---

#Summary

The commercial fishing pressure layers use data from the Sea Around Us Project



#Updates from previous assessment
[Any significant changes in methods from previous analyses?]

***

#Data Source [NOTE: can be copied from README.md in rawdata file]
**Reference**: [citation for source data; website, literature, contact information. Version of data (if relevant). Screenshots if a series of menus had to be navigated to obtain the data.]

**Downloaded**: [date downloaded or received]

**Description**:  [e.g., surface aragonite state]

**Native data resolution**: [e.g., 1 degree, 30 m, etc.]   

**Time range**: [e.g., 1880-1899, monthly data provided for each year] 

**Format**:  [e.g. NetCDF]

***
  
#Methods

##Setup

```{r setup,message=FALSE,warning=FALSE,verbose=FALSE}

#set options for all chunks in code
knitr::opts_chunk$set(warning=FALSE, message=FALSE,fig.width=6, fig.height=6)
source("~/github/ohiprep/src/R/common.R")

#ocean raster at 1km
  ocean = raster(file.path(dir_M, 'model/GL-NCEAS-Halpern2008/tmp/ocean.tif'))
  
#set mollweide projection CRS
  mollCRS=crs('+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +units=m +no_defs')

  
```


## Bring in catch by gear (high and low bycatch) at 1km2

These rasters were created in this [dataprep.R](https://github.com/OHI-Science/ohiprep/blob/master/globalprep/prs_fish/v2015/dataprep.R) script

```{r}
gear_hb = raster(file.path(dir_N,'git-annex/globalprep/prs_fish/v2015/gear_prop_hb_moll_1km_ocean.tif'))
gear_lb = raster(file.path(dir_N,'git-annex/globalprep/prs_fish/v2015/gear_prop_lb_moll_1km_ocean.tif'))
```




***

#Citation information  
[citation information: include if these data will have their own specific citation.]