---
title: 'OHI 2015: Sea Surface Temperature Pressure Layer'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohiprep/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---


#Summary

This script creates the Sea Surface Temperature (SST) layer for the 2015 global Ocean Health Index assessment.

***  

#Updates from previous assessment

The  data was updated using the same data source as previous assessments, but new years of data. CoRTAD v5 now has data from 1982 through 2012. Previous assessments used SST anomaly data up to 2010. The climatological mean for SST anomalies in previous asssessments used all years 1982-2010, and in this assessment we used the mean across all 30 years (1982-2012).

***

#Data Source

Data comes from [CoRTAD version 5](http://www.nodc.noaa.gov/sog/cortad/)

**Native Data Resolution**: 4km   
**Description**: Temperature in Kelvin  
**Time Range**: 1982 - 2012 (weekly averages across all years)  
**Format**: NetCDF  

***  

#Methods

## Setup
```{r setup, message=F,warning=F}
library(raster)
library(RColorBrewer)
library(dplyr)

#paths

dir_N = c('Windows' = '//neptune.nceas.ucsb.edu/data_edit',
          'Darwin'  = '/Volumes/data_edit',
          'Linux'   = '/var/data/ohi')[[ Sys.info()[['sysname']] ]]

cols = rev(colorRampPalette(brewer.pal(11, 'Spectral'))(255)) # rainbow color scheme

# set tmp directory for raster analysis

	tmpdir='~/big/R_raster_tmp'
	dir.create(tmpdir, showWarnings=F)
	rasterOptions(tmpdir=tmpdir)

#ocean raster at 1km
  ocean = raster(file.path(dir_N, 'model/GL-NCEAS-Halpern2008/tmp/ocean.tif'))

#set mollweide projection CRS
  mollCRS=crs('+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +units=m +no_defs')

```

***

## Create 5 year averages and calculate difference from historical mean
```{r, message=F,warning=F, eval=F}

  l   <- list.files(file.path(dir_M,'git-annex/globalprep/prs_sst/v2015/tmp'),pattern='annual_pos_anomalies',full.names=TRUE)
  
# Get 5 year aggregates

  yrs_1985_1989 <- stack(l[4:8])%>%sum(.) # This is the time period we are using for historical comparison
  
  
for(i in 2005:2008){ #i=2005
  
  print(i)
  
  yrs <- c(i,i+1,i+2,i+3,i+4)
  s   <- stack(l[substr(l,35,38)%in%yrs])%>%sum(.)
  
  diff = s - yrs_1985_1989 #calculate difference between recent 5 year mean and historical (1985-1989)
  
  projection(diff) <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
  
  out = projectRaster(diff,crs=mollCRS,progress='text',over=T)%>%
         resample(.,ocean,method='ngb',progress='text')%>%
          mask(.,ocean,filename=paste0(dir_N,'git-annex/globalprep/prs_sst/v2015/output/sst_',min(yrs),'_',max(yrs),'-1985_1989.tif',sep=""),overwrite=T)
  
  ref = quantile(out,prob=0.9999) # calculate the 99.99th quantile  
  
  sprintf('Rescaling')
  
  out_rescale = calc(out,fun=function(x){ifelse(x>0,ifelse(x>ref,1,x/ref),0)},progress='text',
                     filename=paste0(dir_N,'git-annex/globalprep/prs_sst/v2015/output/sst_',min(yrs),'_',max(yrs),'-1985_1989_rescaled.tif',sep=""),overwrite=T)
}
```

***  

## Rescale rasters by 99.99th quantile
```{r, message=F,warning=F, eval=F}    
  ## figure out the scaling value:
  rast_2012 <- raster(file.path(dir_N, 'git-annex/globalprep/prs_sst/v2015/output/sst_2005_2009-1985_1989.tif'))
  quantile(rast_2012,prob=0.9999)  #118.4027
  rast_2013 <- raster(file.path(dir_N, 'git-annex/globalprep/prs_sst/v2015/output/sst_2006_2010-1985_1989.tif'))
  quantile(rast_2013,prob=0.9999)   #133.0371 
  rast_2014 <- raster(file.path(dir_N, 'git-annex/globalprep/prs_sst/v2015/output/sst_2007_2011-1985_1989.tif'))
  quantile(rast_2014,prob=0.9999)  #127.1995
  rast_2015 <- raster(file.path(dir_N, 'git-annex/globalprep/prs_sst/v2015/output/sst_2008_2012-1985_1989.tif'))
  quantile(rast_2015,prob=0.9999)  #130.5288
```
 
*** 

## Rescale using maximum value across rasters   

This is an alternate option to the 99.99th quantile

```{r,message=F,warning=F, eval=F}
  ref <- 133.0371 # maximum value across rasters
  
  for(i in 2005:2008){ #i=2005
    print(i)
    
    final_yr <- c(i+4)

  out <- raster(file.path(dir_neptune_data, 
                          sprintf('git-annex/globalprep/prs_sst/v2015/output/sst_%s_%s-1985_1989.tif', i, final_yr)))  
  out_rescale = calc(out,fun=function(x){ifelse(x>0,ifelse(x>ref,1,x/ref),0)},progress='text',
                     filename=paste0(dir_N,'git-annex/globalprep/prs_sst/v2015/output/sst_',i,'_',final_yr,'-1985_1989_rescaled_v2.tif',sep=""),overwrite=T)
  }
```

***

# Citation information  

Selig, E.R., K.S. Casey, and J.F. Bruno (2010), New insights into global patterns of ocean temperature anomalies: implications for coral reef health and management, Global Ecology and Biogeography, DOI: 10.1111/j.1466-8238.2009.00522.x.

