---
title: 'OHI: World Governance Indicator'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

[REFERENCE RMD FILE: https://cdn.rawgit.com/OHI-Science/ohiprep/master/globalprep/np/v2016/WGI_dataprep.html]

#Summary
This script downloads WGI data and prepares it for a pressures (1 - WGI) and resilience data layer.  

#Updates from previous assessment

None

***

#Data Source 
**Reference**:  http://info.worldbank.org/governance/wgi/index.aspx#home    

**Downloaded**: Sep 6 2017

**Description**:  
The Worldwide Governance Indicators (WGI) project reports aggregate and individual governance indicators for 215 economies over the period 1996–2015, for six dimensions of governance:

* Voice and Accountability
* Political Stability and Absence of Violence
* Government Effectiveness
* Regulatory Quality
* Rule of Law
* Control of Corruption

**Time range**: 1996-2015

***

```{r}

library(ohicore) # devtools::install_github('ohi-science/ohicore@dev')
library(tools)
library(dplyr)
library(tidyr)
library(WDI) # install.packages('WDI')
library(stringr)

# comment out when knitting:
# setwd('globalprep/prs_res_wgi/v2017')

# check website to see what years are available: http://info.worldbank.org/governance/wgi/index.aspx#home
yr_start = 1996
yr_end   = 2015


```

# Obtain the WGI data
Download each of the 6 WGI indicators:
```{r}

## access data ----

## get description of variables:
indicators <-  data.frame(WDI_data[[1]])
indicators[grep("VA.EST", indicators$indicator), ]
indicators[grep("PV.EST", indicators$indicator), ]
indicators[grep("GE.EST", indicators$indicator), ]
indicators[grep("RQ.EST", indicators$indicator), ]
indicators[grep("RL.EST", indicators$indicator), ]
indicators[grep("CC.EST", indicators$indicator), ]

# identify the six indicators
# WDIsearch('violence')# general search
key_voice = WDI(
  WDIsearch('Voice and Accountability: Estimate', field='name')['indicator'],
  country='all',start = yr_start, end=yr_end)

key_polst = WDI(
  WDIsearch('Political Stability and Absence of Violence/Terrorism: Estimate', field='name')['indicator'],
  country='all',start = yr_start, end=yr_end)

key_gvtef = WDI(
  WDIsearch('Government Effectiveness: Estimate', field='name')['indicator'],
  country='all',start = yr_start, end=yr_end)

key_regqt = WDI(
  WDIsearch('Regulatory Quality: Estimate', field='name')['indicator'],
  country='all',start = yr_start, end=yr_end)

key_rolaw = WDI(
  WDIsearch('Rule of Law: Estimate', field='name')['indicator'],
  country='all',start = yr_start, end=yr_end)

key_corrp = WDI(
  WDIsearch('Control of Corruption: Estimate', field='name')['indicator'],
  country='all',start = yr_start, end=yr_end)

```

Combine the indicators.

```{r}
d = key_voice %>% 
  select(country, year, VA.EST) %>%
  left_join(key_polst %>% select(-iso2c), by=(c('country', 'year'))) %>%
  left_join(key_gvtef %>% select(-iso2c), by=(c('country', 'year'))) %>%
  left_join(key_regqt %>% select(-iso2c), by=(c('country', 'year'))) %>%
  left_join(key_rolaw %>% select(-iso2c), by=(c('country', 'year'))) %>%
  left_join(key_corrp %>% select(-iso2c), by=(c('country', 'year'))); head(d); summary(d); sapply(d, class)  
# archived record of raw data: write.csv(d, file.path('raw', 'worldbank_wgi_from_wdi_api.csv'), row.names=F)


```

# Gapfill missing data (part 1)
The first gapfilling occurs when we use the average of previous years data within each country/indicator.  This occurs when a country has data, but not for all years.

Countries without 4 or more indicators are cut (gapfilled later).
```{r}

d <- read.csv('raw/worldbank_wgi_from_wdi_api.csv')

d <- gather(d, "indicator", "value", VA.EST:CC.EST)


## each country has 17 years of data
d_gap_fill  <- d %>%
  group_by(country, year) %>%
  mutate(NA_count_c_y = sum(is.na(value))) %>%  # get count of NA values for the indicators prior to gapfilling, max value is 6 ()
  ungroup() %>%
  group_by(country, indicator) %>%              # this section gap-fills with the mean of values across years within the same region/indicator
  mutate(ind_mean_c_i = mean(value, na.rm=TRUE)) %>%
  ungroup() %>%
  mutate(value = ifelse(is.na(value), ind_mean_c_i, value)) %>%
  group_by(country, year) %>%
  mutate(NA_count_post_gf1 = sum(is.na(value)))     #count NA values for the num of indicators after last gap-fill (some values will be gapfilled by other years of data)


## cut regions with <4 indicators to calculate score:
countries_no_data <- d_gap_fill %>%
  filter(NA_count_post_gf1 > 3)

countries_no_data <- unique(countries_no_data$country)
countries_no_data

# In this case, the countries with minimal data (< 4 indicators ever calculated) are deleted.  
# These will be gap-filled later on if they are deleted now.
d_gap_fill <- d_gap_fill %>%
  filter(!(country %in% countries_no_data))

```

# Calculate overall WGI score for each country
This involves:
* taking the average of the 6 indicators (assuming there are at least 4 of the 6 indicators)
* rescaling the data from 0 to 1 

```{r}

d_calcs  <-  d_gap_fill %>%
  group_by(country, year) %>%
  summarize(score_wgi_scale = mean(value, na.rm=T),
            NA_start = mean(NA_count_c_y),
            NA_post_gf_1 = mean(NA_count_post_gf1)) %>%
  ungroup() 

wgi_range = c(-2.5, 2.5)

d_calcs <- d_calcs %>%
  mutate(score =  (score_wgi_scale - wgi_range[1]) / (wgi_range[2] - wgi_range[1])) %>%
  ungroup(); head(d_calcs); summary(d_calcs)

# document gapfilling
d_calcs <- d_calcs %>%
  mutate(gap_fill = NA_start - NA_post_gf_1,   # if there are values in NA_post_gf_1, it means these weren't gapfilled
         gap_fill = ifelse(is.na(score), 0, gap_fill)) %>%
  select(-NA_start, -NA_post_gf_1)



d_calcs[d_calcs$gap_fill>0, ]     
d_calcs[d_calcs$country == "New Caledonia", ]  # no data, was deleted earlier
d_calcs[d_calcs$country == "Niue", ] # should have gap-fill values between 0-6


## save intermediate file of wgi scores pre-gapfilling (for OHI+ use)
write.csv(d_calcs %>%
            select(country, year, score_wgi_scale, score_ohi_scale = score), 
          file.path('intermediate/wgi_combined_scores_by_country.csv'),
          row.names = FALSE)

```


# Convert country names to ohi regions

```{r}

## We report these regions at a greater spatial resolution:

## Aruba is part of the Netherlands Antilles, but it is reported separately
country_split_1 <- data.frame(country = "Netherlands Antilles", region = c('Bonaire', 'Curacao', 'Saba', 'Sint Maarten', 'Sint Eustatius'))
country_split_2 <- data.frame(country = "Jersey, Channel Islands", region = c('Jersey', 'Guernsey'))
country_split <- rbind(country_split_1, country_split_2)

country_split_data <- country_split %>%
  left_join(d_calcs) %>%
  select(-country) %>%
  rename(country = region)

d_calcs <- d_calcs %>%
  filter(!(country %in% c("Netherlands Antilles", "Jersey, Channel Islands"))) %>%
  rbind(country_split_data)  %>%
  mutate(country = as.character(country))

d_calcs$country[grep("Korea, Dem.", d_calcs$country)] <- "North Korea"



### Function to convert to OHI region ID
d_calcs_rgn <- name_2_rgn(df_in = d_calcs, 
                       fld_name='country', 
                       flds_unique=c('year'))

### Combine the duplicate regions (we report these at lower resolution)
### In this case, we take the weighted average
population_weights <- data.frame(country = c("Virgin Islands (U.S.)", "Puerto Rico",
                                             "China", "Hong Kong SAR, China", "Macao SAR, China"),
                                 population = c(106405, 3725789,
                                         1339724852, 7071576, 636200))

d_calcs_rgn <- d_calcs_rgn %>%
  left_join(population_weights, by="country") %>%
  mutate(population = ifelse(is.na(population), 1, population)) %>%
  group_by(rgn_id, year) %>%
  summarize(score = weighted.mean(score, population),
            gapfill_within_rgn = weighted.mean(gap_fill, population)) %>%
  ungroup() %>%
  filter(rgn_id <= 250)

summary(d_calcs_rgn)

```


# Gapfill missing data (part 2)
This gapfilling occurs when we use assign a territorial region the value of their parent country.

```{r}

## data that describes territories of countries
territory = rgn_master %>% 
  select(rgn_id = rgn_id_2013,
         sov_id) %>%               
  group_by(rgn_id) %>%                  # remove duplicated countries from this rgn_id list                    
  summarize(sov_id = mean(sov_id, na.rm=T)) %>% # duplicates always have the same sov_id (r2 value)
  filter(rgn_id <= 250, rgn_id != 213)

	
## expand to include all years of data
territory <- data.frame(year=yr_start:yr_end) %>% 
  merge(territory, by=NULL) 


# assign territories the values of their country
d_sovs = d_calcs_rgn %>% 
  full_join(territory, by = c('rgn_id', 'year')) %>%
  group_by(sov_id, year) %>%
  mutate(score_gf_territory = mean(score, na.rm=TRUE),
         gapfill_within_rgn = mean(gapfill_within_rgn, na.rm=TRUE))%>%
   filter(!is.na(gapfill_within_rgn)) %>%
  ungroup()

head(d_sovs)
summary(d_sovs)

# format these data
d_gf2 <- d_sovs %>%
  mutate(gapfill_territory = ifelse(is.na(score) & !is.na(score_gf_territory), "territory", "NA")) %>%
  mutate(score = ifelse(is.na(score), score_gf_territory, score)) %>%
  select(rgn_id, year, score, gapfill_within_rgn, gapfill_territory)

# make sure that all regions have a score
regions <- rgn_master %>%
  filter(rgn_typ == "eez") %>%
  filter(rgn_id_2013 <= 250) %>%
  filter(rgn_id_2013 != 213) %>%
  select(rgn_id = rgn_id_2013) %>%
  unique() %>%
  arrange(rgn_id)

d_gf2 <- regions %>%
  left_join(d_gf2)

## check for NA values within "score" variable
## if so, need to gapfill using UN geopolitical regions
summary(d_gf2)

```

# Uninhabited regions

These regions will receive an NA for their score (when established population is < 100 people).  We decided it would be better to give uninhabited regions the scores of their administrative countries.

```{r}
# uninhab <- read.csv('../../../src/LookupTables/rgn_uninhabited_islands.csv') %>%
#   filter(is.na(est_population) | est_population < 100)
# 
# d_gf2 <- d_gf2 %>%
#   mutate(score = ifelse(rgn_id %in% uninhab$rgn_id, NA, score))

```


# Check data 
Comparing this year's values against last year's.  These should be the same unless there have been updates to WGI source data or a change to methods.  For this year, there was a small change that effected a few territorial regions.  In the past, we used the sovereign country value, but in the case, we averaged the sovereign country and the available territorial values. For example, 

Also look at top/bottom 10 regions to make sure these seem reasonable.

```{r}

new2014 <- d_gf2 %>%
  filter(year==2014) %>%
  select(rgn_id, score)

old2014 <- read.csv('../v2016/output/wgi_res_2016.csv') %>%
  select(rgn_id, old_score=resilience_score) %>%
  full_join(new2014)

## should be a 1:1 relationship
plot(old_score ~ score, data=old2014)
abline(0,1, col="red")


## Top/Bottom 10 scorers:

# get region names
regions <- rgn_master %>%
  filter(rgn_typ == "eez") %>%
  filter(rgn_id_2013 <= 250) %>%
  filter(rgn_id_2013 != 213) %>%
  select(rgn_id = rgn_id_2013, rgn_name=rgn_nam_2013) %>%
  unique() %>%
  arrange(rgn_id)


tmp <- d_gf2 %>%
  filter(year==2015) %>%
  arrange(score) %>%
  select(rgn_id, score) %>%
  left_join(regions)

tmp[1:10, ]
tmp[211:220, ]

hist(tmp$score)

```


# Save the data
Gapfilling and score data are saved for each scenario year.  

```{r}


  tmp_data_res <- d_gf2 %>%
    select(rgn_id, year, resilience_score = score)
  write.csv(tmp_data_res, "output/wgi_res.csv", row.names=FALSE)
  
  tmp_data_prs <- d_gf2 %>%
    mutate(score = 1 - score) %>%
    select(rgn_id, year, pressure_score = score)
  write.csv(tmp_data_prs, "output/wgi_prs.csv", row.names=FALSE)
  
  tmp_gf <- d_gf2 %>%
    select(rgn_id, year, gapfill_within_rgn, gapfill_territory)
  write.csv(tmp_gf, "output/wgi_gf.csv", row.names=FALSE)
  

```