---
title: 'OHI 2016: Ocean Acidification Data Prep '
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohiprep/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

#Summary

This script takes the raw data, shared by Woods Hole Institute for Oceanography, and does the following:

   a. Calculates the historical global mean for the decade 1880-1889 (1 raster layer as output)
   b. Calculates the annual mean for each of the 10 years in 2005-2015 (10 raster layers as output)
  
**Notes about the data:**  
This data was shared with us by Ivan Lima from Woods Hole Institue for Oceanography in December 2014 and again February 2016. The data came as NetCDFs in an irregular grid format with a resolution of about 1 degree. The data values are monthly average surface &#937; aragonite saturation state.


#Updates from previous assessment

Added one more year of data (2015) and rescaled values greater than 1 using


Deviation from aragonite saturation state is determined for each year in the study period:

$$\Delta \Omega_{year} = \frac{(\Omega_{base} - \Omega_{year})}{(\Omega_{base} - 1)}$$

Note that current is subtracted from the baseline; this way, a reduction in $\Omega$ becomes a positive pressure value.  It is then normalized by the current mean state; so a decrease in $\Omega$ while the current state is high indicates less pressure than the same decrease when the current state is near 1. 

$\Delta \Omega_{year}$ is then modified to account for increases in aragonite saturation state (pressure = 0) and arag sat state less than 1 (pressure = 1).

***

#Data Source
**Reference**: [Feely et al.(2009)](https://darchive.mblwhoilibrary.org/bitstream/handle/1912/3180/22-4_feely.pdf?sequence=1&isAllowed=y)

**Downloaded**: March 15, 2016

**Description**:  Aragonite Saturation State  $\Omega_{arg}$

**Native data resolution**: 1 degree

**Time range**: 1880-1889 and 2005-2015, monthly data provided for each year

**Format**:  NetCDF

***

# Methods


##Setup  

The main R libraries needed for this analysis are the `raster`, and `ncdf4` packages.
```{r setup,message=FALSE,warning=FALSE,verbose=FALSE}

#set options for all chunks in code
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
source("~/github/ohiprep/src/R/common.R")

#libraries

library(raster)
library(ncdf4)
library(maps)

#define paths for the raw data and OA folder held on git-annex on our NCEAS server, Mazu

raw_dir    = file.path(dir_M,'git-annex/globalprep/_raw_data')
oagit_dir  = file.path(dir_M,'git-annex/globalprep/prs_oa')

# set temporary directory to folder on mazu disk big enough to handle it
tmpdir='~/big/R_raster_tmp'
dir.create(tmpdir, showWarnings=F)
rasterOptions(tmpdir=tmpdir)
```
  
## Load raw  

Read in raw data to get lat and long information and monthly aragonite saturation state values

```{r,message=FALSE,warning=FALSE,eval=FALSE}

#list all 2015-2020 files

files <- list.files(file.path(raw_dir,'WHOI/annual'),pattern=".nc",recursive=T,full.names = T)[12:17] #grab only 2015-2020
        
```

## Step Two  
Get annual averages for each year

```{r,message=FALSE,warning=FALSE,eval=FALSE}
      
#annual_arag is a function that calculates annual average surface aragonite saturation 

  annual_arag<-function(file){
    
    year = substr(file,73,76)
    
    nc = nc_open(file)
            
# longitude values are stored in the variable 'TLONG'

      long <- ncvar_get(nc,varid='TLONG') #same lat and long for both data sets so just reading in from data_nc_18 works

# latitude values are stored in the variable 'TLAT'

      lat <- ncvar_get(nc,varid='TLAT')
    
  
  # select the surface aragonite variable (OARG)

      arag<- ncvar_get(nc,varid="OARG")


# land cells are given large, strange value, so set these to NA

      arag[arag==9969209968386869046778552952102584320]<-NA 

      yr_mean = apply(arag,1:2,mean) # get the annual mean
  
      #create an array with long, lat, aragonite mean data
      A <- array(c(long,lat,yr_mean),dim=c(320,384,3))
      B <- apply(A, 3, cbind)
  
      #lon=x, lat=y
      x = B[,1]
      y = B[,2]
  
  
      C = as.data.frame(B)
      names(C)<-c('x','y','value')
  
      #set extent to lon/lat
      e <- extent(C[,1:2])
      r <- raster(e,ncol=320,nrow=384) #create empty raster with e extent
  
      #rasterize the dataframe
      out <- rasterize(C[,1:2],r,C[,3],fun=function(x,...)mean(x),progress='text') # i had to create a mean function here for "multiple points in a cell"
      extent(out)<-c(0,360,-80,90) #data extent is 0,360,-80, 90 (the -80 is not a typo)
      out <- rotate(out) #shifts data from 0-360 to -180-180
  
  
      plot(out)
  #    map('world',col='gray95',fill=T,border='gray80',add=T)
  
      #notice the empty cells, due to irregular grid, so plot using a different projection
  
      
      # Define initial projection for out
      projection(out) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
  
      # define mollweide projection
      mollCRS <- CRS('+proj=moll')

      # then convert to mollweide and other projections
      out_moll <- projectRaster(out, crs=mollCRS, over=T)
  
  
      # plot results
     plot(out_moll)
      
     print(cellStats(out_moll,stat='mean',na.rm=T))
     
     
  writeRaster(out_moll,filename = paste0(file.path(dir_M,"git-annex/globalprep/prs_oa/v2016/int/global_arag_avg_moll_"),year), format='GTiff', overwrite=T)
    
}

#run the function

# lapply(files,annual_arag)
 
```

## Step Three  
Read in historical mean from 1880-1889

```{r histMean}

hist <- raster(file.path(dir_M,'git-annex/globalprep/prs_oa/v2015/working/global_oa_1880_1889_arag_mean_moll.tif'))

```


## Rescale 

Baseline aragonite saturation state $\Omega_{base}$ is represented by the decadal average from 1880-1889.

Deviation from aragonite saturation state is determined for each year in the study period:

$$\Delta \Omega_{year} = \frac{(\Omega_{base} - \Omega_{year})}{(\Omega_{base} - 1)}$$

Note that current is subtracted from the baseline; this way, a reduction in $\Omega$ becomes a positive pressure value.  It is then normalized by the current mean state; so a decrease in $\Omega$ while the current state is high indicates less pressure than the same decrease when the current state is near 1. 

$\Delta \Omega_{year}$ is then modified to account for increases in aragonite saturation state (pressure = 0) and arag sat state less than 1 (pressure = 1).


```{r}

#Get 2011-2015 rasters

files = c(list.files(file.path(oagit_dir,'v2015/working/annualmean_2005-2014/moll'),full.names = T)[7:10],file.path(oagit_dir,'v2016/int/global_arag_avg_moll_2015.tif'))


#for each layer, all values <=1 are assigned a 1, otherwise old-new/old

oaRescale <- function(file){
  
  yr   = substr(file,nchar(file)-7,nchar(file)-4)         #get year of file
  mean = raster(file)               #get seasonal mean aragonite raster for given year
  diff = (hist-mean)/(hist-1)
  mean[mean<=1]<-1                                #all values at or less than 1 are given a value of 1
  mean[mean>1] <- diff[mean>1]                      #all cells with values greater than 1 are swapped out with their amount of change scaled to how close to 1 
  mean[mean<0]<-0                                 #all values less than 0 (indicating a decrease in acidity) are capped at 0

    writeRaster(mean,filename=paste0(oagit_dir,'/v2016/int/annual_oa_rescaled/oa_rescaled_',yr,sep=""),format='GTiff',overwrite=T)

}

#  sapply(files,oaRescale)

  rescaled = list.files(file.path(oagit_dir,'v2016/int/annual_oa_rescaled'),full.names=T)

```


##Interpolate

Interpolation was done using python and ArcPY. 

```{r interpolate}

int_files <- list.files(file.path(oagit_dir,'v2016/int/annual_oa_rescaled_int'),full.names=T)

```


## Resample

```{r resample}

#ocean is a raster with all land clipped out - at 1km with value of 1. This is used as a mask
ocean = raster(file.path(oagit_dir,'v2016/int/ocean.tif'))

resample = function(file){
  
        yr  = substr(file,91,94)
        r   = raster(file)%>%raster::resample(ocean,method='ngb',progress='text') # resample r to the resolution of 'ocean' (~1km)
    
       writeRaster(r,filename=paste0(oagit_dir,'/v2016/int/annual_oa_rescaled_int_1km/annual_oa_rescaled_int_1km_',yr,sep=''),format='GTiff',overwrite=T)
}

#sapply(int_files,resample)

resampled = list.files(file.path(oagit_dir,'v2016/int/annual_oa_rescaled_int_1km'),full.names=T)

```


##Remove Land

```{r mask}

rm_land <- function(file){
  yr = substr(file,110,113)
  r  = raster(file)%>%
          mask(ocean,filename=paste0(oagit_dir,'/v2016/output/oa_prs_layer_',yr,sep=''),format='GTiff',overwrite=T)
}

#sapply(resampled,rm_land)


```



