---
title: "OHI 2016 - Food Provision: Preparing catch data for BBmsy calculations"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohiprep/src/templates/ohi_hdr.html'
pdf_document:
  toc: true
---


# Summary

This script takes the Sea Around Us Project (SAUP) catch data, provided at a resolution of half-degree cells globally, and aggregates catch to **stock levels**. For the Ocean Health Index, we assume a stock is represented by the FAO region in which the species is caught.

In order to aggregate to FAO regions, we associate each cell to the FAO region and the OHI region in which it is located.

An example of our aggregation proces: New Zealand is located entirely in FAO region 81. All catch reported by New Zealand will be aggregated by species to the FAO region. If a species was reported as caught in both New Zealand waters and in the High Seas of area 81, these two records will be combined into one by summing the catch.

## Updates from previous assessment
The 2016 Food Provision goal now includes new data from the Sea Around Us Project that has been reported at a spatial resolution of half-degree cells. The spatial resolution previously was at the EEZ level. This improvement in data resolution allows OHI to more accurately represent stock locations and calculate BBmsy values. The data from SAUP has also been updated by SAUP to more accurately reflect catch history per country world wide.

***

# Data
The Sea Around Us Project shared the spatialized catch data with OHI on joined to a lookup table that links SAUP region names and ids to the FAO region they are located in. The proportional area of each EEZ within the FAO region was also calculated for overlapping EEZs.


Reference: 

* Downloaded: June 27, 2016
* Description: Tons per half degree cell with information on sector type, industry type, fishing entity, reporting status and taxonomic information
* Native data resolution: Flatfiles (.dat) provided with information for 0.5 degree grid cells
* Time range: 1950 - 2010
* Format: Database (.dat) flatfiles

***

## Setup

``` {r setup, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/',message = FALSE, warning = FALSE)

#setting up provenance
# devtools::install_github('oharac/provRmd')
# library(provRmd)
# prov_setup()

## Libraries
library(readr)
library(data.table)
library(dplyr)
library(parallel)


## Paths for data
path_data               = "/home/shares/ohi/git-annex/globalprep/_raw_data/SAUP/d2016/Data"
file_allocation_data    = "SeaAroundUs/AllocationData.dat"
file_allocation_results = "SeaAroundUs/AllocationResult.dat"
file_taxon              = "SeaAroundUs/taxon.dat"
file_entity             = "FishingEntity.dat"

source('~/github/ohiprep/src/R/common.R')

```

## Load Data

These files are large so using the data.table package is recommended due to R memory limitations.

```{r load_data,eval=F}

# load the Allocation info using fread, and define column names
dt_data           <- fread(file.path(path_data,file_allocation_data), sep=";", header = FALSE)
colnames(dt_data) <- c("UniversalDataID","DataLayerID","FishingEntityID", "Year", "TaxonKey",
                       "InputTypeID", "sector_type_name", "catch_type_name", 
                       "reporting_status_name")


#load the Results data (largest file, usually takes up to 10 minutes to read!)
dt_results           <- fread(file.path(path_data,file_allocation_results), sep=";", header = FALSE)
colnames(dt_results) <- c("UniversalDataID","CellID","AllocatedCatch")
# setkey(dt_results,UniversalDataID) # not necessary the data seems to be already ordered with the keys (univ and Cell IDs)


#load the Taxon data
dt_taxon           <- fread(file.path(path_data,file_taxon), sep=";", header = FALSE)
colnames(dt_taxon) <- c("TaxonKey","Scientific_Name","Common_Name","FunctionalGroupDescription")
setkey(dt_taxon,TaxonKey)


#load the fishing entity data
dt_entity           <- fread(file.path(path_data,file_entity), sep = ";", header=FALSE)
colnames(dt_entity) <- c("FishingEntityID","Name")
setkey(dt_entity,FishingEntityID)
```

# Fix Cell Issues

There are a lot of cells that slighlty overlap the OHI regions shapefile, leaving them with a proportional area less than 1. This would cause us to lose catch when assigning catch to cells. To fix this, we define a vector of cellids that have a proportionArea <1 and are NOT duplicated (i.e. the other portion of the area missing is not accounted for) and assign a proportionArea of 1 to these cells.

```{r cells, eval=F}
###

cells_raw <- read.csv(file.path(dir_M, "git-annex/globalprep/fis/v2015/raw/saup_rasters_to_ohi_rgns.csv"))%>%
              rename(CellID=saup_cell_id) %>%
              group_by(CellID) %>%
              mutate(total_area = sum(proportionArea)) 

### list of cells with areas > 1 (indicates something strange is going on)
bad_ones <- filter(cells_raw, total_area>1) %>%
              arrange(CellID) 

data.frame(bad_ones)

## get a list of cells that have eez or fao categories:
## used later to determine whether cells with OHI regions but no FAO region are an issue:
cells_water <- cells_raw %>%
                group_by(CellID) %>%
                mutate(cell_rgns = paste(sort(unique(rgn_typ)), collapse=", ")) %>%
                arrange(CellID)

# check this went ok:
list <- table(cells_water$cell_rgns)
list
filter(cells_water, cell_rgns=="eez, land")

## check that I got these right!  - these are right (JA - 8/8/26)
eez_fao_cats <- c("eez", "eez-ccamlr", "eez-ccamlr, land-ccamlr", "eez-disputed",
                  "eez-disputed, fao", "eez-disputed, land-disputed", "eez-disputed, land, land-disputed", "eez, eez-ccamlr",
                  "eez, eez-disputed", "eez, eez-disputed, fao", "eez, eez-disputed, land", "eez, eez-disputed, land-disputed",
                  "eez, eez-disputed, land, land-disputed", "eez, fao", "eez, land", "eez, land, land-noeez", "fao")                                   
cells_water <- cells_water %>%
  filter(CellID, cell_rgns %in% eez_fao_cats)


## cells should have a total area of <=1, what causes some cells to have more...and is this a large problem
## No, most cells are fine.
## This seems to happen because of polygon overlap.
## scenario 1: FAO region 262 overlaps other FAO region polygons by a small amount.  In this case a small proportion of the catch within a cell will
## be assigned to two regions.  This will be a small error. (No correction)
## scenario 2: In many cases, the total area is very close to one which may reflect rounding error and is not significant. (Usually no correction)
## scenario 3: In some cases, the regions are small islands where there doesn't appear to be a hole so both the land and underlying eez are counted. ##            This can also occur along any eez/land boundary...but it looks like it mainly happens for islands. If the 
##            overlap is for land/eez within the same region, this will be corrected. 


cells <- read.csv(file.path(dir_M, "git-annex/globalprep/fis/v2015/raw/saup_rasters_to_ohi_rgns.csv")) %>%
  rename(CellID=saup_cell_id) %>%
  group_by(CellID, rgn_id) %>%   # groups land and eez data (that way the cell catch is fully applied to the region...rather than cutting the portion that overlaps land) 
  dplyr::summarise(area = sum(proportionArea)) %>%
  mutate(area = ifelse(area > 1, 1, area))%>%  ## this corrects when there is land/eez overlap within the same region resulting in cell area >1 (scenario 3 above)
  ungroup()  

## Remaining errors from scenario 1 and 2 above:  These are very small errors..nothing to be concerned about.
bad_ones <- cells %>%
  group_by(CellID) %>%
  mutate(total_cell_area = sum(area)) %>%
  filter(total_cell_area > 1) %>%
  arrange(CellID) 
data.frame(bad_ones)


# get the list of cell ids that are duplicated, and use this list of values to adjust the area to equal 1 ONLY for those cells that are not duplicated


## Id duplicated cells:
dup <- cells$CellID[duplicated(cells$CellID)]

## for some reason this is failing for me...
# dup <- cells %>%
#         dplyr::select(CellID) %>%
#         mutate(dup = duplicated(.)) %>%
#         filter(dup==TRUE) %>%
#         collect %>%
#         .[["CellID"]]

## these are the cells that were cut off prematurely due to edge effects, etc.
tmp <- filter(cells, !(CellID %in% dup) & area < 1)
head(tmp)

#read in the dataset matching each cell to an FAO region (need both OHI and FAO region for analysis)
fao_cells <- read.csv( file.path(dir_M, "git-annex/globalprep/fis/v2015/raw/saup_rasters_to_fao_rgns.csv")) %>%
              rename(CellID=saup_cell_id)

cells_df <- cells %>%
              mutate(area = ifelse(CellID %in% dup, area, 1))%>%  # if the cell doesn't cover >1 region, then change cell areas to one to capture entire cell's catch (these are <1 area due to edge effects)
              left_join(fao_cells) 

summary(cells_df)
## One issue of concern:
## Some regions aren't assigned and FAO region value (in places where land > 50% of cell cover)...and we need both for the analysis
#84218/259200

## check to see how many remain after we take out the land cells:

cells_df_water <- cells_df %>%
  filter(CellID %in% cells_water$CellID)
summary(cells_df_water)
#187/252083
# N= 187...no too bad...does not seem worth fretting over
```


***

# Aggregate catch

Using a for loop, aggregate catch per OHI region and FAO area. This catch will be used twice. 

(1) The catch is used to weight scores per region. For this we need to use all catch records, including those not reported at the species level.

(2) The catch data at species level is used to calculate stock status (BBmsy) per stock (remember that our definition of a stock is a species caught within a single FAO area).

```{r, eval =F}

df <- data.frame()


for (i in 1950:2010){

  print(i)
  
#1. subset the allocation data to year i
data_yearly <- dt_data[Year==i,]

#2. Now we want it ordered by UniversalDataID
setkey(data_yearly,UniversalDataID)

#3. Get unique UniversalDataID

udi <- unique(data_yearly$UniversalDataID)

#4. Subset results

results_sub <- dt_results[UniversalDataID %in% udi]

setkey(results_sub,UniversalDataID) #sort by UDI


#5. Join allocation, taxon, entity, resilience data to results to create final catch dataset and removing all catch reported at non-species level


all_data <- results_sub%>%
               left_join(data_yearly)%>%
               left_join(dt_taxon)%>%
               left_join(dt_entity)%>%
               left_join(cells_df_water)%>%
               mutate(catch_prop = AllocatedCatch * area,
                         year = i)%>%
                group_by(rgn_id,fao_id, Scientific_Name, Common_Name, TaxonKey)%>%
                summarise(catch = sum(catch_prop))%>%
                ungroup()%>%
                mutate(year     = i,
                       stock_id = gsub(" ", "_", paste(Scientific_Name, fao_id, sep='-'), fixed=TRUE))%>%
                rename(fao_rgn  = fao_id,
                       tons     = catch)



df = rbind(df,all_data)

}

write.csv(df,file =  file.path(dir_M,'git-annex/globalprep/fis/v2016/int/spatial_catch_saup.csv'),row.names=FALSE)


```


***

# Prep data for BBmsy calculations

Load taxonomic resilience information

```{r resilience,eval=F}

# add the taxon_resilence data to catch for b/bmsy calculations
taxon_res = read.csv('int/taxon_resilience_lookup.csv', stringsAsFactors = FALSE) %>%
              mutate(common = ifelse(common %in% "Silver croaker", paste(common, sciname, sep=" "), common)) %>%
              dplyr::select(Common_Name=common, Resilience)
```

Filter out all stocks that don't meet our conditions:

(1) Keep all stocks that have at least 1000 tons mean annual harvest
(2) Keep all stocks with time series of 20 years or more

```{r prep_data_bbmsy, eval=F}

#set variables to filter by
min_yrs = 20
min_tons = 1000

#read in catch data created above
df <- read.csv(file.path(dir_M,'git-annex/globalprep/fis/v2016/int/spatial_catch_saup.csv'),stringsAsFactors=F)

#create dataset ready to run through catch only models

    stks <- df%>%
              filter(TaxonKey >= 600000,               #remove all records of catch reported at higher taxonomic levels than species
                     tons     > 0)%>%                  #remove records of 0 catch
              select(-rgn_id)%>%                       #remove rgn_id since we aggregate stocks to the FAO level   
              group_by(stock_id,year,fao_rgn,Scientific_Name,Common_Name,TaxonKey)%>%
              summarise(tons = sum(tons))%>%           #calculate total tons per stock
              ungroup()%>%
              group_by(stock_id)%>%
              mutate(nyrs = n(),                       #get the total number of years the stock has records for   
                     avg_ann_catch = mean(tons))%>%    #calculate the mean catch over all catch years
              ungroup()%>%
              filter(avg_ann_catch >= min_tons,        #keep only those stocks that meet our conditions
                              nyrs >= min_yrs)%>%
              left_join(taxon_res)%>%                  #add resilience information
              dplyr::select(year,Scientific_Name,Common_Name,fao_rgn,stock_id,TaxonKey,Resilience,tons)



write.csv(stks, file = 'int/spatial_catch_pre_bbmsy.csv')
```

```{r,echo=F}

stks = read.csv('int/spatial_catch_pre_bbmsy.csv')

DT::datatable(head(stks,n=100))


```

***

# Citation information

Pauly D. and Zeller D. (Editors), 2015. Sea Around Us Concepts, Design and Data (seaaroundus.org)