---
title: 'OHI: Mariculture '
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

[REFERENCE RMD FILE: https://cdn.rawgit.com/OHI-Science/ohiprep/master/globalprep/mar/v2016/mar_dataprep.html]

#Summary
This analysis converts FAO mariculture data into the OHI 2016 global mariculture status score.  This also calculates the genetic escapee from mariculture pressure data.

#Updates from previous assessment
New year of FAO mariculture yield data, but no changes to sustainability or genetic escapee data or general methods.  Code has been improved from last year.  

***

#Data Source 
**Reference**:      
    http://www.fao.org/fishery/statistics/software/fishstatj/en#downlApp
     Release date: March 2016 
FAO Global Aquaculture Production Quantity 1950_2014

**Downloaded**: July 29 2016

**Description**:  Quantity (tonnes) of mariculture for each county, species, year.

**Time range**: 1950-2014 

***
  
#Methods
```{r}
# load libraries, set directories
library(ohicore)  #devtools::install_github('ohi-science/ohicore@dev')
library(dplyr)
library(stringr)
library(tidyr)

## comment out when knitting
#setwd("globalprep/mar/v2016")


### Load FAO-specific user-defined functions
source('mar_fxs.R') # functions specific to mariculture dealing with compound countries
source('../../../src/R/fao_fxn.R') # function for cleaning FAO files
source('../../../src/R/common.R') # directory locations

```

# FAO Mariculuture data

Clean mariculture data: Filter freshwater mariculture, make long format, and clean FAO codes.
```{r}

mar <- read.csv(file.path(dir_M, 'git-annex/globalprep/_raw_data/FAO_mariculture/d2016/GlobalAquacultureProduction_Quantity_1950_2014.csv'), check.names=FALSE, stringsAsFactors=FALSE) ; head(mar) 

mar <- mar %>%
  rename(country = `Country (Country)`,
         FAO_name = `Species (ASFIS species)`,
         fao = `Aquaculture area (FAO major fishing area)`,
         environment = `Environment (Environment)`)
table(mar$environment)  

# include only marine environments
mar <- mar %>%
filter(environment %in% c("Brackishwater", "Marine"))  

## long format and clean FAO codes:
mar <- mar %>%
  select(-Unit) 

mar <- mar %>%
  gather("year", "value", 5:(ncol(mar))) %>%
    fao_clean_data() 

```

Update species names and exclude non-food species.
```{r}

mar_sp <- read.csv('raw/species_list.csv', stringsAsFactors=FALSE) %>%
  select(FAO_name, species, exclude, alias, Taxon_code)
new.spp <- setdiff(mar$FAO_name, mar_sp$FAO_name)
new.spp # check: if dim has 0 rows it means all match
# if length(new.spp) >0 , hand check whether to keep (exclude seaweeds and species harvested for ornamental/medicinal), check if synonyms match Trujillo names

# Remove species not relevant to mariculuture goal (i.e., non-food species)
mar <- left_join (mar, mar_sp, by="FAO_name") 
mar <- filter (mar, exclude==0) 
# change names using species name and the species alias
mar$species <- ifelse(!is.na(mar$alias), mar$alias, mar$species) 

# sum duplicates after name change (this also gets rid of the NA values)
mar <- mar %>%
  filter(!is.na(value)) %>%
  group_by(country, fao, environment, species, year, Taxon_code) %>%
    summarize(value = sum(value)) %>% 
  ungroup()

# remove extra fields
mar = mar %>% 
  select(country, fao, environment, species, year, value, Taxon_code)

# eliminate time-series with all 0s
mar <- mar %>%
  group_by(country, species) %>%
  mutate(total_value = sum(value)) %>%
  filter(total_value > 0) %>%
  select(-total_value) %>%
  ungroup()

```

Convert country names to OHI regions

```{r}

# Divide mariculture from countries that we report as separate regions (assume equal production in all regions)
# Netherlands Antilles: Conch restoration among Aruba, Bonaire, Curacao
# Channel Islands: Jersey and Guernsey
# Bonaire/S.Eustatius/Saba
# Yugoslavia SFR: no longer a country after 1992

mar <- mar %>%
  mutate(country = ifelse(country=="R\xe9union", "Reunion", country)) %>%  # this one is hard to get right
  mar_split()  # function in mar_fxs.R

mar_rgn <- name_2_rgn(df_in = mar, 
                       fld_name='country', 
                       flds_unique=c('species', 'fao', 'environment', 'Taxon_code', 'year'))


### sum values of regions with multiple subregions
mar_rgn <- mar_rgn %>%
  group_by(fao, environment, species, year, Taxon_code, rgn_id) %>%
  summarize(value = sum(value)) %>%
  ungroup()

```

### Gapfilling
Fill in missing years after first harvest with 0 values

```{r}

mar_rgn_spread <- spread(mar_rgn, year, value) 
mar_rgn_gf <- gather(mar_rgn_spread, "year", "value", 6:ncol(mar_rgn_spread)) %>%
  arrange(rgn_id, species, year, Taxon_code, fao, environment)

mar_rgn_gf <- mar_rgn_gf %>%
  mutate(year = as.numeric(as.character(year))) %>%
  mutate(value_w_0 = ifelse(is.na(value), 0, value)) %>%
  group_by(fao, environment, species, Taxon_code, rgn_id) %>%
  mutate(cum_value = cumsum(value_w_0)) %>%
  ungroup() %>%
  filter(cum_value > 0) %>%
  mutate(gap_0_fill = ifelse(is.na(value), "NA_to_zero", "0")) %>%
  mutate(value = ifelse(is.na(value), 0, value)) %>%
  select(-cum_value, -value_w_0)
table(mar_rgn_gf$gap_0_fill)
## 2943 of these out of 24887+2943 cases

```

Remove time series with less than four non-zero datapoints (assume these are not established mariculture programs).
```{r}

mar_rgn_gf = mar_rgn_gf %>% 
  group_by(rgn_id, species, fao, environment) %>%
  mutate (not_0 = length(value[value>0])) %>% 
  filter (not_0>3) %>%
  ungroup() %>% 
  select(rgn_id, species, fao, environment, year, value, Taxon_code, gap_0_fill)

```

Add a unique identifier per cultivated stock that describes species, fao region, and environment.
```{r}

# add a unique identifier per cultivated stock
identifier = mar_rgn_gf %>% 
  select(rgn_id, species, fao, environment) %>% 
  unique() %>% 
  mutate(species_code = 1:n())

mar_rgn_gf = left_join(mar_rgn_gf, identifier)
maric <- mar_rgn_gf

```


Save file to estimate total mariculture yield per country.
```{r}

write.csv(maric, 'output/MAR_FP_data.csv', row.names=FALSE)

```

# Trujillo sustainability scores
(NOTE: These data have been simplified from last year, I eliminated all the fao/environment ditinctions that did not differ in regard to sustainability or genetic escapees.  This just eliminates extra complexity that doesn't alter scores.)

These data describe the sustainability and genetic escapes for country/species combinations (and, in a couple cases, environment and fao region combinations). In cases where these data were not available for a specific county/species, we averaged the data across taxonomic groups to gapfill the missing data.

Convert country names to OHI region names.
```{r}

# Trujillo sustainability data:
sus = read.csv('raw/Truj_label_sust.csv', stringsAsFactors = FALSE, na.strings = NA)

## these need to be re-added (get cut when associated with region ids)
sus_no_rgn <- filter(sus, is.na(country))

# convert country names to OHI region names:
sus_rgn <- name_2_rgn(df_in = sus, 
                       fld_name='country', 
                       flds_unique=c('species_fao', 'fao', 'environment', 'species_Truj'))

sus_rgn <- bind_rows(sus_rgn, sus_no_rgn)

# check the fao spp list in the Trujillo sustainability file matches FAO mariculture species
setdiff(sus_rgn$species_fao, maric$species) # species that are no longer have mariculture industry or are not included due to being freshwater or non-food

sort(setdiff(maric$species, sus_rgn$species_fao)) # species with no Trujillo data

```

# FAO maricultue and sustainability scores

Match the sustainability score to the FAO mariculture data.

The following joins the sustainability scores to regions/species that have Trujillo data.
```{r}

table(sus_rgn$match_type)

# join taxa specific to country/species/environment
c_sp_env = sus_rgn %>% 
  filter(match_type == 'c_sp_env') %>% 
  select(rgn_id, species=species_fao, environment, Sust_c_sp_env = Maric_sustainability)

mar_sus <- maric %>%
  left_join(c_sp_env, by= c("species", "environment", "rgn_id"))

data.frame(filter(mar_sus, rgn_id==14 & species == "Barramundi"))
data.frame(filter(mar_sus, !is.na(Sust_c_sp_env)))

# join taxa specific to country/species/fao region
c_sp_fao = sus_rgn %>% 
  filter(match_type == 'c_sp_fao') %>% 
  select(rgn_id, species=species_fao, fao, Sust_c_sp_fao = Maric_sustainability)

mar_sus <- mar_sus %>%
  left_join(c_sp_fao, by= c("species", "fao", "rgn_id"))

data.frame(filter(mar_sus, rgn_id==218 & species == "Atlantic salmon"))
data.frame(filter(mar_sus, !is.na(Sust_c_sp_fao)))

# join taxa specific to country/species
c_sp = sus_rgn %>% 
  filter(match_type == 'c_sp') %>% 
  select(rgn_id, species=species_fao, Sust_c_sp = Maric_sustainability)

mar_sus <- mar_sus %>%
  left_join(c_sp, by= c("species", "rgn_id"))
summary(mar_sus)

### merge these into a single sustainability score
mar_sus = mar_sus %>% 
  mutate(Sust = ifelse(!is.na(Sust_c_sp_env), Sust_c_sp_env, Sust_c_sp_fao)) %>%
  mutate(Sust = ifelse(is.na(Sust), Sust_c_sp, Sust)) %>%
  select(-Sust_c_sp_env, -Sust_c_sp_fao, -Sust_c_sp)

```


This joins the sustainability data that is gapfilled either at the species level (average of specific species/genera across regions) or at a higher course taxonomic levels and documents which data are gapfilled and how.
```{r}

## Gapfilled at the species/genera level:
gf_sp_sus <- filter(sus_rgn, gapfill != "actuals" & match_type == "species") %>%
  select(species = species_fao, gapfill, Sust_gf_sp = Maric_sustainability)

## check that there are no duplicated species_fao
gf_sp_sus[duplicated(gf_sp_sus$species), ]

# Match gapfilling values by species
mar_sus_gf = mar_sus %>%
  left_join(gf_sp_sus, by = 'species')


# Gapfilled at the coarse taxon level:
gf_taxon_sus <- filter(sus_rgn, gapfill != "actuals" & match_type == "taxon") %>%
  select(Taxon_code=taxon, Sust_gf_taxon = Maric_sustainability)

# Match gapfilling values by species
mar_sus_gf = mar_sus_gf %>%
  left_join(gf_taxon_sus, by = c('Taxon_code'))

summary(mar_sus_gf)
table(mar_sus_gf$gapfill)


#Obtain a sustainability score for each record, and a book-keeping column of whether it's actual or gap-filled
mar_sus_final = mar_sus_gf %>% 
  mutate(gapfill = ifelse(!is.na(Sust), "none", gapfill)) %>% 
  mutate(Sust = ifelse(is.na(Sust), Sust_gf_sp, Sust)) %>%
  mutate(gapfill = ifelse(is.na(Sust), "taxon_average", gapfill)) %>%
  mutate(Sust = ifelse(is.na(Sust), Sust_gf_taxon, Sust)) %>%
  mutate(taxa_code = paste(species, species_code, sep="_")) %>%
  select(rgn_id, species, species_code, taxa_code, year, gapfill_sus = gapfill, gapfill_fao = gap_0_fill, tonnes=value, Sust)


## save data layers
mar_harvest_tonnes = mar_sus_final %>%
  select(rgn_id, taxa_code, year, tonnes)
anyDuplicated(mar_harvest_tonnes)
write.csv(mar_harvest_tonnes, 'output/mar_harvest_tonnes.csv', row.names=F)

mar_harvest_tonnes_gf = mar_sus_final %>%
  select(rgn_id, taxa_code, year, tonnes=gapfill_fao)
write.csv(mar_harvest_tonnes, 'output/mar_harvest_tonnes_gf.csv', row.names=F)


mar_sustainability_score = mar_sus_final %>% 
  select(rgn_id, taxa_code, sust_coeff = Sust) %>% 
  unique()
anyDuplicated(mar_sustainability_score)
write.csv(mar_sustainability_score, 'output/mar_sustainability.csv', row.names=F)

mar_sustainability_score_gf = mar_sus_final %>% 
  select(rgn_id, taxa_code, sust_coeff = gapfill_sus) %>% 
  unique()
write.csv(mar_sustainability_score, 'output/mar_sustainability_gf.csv', row.names=F)

```

# Genetic escapes data

These data are used as a pressure layer to describe the risk of genetic escapees due to mariculture.

First merge with the species data (no gapfilling) for each country/species/fao region combination.

```{r}
# can eliminate the environment category because these have the same scores
esc = sus_rgn %>% 
  filter(!is.na(Genetic.escapees)) %>%
  mutate(match_type = ifelse(match_type == "c_sp_env", "c_sp", match_type)) %>%
  group_by(rgn_id, species=species_fao, fao, match_type, taxon, gapfill) %>%
  summarize(Genetic.escapees = mean(Genetic.escapees)) %>%
  ungroup()
  
# join taxa specific to country/species/fao
c_sp_fao = esc %>% 
  filter(match_type == 'c_sp_fao') %>% 
  select(rgn_id, species, fao, Esc_c_sp_fao = Genetic.escapees) 

mar_esc <- maric %>%
  left_join(c_sp_fao, by= c("species", "fao", "rgn_id"))

data.frame(filter(mar_esc, !is.na(Esc_c_sp_fao)))

# join taxa specific to country/species
c_sp = esc %>% 
  filter(match_type == 'c_sp') %>% 
  select(rgn_id, species, Esc_c_sp = Genetic.escapees)

mar_esc <- mar_esc %>%
  left_join(c_sp, by= c("species", "rgn_id"))
summary(mar_esc)

### merge these into a single sustainability score
mar_esc = mar_esc %>% 
  mutate(Escapees = ifelse(!is.na(Esc_c_sp_fao), Esc_c_sp_fao, Esc_c_sp)) %>%
  select(-Esc_c_sp_fao, -Esc_c_sp)

```

Join the sustainability data that is gapfilled either at the species level (average of specific species/genera across regions) or at a higher course taxonomic levels and documents which data are gapfilled and how.
```{r}
## Gapfilled at the species/genera level:
gf_species_esc <- filter(esc, gapfill != "actuals" & match_type == "species") %>%
  select(species, gapfill, Esc_gf_sp = Genetic.escapees)

## check that there are no duplicated species_fao
gf_species_esc[duplicated(gf_species_esc$species), ]

# Match gapfilling values by species
mar_esc_gf = mar_esc %>%
  left_join(gf_species_esc, by = 'species')


# Gapfilled at the coarse taxon level:
gf_taxon_sus <- filter(esc, gapfill != "actuals" & match_type == "taxon") %>%
  select(Taxon_code=taxon, Esc_gf_taxon = Genetic.escapees)

# Match gapfilling values by species
mar_esc_gf = mar_esc_gf %>%
  left_join(gf_taxon_sus, by = c('Taxon_code'))

summary(mar_esc_gf)
table(mar_esc_gf$gapfill)


#Obtain a sustainability score for each record, and a book-keeping column of whether it's actual or gap-filled
tonnes_esc = mar_esc_gf %>% 
  mutate(gapfill = ifelse(!is.na(Escapees), "none", gapfill)) %>% 
  mutate(Escapees = ifelse(is.na(Escapees), Esc_gf_sp, Escapees)) %>%
  mutate(gapfill = ifelse(is.na(Escapees), "taxon_average", gapfill)) %>%
  mutate(Escapees = ifelse(is.na(Escapees), Esc_gf_taxon, Escapees)) %>%
  select(rgn_id, species, species_code, year, gapfill_escapees = gapfill, tonnes=value, Escapees)

summary(tonnes_esc)

```


Final formatting of the escapee data. This is used as a pressure layer.
```{r}
# for each region/year: average the genetic escape probability for each taxa based on tonnes mariculture
genEscapes <- tonnes_esc %>%
  group_by(rgn_id, year) %>%
  summarize(genEscapes = weighted.mean(Escapees, tonnes, na.rm=TRUE))

# obtain corresponding gapfilling information for each region (average of gapfilled data, weighted by tonnes of mariculture).
genEscapes_gf <- tonnes_esc %>%
  mutate(gapfill = ifelse(gapfill_escapees=="none", 1, 0)) %>%
  group_by(rgn_id, year) %>%
  summarize(genEscapes = weighted.mean(gapfill, tonnes, na.rm=TRUE)) %>%
  ungroup() %>%
  filter(year==2014) %>%
  select(rgn_id, pressures.score=genEscapes) %>%
  mutate(pressures.score=ifelse(pressures.score=="NaN", NA, pressures.score)) %>%
  data.frame()
write.csv(genEscapes_gf, 'output/GenEsc_v2016_gf.csv', row.names=FALSE)

# create the escapee data layers:
for(year in 2012:2016){ #year=2016
data_year <- year-2
    data <- genEscapes %>%
    filter(year == data_year) %>%
    select(rgn_id, pressure_score = genEscapes)
write.csv(data, sprintf('output/GenEsc_v%s.csv', year), row.names=FALSE)  
}


```
***

