---
title: 'OHI: Mariculture '
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

[REFERENCE RMD FILE: https://cdn.rawgit.com/OHI-Science/ohiprep/master/globalprep/mar/v2017/mar_dataprep.html]

#Summary
This analysis converts FAO mariculture data into data used to calculate the OHI global mariculture status score.  This also calculates the genetic escapee from mariculture pressure data.

#Updates from previous assessment
New year of FAO mariculture yield data, but no changes to sustainability or genetic escapee data or general methods.

A few small corrections to make sure the FAO_species fields matched in the species_list and Trujillo data.

***

#Data Source 
**Reference**:      
    http://www.fao.org/fishery/statistics/software/fishstatj/en#downlApp
     Release date: July 2017 
FAO Global Aquaculture Production Quantity 1950_2014

**Downloaded**: 8/10/2017

**Description**:  Quantity (tonnes) of mariculture for each county, species, year.

**Time range**: 1950-2015 

***
  
#Methods
```{r}
# load libraries, set directories
library(ohicore)  #devtools::install_github('ohi-science/ohicore@dev')
library(dplyr)
library(stringr)
library(tidyr)

## comment out when knitting
#setwd("globalprep/mar/v2017")


### Load FAO-specific user-defined functions
source('mar_fxs.R') # functions specific to mariculture dealing with compound countries
source('../../../src/R/fao_fxn.R') # function for cleaning FAO files
source('../../../src/R/common.R') # directory locations

```

# FAO Mariculuture data

Clean mariculture data: Filter freshwater mariculture, make long format, and clean FAO codes.
```{r}

mar <- read.csv(file.path(dir_M, 'git-annex/globalprep/_raw_data/FAO_mariculture/d2017/FAO_GlobalAquacultureProduction_Quantity_1950_2015.csv'), check.names=FALSE, stringsAsFactors=FALSE) ; head(mar) 

# mar <- read.csv('D:/git-annex/globalprep/_raw_data/FAO_mariculture/d2017/FAO_GlobalAquacultureProduction_Quantity_1950_2015.csv', check.names=FALSE, stringsAsFactors=FALSE) ; head(mar) 

mar <- mar %>%
  rename(country = `Country (Country)`,
         FAO_name = `Species (ASFIS species)`,
         fao = `Aquaculture area (FAO major fishing area)`,
         environment = `Environment (Environment)`)
table(mar$environment)  

# include only marine environments
mar <- mar %>%
filter(environment %in% c("Brackishwater", "Marine"))  

## long format and clean FAO codes:
mar <- mar %>%
  select(-Unit) 

## for some reason, I can't provide the data range in gather programatically!
col_num <- ncol(data.frame(mar))
range <- 5:col_num
mar <- mar %>%
  gather(key="year", value="value", 5:70) %>%
    fao_clean_data() 

```

Update species names and exclude non-food species. I simplified the species_list.  I cut the "species"" name columns because it wasn't clear what this was trying to accomplish and created potential error.  
```{r}

mar_sp <- read.csv('raw/species_list.csv', stringsAsFactors=FALSE) %>%
  select(FAO_name, exclude, alias, Taxon_code)
new.spp <- setdiff(mar$FAO_name, mar_sp$FAO_name)
new.spp # check: if dim has 0 rows it means all match
# if length(new.spp) >0 , hand check whether to keep (exclude seaweeds and species harvested for ornamental/medicinal), check if synonyms match Trujillo names

# Remove species not relevant to mariculuture goal (i.e., non-food species)
mar <- left_join (mar, mar_sp, by="FAO_name") 
mar <- filter (mar, exclude==0) 


# change names using species name and the species alias (global changes)
mar$species <- ifelse(!is.na(mar$alias), mar$alias, mar$FAO_name) 

# sum duplicates after name change (this also gets rid of the NA values)
mar <- mar %>%
  filter(!is.na(value)) %>%
  group_by(country, fao, environment, species, year, Taxon_code) %>%
    summarize(value = sum(value)) %>% 
  ungroup()


# eliminate time-series with all 0s
mar <- mar %>%
  group_by(country, species) %>%
  mutate(total_value = sum(value)) %>%
  filter(total_value > 0) %>%
  select(-total_value) %>%
  ungroup()

```

Convert country names to OHI regions

```{r}

# Divide mariculture from countries that we report as separate regions (assume equal production in all regions)
# Netherlands Antilles: Conch restoration among Aruba, Bonaire, Curacao
# Channel Islands: Jersey and Guernsey
# Bonaire/S.Eustatius/Saba
# Yugoslavia SFR: no longer a country after 1992

mar <- mar %>%
  mutate(country = ifelse(country=="R\xe9union", "Reunion", country)) %>%  # this one is hard to get right
  mar_split()  # function in mar_fxs.R

mar_rgn <- name_2_rgn(df_in = mar, 
                       fld_name='country', 
                       flds_unique=c('species', 'fao', 'environment', 'Taxon_code', 'year'))


### sum values of regions with multiple subregions
mar_rgn <- mar_rgn %>%
  group_by(fao, environment, species, year, Taxon_code, rgn_id) %>%
  summarize(value = sum(value)) %>%
  ungroup()

data.frame(filter(mar_rgn, rgn_id==130) %>%
  filter(year==2013) %>%
  arrange(species))

```

For some regions, a specific species can be altered so that it matches more general Trujillo sustainability data.  In this case, I don't want the name changes to be global because some regions may have more specific species data.

(Will explore this in the future, but will not implement this year).
```{r, }
# ## based on looking at the list, make a few name changes to match the regions Trujillo data
# # Chile name modification
# mar_rgn$species[mar_rgn$rgn_id==224 & mar_rgn$species == "Red abalone"] <- "Abalones nei"
# mar_rgn$species[mar_rgn$rgn_id==224 & mar_rgn$species == "Japanese abalone"] <- "Abalones nei"
# 
# # China
# mar_rgn$species[mar_rgn$rgn_id==209 & mar_rgn$species == "Areolate grouper"] <- "Groupers nei"
# mar_rgn$species[mar_rgn$rgn_id==209 & mar_rgn$species == "Greasy grouper"] <- "Groupers nei"
# mar_rgn$species[mar_rgn$rgn_id==209 & mar_rgn$species == "Hong Kong grouper"] <- "Groupers nei"
# mar_rgn$species[mar_rgn$rgn_id==209 & mar_rgn$species == "Orange-spotted grouper"] <- "Groupers nei"
# 
# # Honduras
# mar_rgn$species[mar_rgn$rgn_id==133 & mar_rgn$species == "Whiteleg shrimp"] <- "Penaeus shrimps nei"
# 
# # Italy
# mar_rgn$species[mar_rgn$rgn_id==84 & mar_rgn$species == "Pacific cupped oyster"] <- "Cupped oysters nei"
# 
# # New Zealand
# mar_rgn$species[mar_rgn$rgn_id==162 & mar_rgn$species == "Rainbow abalone"] <- "Abalones nei"
# 
# # Pakistan
# mar_rgn$species[mar_rgn$rgn_id==53 & mar_rgn$species == "Penaeus shrimps nei"] <- "Marine crustaceans nei"
# 
# # Philippines
# mar_rgn$species[mar_rgn$rgn_id==15 & mar_rgn$species == "Whiteleg shrimp"] <- "Penaeus shrimps nei"
# 
# # Portugal
# mar_rgn$species[mar_rgn$rgn_id==183 & mar_rgn$species == "Golden carpet shell"] <- "Marine molluscs nei"
# mar_rgn$species[mar_rgn$rgn_id==183 & mar_rgn$species == "Peppery furrow"] <- "Marine molluscs nei"
# mar_rgn$species[mar_rgn$rgn_id==183 & mar_rgn$species == "Solen razor clams nei"] <- "Razor clams nei"
# mar_rgn$species[mar_rgn$rgn_id==183 & mar_rgn$species == "Atlantic bluefin tuna"] <- "Marine fishes nei"
# mar_rgn$species[mar_rgn$rgn_id==183 & mar_rgn$species == "Meagre"] <- "Marine fishes nei"
# mar_rgn$species[mar_rgn$rgn_id==183 & mar_rgn$species == "Seabasses nei"] <- "Marine fishes nei"
# mar_rgn$species[mar_rgn$rgn_id==183 & mar_rgn$species == "Soles nei"] <- "Marine fishes nei"
# mar_rgn$species[mar_rgn$rgn_id==183 & mar_rgn$species == "White seabream"] <- "Marine fishes nei"
# 
# # Spain
# mar_rgn$species[mar_rgn$rgn_id==182 & mar_rgn$species == "Atlantic bluefin tuna"] <- "Tuna-like fishes nei"
# 
# # Turkey
# mar_rgn$species[mar_rgn$rgn_id==76 & mar_rgn$species == "Atlantic bluefin tuna"] <- "Tuna-like fishes nei"
# mar_rgn$species[mar_rgn$rgn_id==76 & mar_rgn$species == "European seabass"] <- "Seabasses nei"
# 
# ### sum values of regions with multiple subregions
# mar_rgn <- mar_rgn %>%
#   group_by(fao, environment, species, year, Taxon_code, rgn_id) %>%
#   summarize(value = sum(value)) %>%
#   ungroup()


```

### Gapfilling
Fill in missing years after first harvest with 0 values

```{r}

mar_rgn_spread <- spread(mar_rgn, year, value)
dim(mar_rgn_spread)
mar_rgn_gf <- gather(mar_rgn_spread, "year", "value", 6:71) %>%
  arrange(rgn_id, species, year, Taxon_code, fao, environment)

## NA values are converted to zero.  I checked to make sure there 
## weren't instances in which in made more sense to carry the previous
## years data forward as a method of gapfilling. This didn't seem to be the case
mar_rgn_gf <- mar_rgn_gf %>%
  mutate(year = as.numeric(as.character(year))) %>%
  mutate(value_w_0 = ifelse(is.na(value), 0, value)) %>%
  group_by(fao, environment, species, Taxon_code, rgn_id) %>%
  mutate(cum_value = cumsum(value_w_0)) %>%
  ungroup() %>%
  filter(cum_value > 0) %>%
  mutate(gap_0_fill = ifelse(is.na(value), "NA_to_zero", "0")) %>%
  mutate(value = ifelse(is.na(value), 0, value)) %>%
  select(-cum_value, -value_w_0)
table(mar_rgn_gf$gap_0_fill)
## 3344 of these out of 25842+3344 cases

```

Remove time series with less than four non-zero datapoints (assume these are not established mariculture programs).
```{r}

mar_rgn_gf = mar_rgn_gf %>% 
  group_by(rgn_id, species, fao, environment) %>%
  mutate (not_0 = length(value[value>0])) %>% 
  filter (not_0>3) %>%
  ungroup() %>% 
  select(rgn_id, species, fao, environment, year, value, Taxon_code, gap_0_fill)

```

Add a unique identifier per cultivated stock that describes species, fao region, and environment.
```{r}

# add a unique identifier per cultivated stock
identifier = mar_rgn_gf %>% 
  select(rgn_id, species, fao, environment) %>% 
  unique() %>% 
  mutate(species_code = 1:n())

mar_rgn_gf = left_join(mar_rgn_gf, identifier)
maric <- mar_rgn_gf

```


Save file to estimate total mariculture yield per country.
```{r}

write.csv(maric, 'output/MAR_FP_data.csv', row.names=FALSE)

```

# Trujillo sustainability scores

These data describe the sustainability and genetic escapes for country/species combinations (and, in a couple cases, environment and fao region combinations). In cases where these data were not available for a specific county/species, we averaged the data across taxonomic groups to gapfill the missing data.

Convert country names to OHI region names.
```{r}

# Trujillo sustainability data:
sus = read.csv('raw/Truj_label_sust.csv', stringsAsFactors = FALSE, na.strings = NA)

## these need to be re-added (get cut when associated with region ids)
sus_no_rgn <- filter(sus, is.na(country))

# convert country names to OHI region names:
sus_rgn <- name_2_rgn(df_in = sus, 
                       fld_name='country', 
                       flds_unique=c('species_fao', 'fao', 'environment', 'species_Truj'))

sus_rgn <- bind_rows(sus_rgn, sus_no_rgn) %>%
  unique()


# check the fao spp list in the Trujillo sustainability file matches FAO mariculture species
setdiff(sus_rgn$species_fao, maric$species) # species that are no longer have mariculture industry or are not included due to being freshwater or non-food

sort(setdiff(maric$species, sus_rgn$species_fao)) # species with no Trujillo data




```

# FAO maricultue and sustainability scores

Match the sustainability score to the FAO mariculture data.

The following joins the sustainability scores to regions/species that have Trujillo data.
```{r}

table(sus_rgn$match_type)

# join taxa specific to country/species/environment
c_sp_env = sus_rgn %>% 
  filter(match_type == 'c_sp_env') %>% 
  select(rgn_id, species=species_fao, environment, Sust_c_sp_env = Maric_sustainability)

mar_sus <- maric %>%
  left_join(c_sp_env, by= c("species", "environment", "rgn_id"))

# join taxa specific to country/species/fao region
c_sp_fao = sus_rgn %>% 
  filter(match_type == 'c_sp_fao') %>% 
  select(rgn_id, species=species_fao, fao, Sust_c_sp_fao = Maric_sustainability)

mar_sus <- mar_sus %>%
  left_join(c_sp_fao, by= c("species", "fao", "rgn_id"))

data.frame(filter(mar_sus, rgn_id==218 & species == "Atlantic salmon"))
data.frame(filter(mar_sus, !is.na(Sust_c_sp_fao)))

# join taxa specific to country/species
c_sp = sus_rgn %>% 
  filter(match_type == 'c_sp') %>% 
  select(rgn_id, species=species_fao, Sust_c_sp = Maric_sustainability)

mar_sus <- mar_sus %>%
  left_join(c_sp, by= c("species", "rgn_id"))
summary(mar_sus)

### merge these into a single sustainability score
mar_sus = mar_sus %>% 
  mutate(Sust = ifelse(!is.na(Sust_c_sp_env), Sust_c_sp_env, Sust_c_sp_fao)) %>%
  mutate(Sust = ifelse(is.na(Sust), Sust_c_sp, Sust)) %>%
  select(-Sust_c_sp_env, -Sust_c_sp_fao, -Sust_c_sp)

```


This joins the sustainability data that is gapfilled either at the species level (average of specific species/genera across regions) or at a higher course taxonomic levels and documents which data are gapfilled and how.
```{r}

## Gapfilled at the species/genera level:
gf_sp_sus <- filter(sus_rgn, gapfill != "actuals" & match_type == "species") %>%
  select(species = species_fao, gapfill, Sust_gf_sp = Maric_sustainability)

## check that there are no duplicated species_fao
gf_sp_sus[duplicated(gf_sp_sus$species), ]

# Match gapfilling values by species
mar_sus_gf = mar_sus %>%
  left_join(gf_sp_sus, by = 'species')


# Gapfilled at the coarse taxon level:
gf_taxon_sus <- filter(sus_rgn, gapfill != "actuals" & match_type == "taxon") %>%
  select(Taxon_code=taxon, Sust_gf_taxon = Maric_sustainability)

# Match gapfilling values by species
mar_sus_gf = mar_sus_gf %>%
  left_join(gf_taxon_sus, by = c('Taxon_code'))

summary(mar_sus_gf)
table(mar_sus_gf$gapfill)


#Obtain a sustainability score for each record, and a book-keeping column of whether it's actual or gap-filled
mar_sus_final = mar_sus_gf %>% 
  mutate(gapfill = ifelse(!is.na(Sust), "none", gapfill)) %>% 
  mutate(Sust = ifelse(is.na(Sust), Sust_gf_sp, Sust)) %>%
  mutate(gapfill = ifelse(is.na(Sust), "taxon_average", gapfill)) %>%
  mutate(Sust = ifelse(is.na(Sust), Sust_gf_taxon, Sust)) %>%
  mutate(taxa_code = paste(species, species_code, sep="_")) %>%
  select(rgn_id, species, species_code, taxa_code, year, gapfill_sus = gapfill, gapfill_fao = gap_0_fill, tonnes=value, Sust)


## save data layers
mar_harvest_tonnes = mar_sus_final %>%
  select(rgn_id, taxa_code, year, tonnes)
anyDuplicated(mar_harvest_tonnes)
write.csv(mar_harvest_tonnes, 'output/mar_harvest_tonnes.csv', row.names=F)

mar_harvest_tonnes_gf = mar_sus_final %>%
  select(rgn_id, taxa_code, year, tonnes=gapfill_fao)
write.csv(mar_harvest_tonnes, 'output/mar_harvest_tonnes_gf.csv', row.names=F)


mar_sustainability_score = mar_sus_final %>% 
  mutate(year = 2012) %>%
  select(rgn_id, year, taxa_code, sust_coeff = Sust) %>% 
  unique()
anyDuplicated(mar_sustainability_score)
write.csv(mar_sustainability_score, 'output/mar_sustainability.csv', row.names=F)

mar_sustainability_score_gf = mar_sus_final %>% 
  select(rgn_id, taxa_code, sust_coeff = gapfill_sus) %>% 
  unique()
write.csv(mar_sustainability_score, 'output/mar_sustainability_gf.csv', row.names=F)

```

# Genetic escapes data

These data are used as a pressure layer to describe the risk of genetic escapees due to mariculture.

First merge with the species data (no gapfilling) for each country/species/fao region combination.

```{r}
# can eliminate the environment category because these have the same scores
esc = sus_rgn %>% 
  filter(!is.na(Genetic.escapees)) %>%
  mutate(match_type = ifelse(match_type == "c_sp_env", "c_sp", match_type)) %>%
  group_by(rgn_id, species=species_fao, fao, match_type, taxon, gapfill) %>%
  summarize(Genetic.escapees = mean(Genetic.escapees)) %>%
  ungroup()
  
# join taxa specific to country/species/fao
c_sp_fao = esc %>% 
  filter(match_type == 'c_sp_fao') %>% 
  select(rgn_id, species, fao, Esc_c_sp_fao = Genetic.escapees) 

mar_esc <- maric %>%
  left_join(c_sp_fao, by= c("species", "fao", "rgn_id"))

data.frame(filter(mar_esc, !is.na(Esc_c_sp_fao)))

# join taxa specific to country/species
c_sp = esc %>% 
  filter(match_type == 'c_sp') %>% 
  select(rgn_id, species, Esc_c_sp = Genetic.escapees)

mar_esc <- mar_esc %>%
  left_join(c_sp, by= c("species", "rgn_id"))
summary(mar_esc)

### merge these into a single sustainability score
mar_esc = mar_esc %>% 
  mutate(Escapees = ifelse(!is.na(Esc_c_sp_fao), Esc_c_sp_fao, Esc_c_sp)) %>%
  select(-Esc_c_sp_fao, -Esc_c_sp)

```

Join the sustainability data that is gapfilled either at the species level (average of specific species/genera across regions) or at a higher course taxonomic levels and documents which data are gapfilled and how.
```{r}
## Gapfilled at the species/genera level:
gf_species_esc <- filter(esc, gapfill != "actuals" & match_type == "species") %>%
  select(species, gapfill, Esc_gf_sp = Genetic.escapees)

## check that there are no duplicated species_fao
gf_species_esc[duplicated(gf_species_esc$species), ]

# Match gapfilling values by species
mar_esc_gf = mar_esc %>%
  left_join(gf_species_esc, by = 'species')


# Gapfilled at the coarse taxon level:
gf_taxon_sus <- filter(esc, gapfill != "actuals" & match_type == "taxon") %>%
  select(Taxon_code=taxon, Esc_gf_taxon = Genetic.escapees)

# Match gapfilling values by species
mar_esc_gf = mar_esc_gf %>%
  left_join(gf_taxon_sus, by = c('Taxon_code'))

summary(mar_esc_gf)
table(mar_esc_gf$gapfill)


#Obtain a sustainability score for each record, and a book-keeping column of whether it's actual or gap-filled
tonnes_esc = mar_esc_gf %>% 
  mutate(gapfill = ifelse(!is.na(Escapees), "none", gapfill)) %>% 
  mutate(Escapees = ifelse(is.na(Escapees), Esc_gf_sp, Escapees)) %>%
  mutate(gapfill = ifelse(is.na(Escapees), "taxon_average", gapfill)) %>%
  mutate(Escapees = ifelse(is.na(Escapees), Esc_gf_taxon, Escapees)) %>%
  select(rgn_id, species, species_code, year, gapfill_escapees = gapfill, tonnes=value, Escapees)

summary(tonnes_esc)

```


Final formatting of the escapee data. This is used as a pressure layer.
```{r}
# for each region/year: average the genetic escape probability for each taxa based on tonnes mariculture
genEscapes <- tonnes_esc %>%
  group_by(rgn_id, year) %>%
  summarize(genEscapes = weighted.mean(Escapees, tonnes, na.rm=TRUE))

# obtain corresponding gapfilling information for each region (average of gapfilled data, weighted by tonnes of mariculture).
genEscapes_gf <- tonnes_esc %>%
  mutate(gapfill = ifelse(gapfill_escapees=="none", 1, 0)) %>%
  group_by(rgn_id, year) %>%
  summarize(genEscapes = weighted.mean(gapfill, tonnes, na.rm=TRUE)) %>%
  ungroup() %>%
  filter(year==2015) %>%
  select(rgn_id, pressures.score=genEscapes) %>%
  mutate(pressures.score=ifelse(pressures.score=="NaN", NA, pressures.score)) %>%
  data.frame()
write.csv(genEscapes_gf, 'output/GenEsc_gf.csv', row.names=FALSE)

# create the escapee data layers:
    data <- genEscapes %>%
    select(rgn_id, year, pressure_score = genEscapes)
write.csv(data, 'output/GenEsc.csv', row.names=FALSE)  

old <- read.csv("../v2016/output/GenEsc_v2016.csv") %>%
  select(rgn_id, prs_score_old=pressure_score)

new <- read.csv("../v2017/output/GenEsc.csv") %>%
  filter(year == 2014)

full_join(old, new, by="rgn_id")

tmp_old <- read.csv("../v2016/raw/Truj_label_sust.csv")
tmp_new <- read.csv("../v2017/raw/Truj_label_sust.csv")

test <- full_join(tmp_new, tmp_old, by=c("country", "species_fao", "fao", "environment" ))
```
***

