Liveliehoods and Economies 2012 vs 2013 Report
==============================================

```
From: Benjamin Halpern <halpern@bren.ucsb.edu>
Date: Sun, Feb 16, 2014 at 4:37 PM
Subject: before lab meeting
To: Ben Best <bbest@nceas.ucsb.edu>
Cc: Katie Longo <longo@nceas.ucsb.edu>, Julia Stewart Lowndes <lowndes@nceas.ucsb.edu>, Melanie Frazier <frazier@nceas.ucsb.edu>
```

Hi BB,
As soon as possible, but at the latest by the start of lab meeting on Wednesday, please send me stats on:

1. which countries had ECO change >10 points between 2012 and 2013
2. which countries used georegional gap filled data for revenue vs. their own data in 2012 vs. 2013
3. which countries had LIV change >10 points between 2012 and 2013
4. which countries LIV scores were based on jobs data only vs. jobs + wages data
5. which countries used georegional gap filled data for jobs vs. their own data in 2012 vs. 2013
6. which countries used georegional gap filled data for wages vs. their own data in 2012 vs. 2013

thanks,
Ben


```{r init, echo=FALSE}

# libraries
library(reshape2)
library(plyr)
library(knitr)
suppressPackageStartupMessages(library(dplyr))

# working directory
wd = '~/GitHub_Mac/ohidata/Global/NCEAS-LivelihoodsEconomiesReport_v2013'
dir_local = '/Volumes/local_edit'
dir_data  = '/Volumes/data_edit'
dir_out   = 'report_LE_output'
setwd(wd)
dir.create(dir_out, showWarnings=F)

# get official Radical(rad) results
rad  = read.csv(file.path(dir_local, 'src/toolbox/scenarios/global_2013a/results/OHI_results_for_Radical_2013-10-09.csv'), na.strings='') # head(rad); table(rad[,c('goal','dimension')])

# get region names, including GLOBAL
rgns = read.csv(file.path(dir_data , 'model/GL-NCEAS-OceanRegions_v2013a/data/rgn_details.csv'), na.strings='')  %.%
  select(rgn_id, rgn_nam) %.%
  rbind(data.frame(rgn_id=0, rgn_nam='GLOBAL')) %.%
  rename(c(rgn_id='region_id', rgn_nam='region_name')) # head(rgns); tail(rgns)

# merge Radical results with region name
rad = rad %.%
  merge(rgns) # head(rad); tail(rad)

# get status metric actuals
status_metric_2012 = read.csv(file.path(dir_local, 'src/model/global2013/livelihoods/status/data_2012/global_li_status_model_combined.csv'), na.strings='') # head(status_metric_2012)
status_metric_2013 = read.csv(file.path(dir_local, 'src/model/global2013/livelihoods/status/data_2013/global_li_status_model_combined.csv'), na.strings='') # head(status_metric_2013)

# get 2013 region names per Nature 2012 country
rgn_country = read.csv(file.path(dir_data, 'model/GL-NCEAS-OceanRegions_v2013a/manual_output/rgn2013_to_country2012.csv'), na.strings='') # head(rgn_country)
rgn_country = rgn_country %.%
  select(rgn_id, country_id) %.%
  rename(c(rgn_id='region_id', country_id='iso3166')) %.%
  merge(rgns, by='region_id')

# update status_metrics with 2013 region ids and names
status_metric_2012 = merge(status_metric_2012, rgn_country, by='iso3166')
status_metric_2013 = merge(status_metric_2013, rgn_country, by='iso3166')

# info function for comparing
info = function(x){ print(head(x)); print(dim(x)); print(summary(x)) }
```

## 1. ECO score change >10 points between 2012 and 2013
```{r q1, results='asis', echo=FALSE}

# query
q1 = rad %.%
  filter(goal=='ECO', dimension=='score') %.%
  dcast(region_id + region_name ~ scenario, value.var='value') %.%
  rename(c('2012'='score_2012', '2013'='score_2013')) %.%
  mutate(dif_abs = abs(score_2013 - score_2012)) %.%
  filter(dif_abs > 10) %.%
  arrange(desc(dif_abs))

# write
write.csv(q1, file.path(dir_out, 'q1.csv'))

# display
row.names(q1) = 1:nrow(q1)
kable(q1, format='html', row.names=T)
```

## 2. Revenue georegional gap filled for 2012, but actuals for 2013
```{r q2, results='asis', echo=FALSE}

# query
q2 = anti_join(status_metric_2013, status_metric_2012, by=c('metric','iso3166','region_id','region_name')) %.%
  filter(metric=='rev') %.%
  arrange(iso3166, region_name) %.%
  select(iso3166, region_id, region_name, score, n_sector)

# write
write.csv(q2, file.path(dir_out, 'q2.csv'))

# display
row.names(q2) = 1:nrow(q2)
kable(q2, format='html', row.names=T)
```

## 3. LIV change >10 points between 2012 and 2013
```{r q3, results='asis', echo=FALSE}
# query
q3 = rad %.%
  filter(goal=='LIV', dimension=='score') %.%
  dcast(region_id + region_name ~ scenario, value.var='value') %.%
  rename(c('2012'='score_2012', '2013'='score_2013')) %.%
  mutate(dif_abs = abs(score_2013 - score_2012)) %.%
  filter(dif_abs > 10) %.%
  arrange(desc(dif_abs))

# write
write.csv(q3, file.path(dir_out, 'q3.csv'))

# display
row.names(q3) = 1:nrow(q3)
kable(q3, format='html', row.names=T)
```

## 4. LIV scores were based on jobs data only vs. jobs + wages data

### 2012
```{r q4_2012, results='asis', echo=FALSE}

q4_2012 = status_metric_2012 %.%
  dcast(iso3166 + region_id + region_name ~ metric, value.var='score') %.%
  filter(!is.na(jobs) & is.na(wage)) %.%
  select(iso3166, region_id, region_name, jobs, wage)

# write
write.csv(q4_2012, file.path(dir_out, 'q4_2012.csv'))

# display
row.names(q4_2012) = 1:nrow(q4_2012)
kable(q4_2012, format='html', row.names=T)
```

### 2013
```{r q4_2013, results='asis', echo=FALSE}

q4_2013 = status_metric_2013 %.%
  dcast(iso3166 + region_id + region_name ~ metric, value.var='score') %.%
  filter(!is.na(jobs) & is.na(wage)) %.%
  select(iso3166, region_id, region_name, jobs, wage)

# write
write.csv(q4_2013, file.path(dir_out, 'q4_2013.csv'))

# display
row.names(q4_2013) = 1:nrow(q4_2013)
kable(q4_2013, format='html', row.names=T)
```

## 5. Jobs georegional gap filled for 2012, but actuals for 2013
```{r q5, results='asis', echo=FALSE}

# query
q5 = anti_join(status_metric_2013, status_metric_2012, by=c('metric','iso3166','region_id','region_name')) %.%
  filter(metric=='jobs') %.%
  select(iso3166, region_id, region_name, score, n_sector)

# write
write.csv(q5, file.path(dir_out, 'q5.csv'))

# display
kable(q5, format='html', row.names=T)
```
None

## 6. Wages georegional gap filled for 2012, but actuals for 2013
```{r q6, results='asis', echo=FALSE}

# query
q6 = anti_join(status_metric_2013, status_metric_2012, by=c('metric','iso3166','region_id','region_name')) %.%
  filter(metric=='wage') %.%
  arrange(iso3166, region_name) %.%
  select(iso3166, region_id, region_name, score, n_sector)

# write
write.csv(q6, file.path(dir_out, 'q6.csv'))

# display
row.names(q6) = 1:nrow(q6)
kable(q6, format='html', row.names=T)
```
