---
title: "Select EEZs from Global Shapefile"
output: html_document
---

#Setup

```{r setup, include=FALSE}

source('~/github/ohiprep/src/R/common.R')

library(raster)
library(rgdal)


```

#Read in global shapefile

You will have to point the path to where the global OHI EEZ file is located on your local computer.

```{r}

#use readOGR to read in the global shapefile.
## dsn: path name, layer: name of the shapefile

eezs = readOGR(dsn = file.path(dir_M,'git-annex/globalprep/spatial/v2015/data'), layer = 'regions_gcs')


```

#Look at global data

```{r}
head(eezs@data)
```

#Specify countries

```{r}

countries = c("Sweden", "Finland", "Denmark", "Germany", "Estonia", "Poland", "Latvia", "Lithuania", "Russia")

```

#Select EEZs and remove all others

```{r}

myShape <- eezs%>%subset(rgn_nam%in%countries)%>%subset(rgn_typ=='eez')

plot(myShape)

```

#Draw extent to crop

You can use raster::drawExtent() to draw the bounding box for cropping your shapefile if you don't have a predefined extent. Otherwise use extent() to set it.

```{r}

ext <- drawExtent(show=TRUE,col='red')

#ext <- extent()

ext

myShape <- crop(myShape,ext)


```

#Save shapefile

Use rgdal::writeOGR() to save the shapefile. Set the driver equal to ESRI Shapefile.

```{r save}

writeOGR(myShape,dsn='globalprep/spatial',layer='bhi_global_eezs', driver = "ESRI Shapefile")

```











