---
title: "OHI 2015: Marine Plastics Layer"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohiprep/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

#Summary

This data was used in the Clean Waters goal for the 2015 global Ocean Health Index assessment

***

#Data Source

**Reference**: [Eriksen et al. 2014](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0111913)

**Downloaded**: December 10, 2014 directly from authors

**Native Data Resolution**:   

**Values**: Count (number/km2) and weight (g/km2) across 4 size classes  

**Time Range**: N/A

**Format**: GeoTIFF

***

# Methods
```{r setup, message=F,warning=F}
# set tmp directory

tmpdir='~/big/R_raster_tmp'
dir.create(tmpdir, showWarnings=F)

# paths
    dir_M = c('Windows' = '//mazu.nceas.ucsb.edu/ohi',
              'Linux'   = '/home/shares/ohi')[[ Sys.info()[['sysname']] ]]

#libraries

library(raster)
library(rgdal)
library(rasterVis)
library(RColorBrewer)

#set colors for plotting

cols = rev(colorRampPalette(brewer.pal(11, 'Spectral'))(255)) # rainbow color scheme

#use my theme for levelplot
mytheme <- rasterTheme(region = rev(colorRampPalette(brewer.pal(11, 'Spectral'))(255)))

#set working directory on neptune
data_wd = file.path(dir_M,'git-annex/globalprep/cw_pressure_trash')
```

## Read in data

```{r read data, warning=F,message=F}
#bring in ocean raster to clip out land
ocean = raster(file.path(dir_M,'model/GL-NCEAS-Halpern2008/tmp/ocean.tif'))


# plastics data
count  = list.files(path=file.path(data_wd,'v2015/globalplastic_wd_cd_rasters_180'),pattern='count_*',full.names=T)
weight = list.files(path=file.path(data_wd,'v2015/globalplastic_wd_cd_rasters_180'),pattern='weight_*',full.names=T)

# There are 3 extra weight files from the data source. For sizes 2-4, there is a file with a '2' at the end of it. These are the rasters
# that work, while the other three do not. As an example:

#raster(file.path(data_wd,'v2015/globalplastic_wd_cd_rasters_180/weight_density_size2_180.tif')) #gives an error
#but
raster(file.path(data_wd,'v2015/globalplastic_wd_cd_rasters_180/weight_density_size2_180_2.tif')) #works! These are the ones we will use, along with weight size 1
# which appears to work
```

## Unlog data

Data came to use logged (using base 10 log) so need to 'unlog' the data first 

```{r unlog, message=F,warning=F,eval=F}

  unlog = function(file){

  name = unlist(strsplit(file,'/','.'))[3] #split filename, grab second string to use in naming tif
  r = raster(file)
  out = 10^r
  
  writeRaster(out,filename=paste0(data_wd,'v2015/tmp/unlog/unlog_',name,sep=''),overwrite=T,format='GTiff')
  
}

    sapply(count,unlog)
    sapply(weight,unlog)
```

## Stack all rasters within weight and count
```{r stack, message=F,warning=F}
    weight = stack(list.files(file.path(data_wd,'v2015/tmp/unlog'),'unlog_weight_*',full.names=T))
    count  = stack(list.files(file.path(data_wd,'v2015/tmp/unlog'),'unlog_count_*',full.names=T))

```

## Sum across 4 datasets
```{r sum, message=F,warning=F}
    w_sum = calc(weight,fun=sum)
    c_sum = calc(count,fun=sum)
```

## Reproject to mollweide
```{r reproject, message=F, warning=F, eval=F}

#defining initial projection. Helps avoid errors when reprojecting to mollweide
    projection(w_sum) <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"  


    projection(c_sum) <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"  
    mollCRS <- CRS('+proj=moll') #set mollweide CRS


    w_moll <- projectRaster(w_sum, crs=mollCRS,over=T,progress='text')#,filename='v2015/tmp/weight_sum_moll.tif',overwrite=T)

    c_moll <- projectRaster(c_sum, crs=mollCRS,over=T,progress='text')#,filename='v2015/tmp/count_sum_moll.tif',overwrite=T)

```

## Resample to 1km 
```{r resample, message=F,warning=F,eval=F}
    resamp_w   <- resample(w_moll,ocean,progress='text',filename=file.path(data_wd,'v2015/tmp/weight_sum_moll_1km.tif'),overwrite=T,method='ngb')

    resamp_c   <- resample(c_moll,ocean,progress='text',filename=file.path(data_wd,'v2015/tmp/count_sum_moll_1km.tif'),overwrite=T,method='ngb')

```

## Read in mollweide rasters
```{r moll rasters,  message=F, warning=F}
count_moll      = raster(file.path(data_wd,'v2015/tmp/count_sum_moll.tif'))
count_moll_1km  = raster(file.path(data_wd,'v2015/tmp/count_sum_moll_1km.tif'))
weight_moll     = raster(file.path(data_wd,'v2015/tmp/weight_sum_moll_1km.tif'))
weight_moll_1km = raster(file.path(data_wd,'v2015/tmp/weight_sum_moll_1km.tif'))

```

## Mask out ocean
```{r mask ocean,warning=F,message=F}
#     count_mask = mask(count_moll_1km,ocean,progress='text',filename = file.path(data_wd,'v2015/tmp/count_sum_moll_1km_clip.tif'),overwrite=T)
#     weight_mask = mask(weight_moll_1km,ocean,progress='text',filename=file.path(data_wd,'v2015/tmp/weight_sum_moll_1km_clip.tif'),overwrite=T)

    count_mask  = raster(file.path(data_wd,'v2015/tmp/count_sum_moll_1km_clip.tif'))
    weight_mask = raster(file.path(data_wd,'v2015/tmp/weight_sum_moll_1km_clip.tif'))
    
    plot(count_mask,col=cols,main='Count density (pieces/km2)')
    plot(weight_mask,col=cols,main='Weight density (g/km2)')
```

## Log Transform
```{r log, warning=F,message=F}

#     w_log = calc(weight_mask,fun=function(x){log(x+1)},filename=file.path(data_wd,'v2015/tmp/weight_sum_moll_1km_clip_log.tif'),overwrite=T)
#     c_log = calc(count_mask,fun=function(x){log(x+1)},filename=file.path(data_wd,'v2015/tmp/count_sum_moll_1km_clip_log.tif'),overwrite=T)

    w_log = raster(file.path(data_wd,'v2015/tmp/weight_sum_moll_1km_clip_log.tif'))
    c_log = raster(file.path(data_wd,'v2015/tmp/count_sum_moll_1km_clip_log.tif'))
    
    plot(w_log,col=cols,main='Weight density\nlog(g/km2)')
    plot(c_log,col=cols,main='Count density\nlog(pieces/km2)')
```

## Reference point 

The reference point is the 99.99th quantile of the data.
```{r quantile,message=F,warning=F,eval=F}
w_ref = quantile(w_log,prob=c(0.001,0.01,0.1,0.25,0.5,0.75,0.9,0.99,0.999,0.9999))
c_ref = quantile(c_log,prob=c(0.001,0.01,0.1,0.25,0.5,0.75,0.9,0.99,0.999,0.9999))

w_99 = w_ref[10]
c_99 = c_ref[10]

histogram(w_log,main='Weight density (log(g/km2))')
histogram(c_log,main='Count density (log(pieces/km2))')
```

## Rescale

Using the reference point, rescale the data so that all values are between 0 and 1.
```{r rescale, message=F,warning=F}
# w_rescale = calc(w_log,fun=function(x){ifelse(x>w_99,1,x/w_99)},filename=file.path(data_wd,'v2015/output/weight_rescale.tif'),overwrite=T)

w_rescale = raster(file.path(data_wd,'v2015/output/weight_rescale.tif'))
#raster:plot(w_rescale,main='Pressure layer: weight density (g/km2)',col=cols)

# c_rescale = calc(c_log,fun=function(x){ifelse(x>c_99,1,x/c_99)},filename=file.path(data_wd,'v2015/output/count_rescale.tif'),overwrite=T)

c_rescale = raster(file.path(data_wd,'v2015/output/count_rescale.tif'))
#raster:plot(c_rescale,main='Pressure layer: count density (pieces/km2)',col=cols)
```

## Compare

Compare the differences between weight and count data to help decide what layer to use

```{r compare,message=F,warning=FALSE}

# This provides some general statistics and visualizations:
compare <- stack(w_rescale, c_rescale) 
pairs(compare)    

## Here is a scatter plot comparison:
#plot(w_rescale, c_rescale, ylab="Count", xlab="Weight", maxpixels=10000000, col=rgb(0,0,0,0.2))   
```
