---
title: 'OHI: SPP'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohiprep/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

[REFERENCE RMD FILE: https://rawgit.com/OHI-Science/ohiprep/master/globalprep/Pressures_OceanAcidification/v2015/oa_create_layer_2015.html]

#Summary

Spatial data from IUCN and Aquamaps is combined with extinction risk information from IUCN to generate regional scores for the Species subgoal.  A region's status is based upon an area-weighted average of species health across each global reporting region.

From Halpern et al (2012):

> The target for the Species sub-goal is to have all species at a risk status of Least Concern. We scaled the lower end of the biodiversity goal to be 0 when 75% species are extinct, a level comparable to the five documented mass extinctions and would constitute a catastrophic loss of biodiversity. The Status of assessed species was calculated as the area- and threat status-weighted average of the number of threatened species within each 0.5 degree grid cell.

**Mean risk status per cell:**

$$\bar{R}_{cell} = \frac{\displaystyle\sum_{species}(Risk)}{n_{spp}}$$

**Mean risk status per region:**

$$\bar{R}_{SPP} = \frac{\displaystyle\sum_{cells}(\bar{R}_{cell} * A_{cell} * pA_{cell-rgn})}{A_{rgn}}$$

**Species goal model**

$$X_{SPP} = \frac{((1 - \bar{R}_{SPP}) - 0.25)}{(1 - 0.25)} * 100%$$

where:

* $X_{SPP}$ is Species goal status
* $\bar{R}_{cell}$ is mean extinction risk for one cell
* $\bar{R}_{SPP}$ is area-weighted mean extinction risk for a region
* $A_{cell}$ is cell area
* $pA_{cell-rgn}$ is percent of cell area included in region
* *Risk* is scaled value for species extinction risk category, based on: 
    * 'LC' = 0.0, 'NT' = 0.2, 'VU' = 0.4, 'EN' = 0.6, 'CR' = 0.8, 'EX' = 1.0
* SPP trend is calculated in a similar area-weighted mean, but population trend values are assigned according to:
    * 'Decreasing' = -0.5, 'Stable' = 0.0, 'Increasing' = +0.5

#Updates from previous assessment

Changes since 2015 SPP subgoal for global OHI:

* The 2016 assessment now includes bird species, based upon BirdLife International data.  "Marine" bird species include any bird species with overlap of marine cells. (?) Prior assessments did not include this data.
* Previous assessments counted overlapping species as any species with a binomial included in both data sets (AquaMaps and IUCN/BirdLife).  Synonyms were matched and typos were fixed manually where possible.  This assessment takes advantage of the `taxize` package to identify and match synonyms for better match rate.
* Previous assessments used an AquaMaps threshold of 40% to determine whether a species counted as "present" within a cell.  Based on the IUCN/AquaMaps manuscript, we are now using a 0% threshold for AquaMaps as it more closely resembles the "limits of distribution" criterion used for IUCN inclusion.
* Previous assessments ignored subpopulations for SPP; with IUCN's newest data, a subpopulation field identifies distinct subpopulations (and IUCN risk codes) for a number of species (mostly mammals, but a handful of others as well).  This allows us to incorporate subpopulation scores in the global assessment.

***

#Data Source

IUCN:
* __Reference__: 
    * IUCN 2015. The IUCN Red List of Threatened Species. Version 2015-4. <http://www.iucnredlist.org>.
        * Shapefiles available from: http://www.iucnredlist.org/technical-documents/spatial-data
        * __Downloaded__: 21 December 2015.
    * BirdLife International and NatureServe (2015) Bird species distribution maps of the world. BirdLife International, Cambridge, UK and NatureServe, Arlington, USA.
        * Zipped shapefile available from: http://www.biodiversityinfo.org/spcdownload/r5h8a1/ - username and password via BirdLife International.  
        * __Downloaded__: 14 December 2015.
* __Description__:  Shapefiles containing polygons of assessed species ranges; each shapefile represents all assessed species within a comprehensively-assessed (i.e. >90% assessed) taxonomic group.
* __Native data resolution__: NA
* __Time range__: NA
* __Format__:  Shapefile

AquaMaps:
* __Reference__: 
    * Kaschner, K., J. Rius-Barile, K. Kesner-Reyes, C.Garilao, S.O. Kullander, T. Rees and R. Froese (2013). AquaMaps: Predicted range maps for aquatic species. World wide web electronic publication, www.aquamaps.org, Version 08/2015.
        * __Downloaded__: August 2015.
* __Description__:  .sql files containing information to recreate rasters of cell attributes and cell-by-cell probability of occurrence for 22889 marine species.
* __Native data resolution__: 0.5° latitude and longitude  
* __Time range__: NA
* __Format__:  SQL database files converted into .csv

***
  
#Methods
[R code used to generate the data. Or, alternatively, description of the code/data files used to generate the data.]

***

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = FALSE, message = FALSE, warning = FALSE)

library(foreign)
library(data.table) # for fread()
library(sp)
library(rgdal)
library(raster)
library(maptools)
library(readr)      # for read_csv()

setwd('~/github/ohiprep')
source('src/R/common.R')

goal     <- 'globalprep/SPP_ICO'
scenario <- 'v2016'
dir_anx  <- file.path(dir_neptune_data, 'git-annex', goal) 
dir_data_am    <- file.path(dir_neptune_data, 'git-annex/globalprep/_raw_data', 'aquamaps/v2015') 
dir_data_iucn  <- file.path(dir_neptune_data, 'git-annex/globalprep/_raw_data', 'iucn_spp') 
dir_data_bird  <- file.path(dir_neptune_data, 'git-annex/globalprep/_raw_data', 'birdlife_intl') 
dir_git  <- file.path('~/github/ohiprep', goal)

source(file.path(dir_git, scenario, 'spp_fxn.R'))
# SPP-specific and ICO-specific functions
```


``` {r Ingest Aquamaps data to csv from sql, eval = FALSE}

source(file.path(dir_git, 'R/ingest_aquamaps.R'))
ingest_aquamaps()
```

``` {r Ingest IUCN species list, eval = FALSE}

source(file.path(goal, 'R/ingest_iucn.R'))
### Pulls species list from IUCN
### Scrapes data from iucn site for each species on the list; saves to cache directory.
### Extracts habitats from cached files; filters IUCN list down to just marine species.
### Fixes older IUCN codes, cleans up scinames, gets rid of infraranks (subspecies).
### Determines subpopulations/parents and population trends from scraped files.
### Main outputs, saved to git-annex/globalprep/SPP_ICO/v201x/int: 
### * spp_iucn_all.csv      - full list of IUCN species pulled from web, some cleaning.
### * spp_iucn_habitats.csv - list of IUCN species (by iucn_sid) and corresponding habitat.
### * spp_iucn_marine.csv   - prepped list: cleaned marine list with subpops and trends.
```

``` {r Generate spp lookup}

spp_all <- create_spp_master_lookup(source_pref = 'iucn', fn_tag = '', reload = TRUE)
### | am_sid | sciname | am_category | iucn_sid | iucn_category | popn_trend | popn_category | 
### | info_source | spp_group | id_no | objectid | spatial_source | category_score | trend_score |
### NOTE: as currently stands, ignores AquaMaps-mapped species with 
###   subpopulations (fix_am_subpops() is disabled)

spp_all_subpops1 <- spp_all %>%
  filter(!is.na(iucn_subpop)) %>%
  unique()
spp_all_subpops2 <- spp_all %>%
  filter(str_detect(spatial_source, 'subpop') | str_detect(spatial_source, 'parent')) %>%
  group_by(sciname) %>%
  filter(sum(!is.na(iucn_subpop)) > 0) %>%
  unique()

spp_all_spatially_extinct <- spp_all %>%
  filter(presence == 5)

spp_all1 <- spp_all %>%
  filter(!(str_detect(spatial_source, 'iucn_subpop') & is.na(iucn_subpop))  |    ### subpop tag with no subpop location = FALSE
         is.na(spatial_source)) %>%                                              ### spatial_source is NA = TRUE (keep NAs in for now)
  filter(!(str_detect(spatial_source, 'iucn_parent') & !is.na(iucn_subpop)) |    ### parent tag with no subpop location = TRUE
         is.na(spatial_source))                                                  ### spatial_source is NA = TRUE (keep NAs in for now)
  ### This gets rid of any subpops that don't have identified shapefiles... 
  ### NOTE: this still doesn't uniquely identify which ID number goes with
  ### which subpopulation though!

spp_subpops_parents <- spp_all %>% 
  filter(str_detect(spatial_source, 'subpop') | str_detect(spatial_source, 'parent'))
table(spp_subpops_parents$spp_group)

spp_all <- spp_all %>%
  fix_am_subpops() %>%
  fix_iucn_subpops(spp_iucn_maps) 



write_csv(spp_all, file.path(dir_anx, scenario, 'int/spp_all.csv'))
```


``` {r Generate lookup IUCN species to LOICZID, eval = FALSE}

# x <- c("DAMSELFISH", "WRASSE", "SURGEONFISH_TANGS_UNICORNFISH", 
#        "COMBTOOTHBLENNIES", "TUNAS_BILLFISHES", "GROUPERS", "PUFFERFISH", "BUTTERFLYFISH", 
#        "LOBSTERS", "CONESNAILS", "CORALS_Part3", 
#        "CORALS_Part1", "SEACUCUMBERS", "MARINE_MAMMALS")
#   
# y <- c("SEABREAMS_PORGIES", "ANGELFISH", "BONEFISH_TARPONS", "CORALS_Part2", 
#        "REPTILES", "SEAGRASSES", "MANGROVES", "TERRESTRIAL_MAMMALS")
# z <- c("TUNAS_BILLFISHES", "MARINE_MAMMALS", "REPTILES", "SEABREAMS_PORGIES")

maps_list <- spp_all %>%
  filter(str_detect(spatial_source, 'iucn') & !str_detect(spatial_source, 'bli')) %>%
  select(sciname | iucn_sid | id_no (opt) | spp_group) %>%
  unique()

extract_loiczid_per_spp(maps_list, fn_tag = NULL, reload = TRUE)
### Extract loiczid cell IDs for each species within each species group.  Save 
### a .csv file for that group, with fields:
###       sciname | iucn_sid | presence | subpop | LOICZID | prop_area
### * presence codes: 1 extant; 2 prob extant (discontinued); 3 Possibly Extant;
###                   4 Possibly Extinct; 5 Extinct (post 1500); 6 Presence Uncertain
### NOTES: this takes a long time - multiple hours for some of the shape files.  
### * reload = FALSE allows it to skip extraction on groups with files already present.

```

``` {r SPP Generate species per cell tables for Aquamaps and IUCN }

am_cells_spp_sum <- process_am_summary_per_cell(reload = TRUE)
### NOTE: keyed data.table works way faster than the old inner_join or merge.
### loiczid | mean_cat_score | mean_trend_score | n_cat_species | n_trend_species
### AM does not include subspecies or subpops: every am_sid corresponds to exactly one sciname.

iucn_cells_spp_sum <- process_iucn_summary_per_cell(reload = TRUE)
### loiczid | mean_cat_score | mean_trend_score | n_cat_species | n_trend_species
### IUCN includes subpops - one sciname corresponds to multiple iucn_sid values.

```

``` {r  SPP Global Summarize mean category and trend per cell and per region}

summary_by_loiczid <- process_means_per_cell(am_cells_spp_sum, iucn_cells_spp_sum)
### This returns dataframe with variables:
### loiczid | weighted_mean_cat | weighted_mean_trend

rgn_cell_lookup <- extract_cell_id_per_region(reload = FALSE)
### | sp_id | loiczid | proportionArea | csq | cell_area
### saves lookup table to git-annex/globalprep/SPP_ICO/rgns/cellID_region_gcs_global.csv
### To use a different region shapefile, add an argument of: rgn_layer = '<rgn file here, without extension>'
### To use a shape file from a different directory (other than git-annex/globalprep/spatial/v2015/data), add an ogr_location argument.
### To run a different type of analysis, add an ohi_type argument with global (default), HS, or AQ

summary_by_rgn     <- process_means_per_rgn(summary_by_loiczid, rgn_cell_lookup)
### This returns dataframe with variables:
### sp_id | rgn_mean_cat | rgn_mean_trend | status

### Create final outputs:
if(!exists('summary_by_rgn')) 
  summary_by_rgn <- read.csv(file.path(dir_git, scenario, 'summary/rgn_summary.csv'))
spp_status <- summary_by_rgn %>%
  select(rgn_id, score = status)
spp_trend <- summary_by_rgn %>%
  select(rgn_id, score = rgn_mean_trend)
write_csv(spp_status, file.path(dir_git, scenario, 'data/spp_status_global.csv'))
write_csv(spp_trend,  file.path(dir_git, scenario, 'data/spp_trend_global.csv'))
```

``` {r SPP 3nm}

### Summarize mean category and trend per region within 3 nm of shore -----
### create final outputs for 3nm zone:
### This version is for the 3 nm coastal zone cells...
rgn_cell_lookup_3nm <- extract_cell_id_per_region(reload = FALSE, 
                                                  ogr_location = file.path(dir_neptune_data, 'git-annex/Global/NCEAS-Regions_v2014/data'),
                                                  rgn_layer = 'rgn_offshore3nm_gcs')
### | sp_id | loiczid | proportionArea | csq | cell_area
### saves lookup table to git-annex/globalprep/SPP_ICO/rgns/cellID_rgn_offshore3nm_gcs_global.csv

summary_by_rgn_3nm <- process_means_per_rgn(summary_by_loiczid, rgn_cell_lookup_3nm, rgn_note = '3nm')

if(!exists('summary_by_rgn_3nm')) 
  summary_by_rgn_3nm <- read.csv(file.path(dir_git, scenario, 'summary/rgn_summary_3nm.csv'))
spp_status_3nm <- summary_by_rgn_3nm %>%
  select(rgn_id, score = status)
spp_trend_3nm <- summary_by_rgn_3nm %>%
  select(rgn_id, score = rgn_mean_trend)
write_csv(spp_status_3nm, file.path(dir_git, scenario, 'data/spp_status_3nm.csv'))
write_csv(spp_trend_3nm,  file.path(dir_git, scenario, 'data/spp_trend_3nm.csv'))

```

``` {R SPP HS and Antarctic}

###Summarize mean category and trend per region -----
##############################################################################=
### create final outputs for HS zone:
### This version is for the high seas cells...
rgn_cell_lookup_hs <- extract_cell_id_per_region(reload = FALSE, rgn_layer = 'regions_gcs', ohi_type = 'HS')
### | sp_id | loiczid | proportionArea | csq | cell_area
### saves lookup table to git-annex/globalprep/SPP_ICO/rgns/cellID_region_gcs_HS.csv

summary_by_rgn_hs <- process_means_per_rgn(summary_by_loiczid, rgn_cell_lookup_hs, rgn_note = 'HS')

if(!exists('summary_by_rgn_hs')) 
  summary_by_rgn_hs <- read.csv(file.path(dir_git, scenario, 'summary/rgn_summary_hs.csv'))
spp_status_hs <- summary_by_rgn_hs %>%
  select(rgn_id, score = status)
spp_trend_hs <- summary_by_rgn_hs %>%
  select(rgn_id, score = rgn_mean_trend)
write_csv(spp_status_hs, file.path(dir_git, scenario, 'data/spp_status_hs.csv'))
write_csv(spp_trend_hs,  file.path(dir_git, scenario, 'data/spp_trend_hs.csv'))


### create final outputs for AQ zone:
### This version is for the Antarctic cells...
rgn_cell_lookup_aq <- extract_cell_id_per_region(reload = FALSE, rgn_layer = 'regions_gcs', ohi_type = 'AQ')
### | sp_id | loiczid | proportionArea | csq | cell_area
### saves lookup table to git-annex/globalprep/SPP_ICO/rgns/cellID_region_gcs_AQ.csv

summary_by_rgn_aq <- process_means_per_rgn(summary_by_loiczid, rgn_cell_lookup_aq, rgn_note = 'AQ')

if(!exists('summary_by_rgn_aq')) 
  summary_by_rgn_aq <- read.csv(file.path(dir_git, scenario, 'summary/rgn_summary_aq.csv'))
spp_status_aq <- summary_by_rgn_aq %>%
  select(rgn_id, score = status)
spp_trend_aq <- summary_by_rgn_aq %>%
  select(rgn_id, score = rgn_mean_trend)
write_csv(spp_status_aq, file.path(dir_git, scenario, 'data/spp_status_aq.csv'))
write_csv(spp_trend_aq,  file.path(dir_git, scenario, 'data/spp_trend_aq.csv'))
```

``` {r SPP swapping priority on AM and IUCN spatial source, eval = FALSE}

### for AquaMaps/IUCN paper, using OHI case study
scenario <- 'vAM_IUCN'

pref_flag_vec <- c('_IUCNpref', '_AMpref')
prob_filt_vec <-   c(0.40, 0)

for (pref_flag in pref_flag_vec) {   ### pref_flag  <- pref_flag_vec[1]
  for (prob_filt in prob_filt_vec) { ### prob_filt <- prob_filt_vec[1]
    message('Processing parameters: ', pref_flag, ', prob = ', prob_filt)
    spp_all <- create_spp_master_lookup(source_pref = ifelse(pref_flag == '_AMpref', 'am', 'iucn'), 
                                        fn_tag = pref_flag, 
                                        reload = TRUE)
    # only affected by am/iucn preference
    spp_all <- spp_all %>%
      select(-iucn_subpop) %>% ### need to figure out how to use this info
      select(-sciname) %>%     ### get rid of this to eliminate duped names
      filter(!is.na(spatial_source)) %>% 
      unique()
        
    am_cells_spp_sum <- process_am_summary_per_cell(fn_tag = sprintf('%s_prob%s', pref_flag, prob_filt), prob_filter = prob_filt, reload = TRUE)
    # affected by prob filter and preference flag
    
    iucn_cells_spp_sum <- process_iucn_summary_per_cell(fn_tag = sprintf('_pref%s', pref_flag), reload = TRUE)
    # affected by preference flag
    
    sum_by_loiczid <- process_means_per_cell(am_cells_spp_sum, iucn_cells_spp_sum, fn_tag = sprintf('%s_%s', pref_flag, prob_filt))
    rgn_cell_lookup <- extract_cell_id_per_region(reload = FALSE)
    sum_by_rgn     <- process_means_per_rgn(sum_by_loiczid, rgn_cell_lookup, rgn_note = sprintf('%s_%s', pref_flag, prob_filt))
    ### This returns dataframe with variables:
    ### sp_id | rgn_mean_cat | rgn_mean_trend | status
    
    ### Create final outputs:
    spp_status <- sum_by_rgn %>%
      select(rgn_id, score = status)
    spp_trend <- sum_by_rgn %>%
      select(rgn_id, score = rgn_mean_trend)
    write_csv(spp_status, 
              file.path(dir_git, scenario, 
                        sprintf('data/spp_status_global%s_prob%s.csv', pref_flag, prob_filt)))
    write_csv(spp_trend,  
              file.path(dir_git, scenario, 
                        sprintf('data/spp_trend_global%s_prob%s.csv', pref_flag, prob_filt)))
  }
}
  

```

# TO DO: deal with populations for IUCN species.
#   examine range maps - are there overlaps between parent and subspecies/subpops?
#   - if not - no need to worry.
#   - if so  - worry!  once extracted, overlap info is lost.
#     - do a difference between parent and each subpop, and then use that for extract?
#     - if overlap is consistent (i.e. parent fully covers each subpop), just subtract prop_area?
#       - what if two subpopulations in same cell - might not overlap; so subtract won't work.
#   ??? are parents already taken out of the iucn_spp_marine_global? that's what it seems like.
#       In ingest_iucn script, looks like subpops are subsetted OUT of the main list.
#       But then merged back in, and then parents subsetted out.
#   TO DO: confirm that this is the right way to do things?  Does 'parent' really just consist
#       of sum of 'subpopulations' or could it encompass other areas?
#     * find the full list of parents and subpopulations for marine species, and see how
#       many are IUCN range map species.  Select by sci name, spot check a few in QGIS or ArcGIS.
# subpops <- read_csv(file.path(dir_anx, scenario, 'intermediate/spp_iucn_marine_subpopulations.csv')) %>% 
#   rename(parent = sid)
# spp_all_subpops <- spp_all %>%
#   filter(spatial_source == 'iucn') %>%
#   left_join(subpops,
#             by = c('iucn_sid' = 'subpopulations')) %>%
#   filter(!is.na(parent)) %>%
#   select(iucn_sid, parent, sciname, iucn_category, spp_group, spatial_source)

# NOTES ON model.R - should be covered in the above code.
# ??? compare BB and MF versions - BB version involves 3nm regions? MF version is for AQ and HS
# Averages IUCN score per cell for all species in that cell; then takes area-weighted average of those scores for each region
# .csv of score and trend output.
# ??? calculate this for both SPP and ICO - SPP is area-weighted (so rare species don't count as much), but ICO is not (rare species count just as much as common)

# some checks: 

# examine species names for mismatches/typos/name variations.  Similar defined as:
# * first five letters of one field (genus or species) match for both AM and IUCN.
# * exact match for other field.
sim_names <- check_sim_names(spp_all)

# check to see how closely IUCN category info matches between IUCN data and AM data
category_check <- spp_all %>% 
  filter(!is.na(iucn_category) & !is.na(am_category)) %>%
  select(sciname, am_category, iucn_category) %>%
  mutate(am_category   = as.character(am_category),
         iucn_category = as.character(iucn_category),
         cat_match = (am_category == iucn_category))
sum(category_check$cat_match)/nrow(category_check)
# Out of 2962 species that include IUCN category from both AM and IUCN,
# 2862 of them are matches.  96.6% match.

