---
title: 'OHI: Species subgoal'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohiprep/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = FALSE, message = FALSE, warning = FALSE)

library(foreign)
library(data.table)
library(sp)
library(rgdal)
# library(raster)
# library(maptools)
library(readr)

source('~/github/ohiprep/src/R/common.R')

goal     <- 'globalprep/spp_ico'
scenario <- 'v2016'
dir_anx  <- file.path(dir_M, 'git-annex', goal) 
dir_data_am    <- file.path(dir_M, 'git-annex/globalprep/_raw_data', 'aquamaps', 'd2015') 
dir_data_iucn  <- file.path(dir_M, 'git-annex/globalprep/_raw_data', 'iucn_spp', 'd2015') 
dir_data_bird  <- file.path(dir_M, 'git-annex/globalprep/_raw_data', 'birdlife_intl', 'd2015') 
dir_git  <- file.path('~/github/ohiprep', goal)

source(file.path(dir_git, scenario, 'spp_fxn.R'))
if(!file.exists(file.path(dir_git, 'README.md'))) {
  warning(sprintf('No README detected in %s', dir_git))
}
if(!file.exists(file.path(dir_git, scenario, 'README.md'))) {
  warning(sprintf('No README detected in %s', file.path(dir_git, scenario)))
}
# SPP-specific and ICO-specific functions
```

<!--[REFERENCE RMD FILE: https://rawgit.com/OHI-Science/ohiprep/master/globalprep/Pressures_OceanAcidification/v2015/oa_create_layer_2015.html]-->

# Summary

Spatial data from IUCN and Aquamaps is combined with extinction risk information from IUCN to generate regional scores for the Species subgoal.  A region's status is based upon an area-weighted average of species health across each global reporting region.

From Halpern et al (2012):

> The target for the Species sub-goal is to have all species at a risk status of Least Concern. We scaled the lower end of the biodiversity goal to be 0 when 75% species are extinct, a level comparable to the five documented mass extinctions and would constitute a catastrophic loss of biodiversity. The Status of assessed species was calculated as the area- and threat status-weighted average of the number of threatened species within each 0.5 degree grid cell.

**Mean risk status per cell:**

$$\bar{R}_{cell} = \frac{\displaystyle\sum_{species}(Risk)}{n_{spp}}$$

**Mean risk status per region:**

$$\bar{R}_{SPP} = \frac{\displaystyle\sum_{cells}(\bar{R}_{cell} * A_{cell} * pA_{cell-rgn})}{A_{rgn}}$$

**Species goal model**

$$X_{SPP} = \frac{((1 - \bar{R}_{SPP}) - 0.25)}{(1 - 0.25)} * 100%$$

where:

* $X_{SPP}$ is Species goal status
* $\bar{R}_{cell}$ is mean extinction risk for one cell
* $\bar{R}_{SPP}$ is area-weighted mean extinction risk for a region
* $A_{cell}$ is cell area
* $pA_{cell-rgn}$ is percent of cell area included in region
* *Risk* is scaled value for species extinction risk category, based on: 
    * 'LC' = 0.0, 'NT' = 0.2, 'VU' = 0.4, 'EN' = 0.6, 'CR' = 0.8, 'EX' = 1.0
* SPP trend is calculated in a similar area-weighted mean, but population trend values are assigned according to:
    * 'Decreasing' = -0.5, 'Stable' = 0.0, 'Increasing' = +0.5

# Updates from previous assessment

Changes since 2015 SPP subgoal for global OHI:

* The 2016 assessment now includes bird species, based upon BirdLife International data.  "Marine" bird species include any bird species with overlap of marine cells. (?) Prior assessments did not include this data.
* Previous assessments counted overlapping species as any species with a binomial included in both data sets (AquaMaps and IUCN/BirdLife).  Synonyms were matched and typos were fixed manually where possible.  This assessment takes advantage of the `taxize` package to identify and match synonyms for better match rate.
* Previous assessments used an AquaMaps threshold of 40% to determine whether a species counted as "present" within a cell.  Based on the IUCN/AquaMaps manuscript, we are now using a 0% threshold for AquaMaps as it more closely resembles the "limits of distribution" criterion used for IUCN inclusion.
* Previous assessments ignored subpopulations for SPP; with IUCN's newest data, a subpopulation field identifies distinct subpopulations (and IUCN risk codes) for a number of species (mostly mammals, but a handful of others as well).  This allows us to incorporate subpopulation scores in the global assessment.
    * From http://www.iucnredlist.org/details/3897/0 (Loggerhead sea turtle):
        * Rationale: The global population of the Loggerhead Turtle (Caretta caretta) comprises 10 subpopulations (see Figure 2 in the Supplementary Material) that vary widely in population size, geographic range, and population trends, and are the appropriate units for assessment of global conservation status for this species (Wallace et al. 2010, 2011). As such, assessments have been completed for each of the 10 subpopulations, in addition to the combined global population assessment required by the IUCN (see Table 1 in the Supplementary Material). At the global level, both geographic distribution and population size are much larger than required to qualify for a threatened category. The available long-term series of nest counts (used as an index of population abundance) show an important decrease in the past (47%). Therefore, the Loggerhead Turtle is considered as Vulnerable under current IUCN Red List Criteria (criterion A2b). The previous listing, published in 1996, was Endangered under criterion A1bd (Marine Turtle Specialist Group 1996).
        * __Results indicate that the Loggerhead Turtle, as a single taxonomic entity, will not go extinct globally in the next generation according to any Red List criteria. However, the global listing is not an appropriate representation of the conservation status__ of the biologically relevant subpopulations that make up the global Loggerhead Turtle population. Subpopulation assessments demonstrated wide variation not only in status of individual subpopulations (as indicated by IUCN Red List Categories), but also in the criteria under which the individual subpopulations qualified for a threatened category (see Table 1 in the Supplementary Material). For these reasons, the __subpopulation-level assessments for the Loggerhead Turtle should be given priority in evaluating the true global conservation status__ of this species. __This conclusion follows the precedent for other long-lived, widely distributed species,__ including the Leatherback Turtle (Wallace et al. 2013).

***

# Data Sources

IUCN:

* __Reference__: 
    * IUCN 2015. The IUCN Red List of Threatened Species. Version 2015-4. <http://www.iucnredlist.org>.
        * Shapefiles available from: http://www.iucnredlist.org/technical-documents/spatial-data
        * __Downloaded__: 21 December 2015.
    * BirdLife International and NatureServe (2015) Bird species distribution maps of the world. BirdLife International, Cambridge, UK and NatureServe, Arlington, USA.
        * Zipped shapefile available from: http://www.biodiversityinfo.org/spcdownload/r5h8a1/ - username and password via BirdLife International.  
        * __Downloaded__: 14 December 2015.
* __Description__:  Shapefiles containing polygons of assessed species ranges; each shapefile represents all assessed species within a comprehensively-assessed (i.e. >90% assessed) taxonomic group.
* __Native data resolution__: NA
* __Time range__: NA
* __Format__:  Shapefile

AquaMaps:

* __Reference__: 
    * Kaschner, K., J. Rius-Barile, K. Kesner-Reyes, C.Garilao, S.O. Kullander, T. Rees and R. Froese (2013). AquaMaps: Predicted range maps for aquatic species. World wide web electronic publication, www.aquamaps.org, Version 08/2015.
        * __Downloaded__: August 2015.
* __Description__:  .sql files containing information to recreate rasters of cell attributes and cell-by-cell probability of occurrence for 22889 marine species.
* __Native data resolution__: 0.5° latitude and longitude  
* __Time range__: NA
* __Format__:  SQL database files converted into .csv

***
  
# Methods

## Extract AquaMaps data from .sql files

AquaMaps data for the 2016 assessment was provided as .sql files, as in previous years, that can be used to generate an SQL database.  Each line in the .sql is a command to populate the SQL database.

* Original files extracted from `aquamaps_2015_full_dataset_ohi.zip`:
    * `hcaf_ohi.sql`
    * `speciesoccursum_ohi.sql`
    * `hcaf_species_native_ohi.sql`
* Processed files saved to git-annex:
    * `hcaf_truncated.csv`
    * `speciesoccursum.csv`
    * `hcaf_sp_native_trunc.csv`

To extract data, we instead scan each line for `CREATE TABLE` and `INSERT INTO` commands to create and save dataframes.  Note that the `am_extract_2015.R` script discards much of the data from these .sqls that is not used within the OHI Species Goal processing (thus "truncated").  This speeds up read time and processing time and avoids parsing issues with some of the rows/columns.
Note also that the hcaf_species_native_ohi.sql file does not originally contain LOICZID information; this is added in the extract script, since LOICZID as a cell identifier is faster and less memory intensive than CsquareCode (integer vs character string).

``` {r Ingest Aquamaps data to csv from sql, eval = FALSE}

source(file.path(dir_git, 'R/am_extract_2015.R'))
ingest_aquamaps()

```

``` {r pull headers from AquaMaps files}
hcaf_tmp      <- read_csv(file.path(dir_data_am, 'csv/hcaf_truncated.csv'),       n_max = 5)
sppoccsum_tmp <- read_csv(file.path(dir_data_am, 'csv/speciesoccursum.csv'),      n_max = 5)
hcaf_sp_n_tmp <- read_csv(file.path(dir_data_am, 'csv/hcaf_sp_native_trunc.csv'), n_max = 5)
```

`r knitr::kable(hcaf_tmp,      caption = 'Half-degree cell info from hcaf_truncated.csv (first few rows)')`

`r knitr::kable(sppoccsum_tmp, caption = 'Species info from speciesoccursum.csv (first few rows)')`

`r knitr::kable(hcaf_sp_n_tmp, caption = 'Species-to-cell lookup from hcaf_sp_native_trunc.csv (first few rows)')`

## Ingest IUCN species list

To identify appropriate IUCN species for the analysis, we identified all IUCN Red List species whose habitat included "marine" designation.  The `ingest_iucn.R` script scrapes this data directly from the IUCN Red List website.

* Starting point: Access Red List API at http://api.iucnredlist.org/index/all.csv.  This provides a list of all Red List species, including unique IUCN code, taxonomic nomenclature, and Red List extinction risk.
* For all identified species, we access species-specific info at http://api.iucnredlist.org/details/X/0 where X is the unique IUCN species ID; this page is saved to a cache directory on git-annex as a .htm file.  At the same time, we scrape the "habitats" field from the saved page using XML tags.
* All species with "marine" habitat are then scraped to find details on population trend and subpopulation ID numbers.
* After some cleaning of scientific names (accents, spaces, html tags, etc) the file is saved to git-annex for later use.

Processed files, saved to `git-annex/globalprep/spp_ico/v201x/int`: 
* `spp_iucn_all.csv` - full list of IUCN species pulled from web, some cleaning.
* `spp_iucn_habitats.csv` - list of IUCN species (by iucn_sid) and corresponding habitat.
* `spp_iucn_marine.csv` - prepped list: cleaned marine list with subpops and trends.

``` {r Ingest IUCN species list, eval = FALSE}

source(file.path(goal, 'R/ingest_iucn.R'))

```

``` {r pull headers from IUCN marine species list}
mar_tmp <- read_csv(file.path(dir_anx, scenario, 'int/spp_iucn_marine.csv'), n_max = 5)
```

`r knitr::kable(mar_tmp, caption = 'IUCN marine species list and info from spp_iucn_marine.csv (first few rows)')`

## Generate full species lookup table

Having processed AquaMaps and IUCN species raw data, we can now prepare a full combined list of all species to be included in the OHI SPP goal.  The function `create_spp_master_lookup()` creates the full lookup table:

* Load AquaMaps species info from `speciesoccursum.csv`
    * Verify `sciname` field; the `verify_scinames()` function uses `taxize::gnr_resolve()` to compare AM scinames to accepted names from Encyclopedia of Life and NCBI databases - this helps resolve differences in naming conventions and species aliases.
* Load IUCN species info from `spp_iucn_marine.csv`
    * Verify `sciname` field as above.
* Join the two species lists by verified scientific names as the best common identifier.
* Create an extinction risk category field for each row, using IUCN category, or failing that the category listed in AquaMaps; failing that, assign NA.

At this point, the list is all AquaMaps species and all marine-identified IUCN species.  Next, identify the source of spatial distribution data for each species, if available.

* From the downloaded IUCN shapefiles, determine which species have spatial data.
* Depending on preference and availability of spatial source (AquaMaps or IUCN), tag each row with a specific `spatial_source` ("am" or "iucn").
* Follow a similar process for IUCN species included in the BirdLife International geodatabase (tag as "iucn-bli").

Now having identified spatial source availability and preference for all species on the list:

* Clean out duplicated species
* Assign numeric values to population Red List category and population trend.
* Save result to `git-annex/globalprep/spp_ico/v2016/int/spp_all_raw.csv`
* Filter out subpop species rows (species with a non-NA `parent_sid`) that do not have a named subpopulation location (i.e. `iucn_subpop` field is NA, instead of the name of a subpopulation)
* Save final list to `git-annex/globalprep/spp_ico/v2016/int/spp_all_cleaned.csv`

``` {r Generate spp lookup}

reload <- FALSE

if(reload) {
  ### NOTE: Manually set up the subpopulations for 2016: see subpops_list.csv.
  ### For each of these species, the shapefile was modified so that the id_no
  ### field related to the subpop iucn_sid instead of the parent iucn_sid.

  spp_all <- create_spp_master_lookup(source_pref = 'iucn', fn_tag = '', reload = TRUE)
  ### | am_sid | sciname | am_cat | iucn_sid | iucn_cat | pop_trend | pop_cat | 
  ### | info_source | spp_group | id_no | objectid | spatial_source | cat_score | trend_score |
  
  ### check subpops: all spp with an IUCN subpop code
  spp_all_subpops <- spp_all %>%
    filter(!is.na(iucn_subpop)) %>%
    unique()
  
  ### if there is any cleaning to be done, this is where it would go.
  
  write_csv(spp_all, file.path(dir_anx, scenario, 'int/spp_all_cleaned.csv'))
} else {
  spp_all <- read_csv(file.path(dir_anx, scenario, 'int/spp_all_cleaned.csv'))
}

```

`r knitr::kable(head(spp_all), caption = 'Combined species list from spp_all_cleaned.csv (first few rows)')`

## Spatialize species information using AquaMaps and IUCN spatial data

### Extract IUCN polygons to half-degree cells

We extract IUCN polygon presence to the same half-degree cells as AquaMaps to simplify the analysis.
* The `spp_all` species list includes a field `spp_group` that identifies which shapefile contains the spatial information for a given species.
    * for each species group, specific species are identified by comparing `iucn_sid` from dataframe to `id_no` within the shapefile.
* Extract loiczid cell IDs for each species within each species group.  Save a .csv file for that group, with fields:
    * sciname | iucn_sid | presence | subpop | LOICZID | prop_area
    * presence codes: 1 extant; 2 prob extant (discontinued); 3 Possibly Extant; 4 Possibly Extinct; 5 Extinct (post 1500); 6 Presence Uncertain
* __NOTE:__ this takes a long time - multiple hours for some of the shape files.  
    * by passing a filtered data frame to the function, you can focus the process only on new or updated shapefiles
    * reload = FALSE allows the function to skip extraction on groups with files already present.  Set to TRUE if you need to extract an updated shapefile (or change the shapefile name, or delete the previous extraction...)

``` {r Generate lookup IUCN species to LOICZID, eval = FALSE}

### set up maps_list for all standard (non-bird) IUCN species...
maps_list_iucn <- spp_all %>%
  filter(str_detect(spatial_source, 'iucn') & !str_detect(spatial_source, 'bli')) %>%
  dplyr::select(sciname, iucn_sid, spp_group) %>%
  unique()

extract_loiczid_per_spp(maps_list_iucn,
                        shp_dir = file.path(dir_data_iucn, 'iucn_shp'), 
                        fn_tag = scenario, 
                        reload = FALSE)

### set up maps_list for all bird IUCN species...
maps_list_bli <- spp_all %>%
  filter(str_detect(spatial_source, 'bli')) %>%
  dplyr::select(sciname, iucn_sid, spp_group) %>%
  mutate(spp_group = 'BOTW') %>%
  unique()

extract_loiczid_per_spp(maps_list_bli,
                        shp_dir = dir_data_bird, 
                        fn_tag = scenario, 
                        reload = FALSE)

```

### Generate cell-by-cell summary of species

For each half-degree cell, tally up the number of species present and determine a mean species risk value and population trend value for the cell.  

* At this point, the species-cell lists are filtered to species with valid extinction risk categories - i.e. not DD and not NA.
* Since this averaging is done for each data set separately, we also track the species count per cell used to determine both the risk and the trend (separately, since many species with a risk value have no trend information, i.e. NA).  These counts are used to weight the values when the two are combined.
* Data-set specific idiosyncracies:
    * For AquaMaps, we apply a threshold to set the minimum probability of occurrence that determines species "presence."
    * For IUCN, no threshold is needed; but the shapefiles include a "presence" attribute in which a value of 5 indicates a region in which a subpopulation has become extinct.  We use this to manually reset local extinction risk and trend to EX and NA respectively.
    * Note that for IUCN, we determine the proportional area when extracting polygons; currently we just consider any presence to fill the cell (similar to assuming even a low AquaMaps probability to indicate presence within the entire cell).

``` {r SPP Generate species per cell tables for Aquamaps and IUCN }

am_cells_spp_sum <- process_am_summary_per_cell(spp_all, fn_tag = '', prob_filter = 0, reload = FALSE)
### NOTE: keyed data.table works way faster than the old inner_join or merge.
### loiczid | mean_cat_score | mean_trend_score | n_cat_species | n_trend_species
### AM does not include subspecies or subpops: every am_sid corresponds to exactly one sciname.

iucn_cells_spp_sum <- process_iucn_summary_per_cell(spp_all, fn_tag = '', reload = FALSE)
### loiczid | mean_cat_score | mean_trend_score | n_cat_species | n_trend_species
### IUCN includes subpops - one sciname corresponds to multiple iucn_sid values.

```

`r knitr::kable(head(am_cells_spp_sum), caption = 'AquaMaps cell summary (first few rows)')`

`r knitr::kable(head(iucn_cells_spp_sum), caption = 'IUCN cell summary (first few rows)')`

## Summarize status and trend by region

Finally we take the two cell-by-cell summaries and combine, using a species-count weighting to determine the mean category and trend per cell.  Cells are aggregated to regions, to calculate an area-weighted regional mean category, trend, and status.

These are then saved to status and trend layer outputs for global (shown in table) as well as 3 nautical mile, Antarctic, and High Seas regions.

``` {r  SPP Global Summarize mean category and trend per cell and per region}

sum_by_loiczid <- process_means_per_cell(am_cells_spp_sum, iucn_cells_spp_sum, fn_tag = '')
### This returns dataframe with variables:
### loiczid | weighted_mean_cat | weighted_mean_trend | n_cat_spp | n_tr_spp
  
rgn_cell_lookup <- extract_cell_id_per_region(reload = FALSE)
### | rgn_id | rgn_name | loiczid | proportionArea | csq | cell_area

### saves lookup table to git-annex/globalprep/spp_ico/rgns/cellID_region_gcs_global.csv
### To use a different region shapefile, add an argument of: rgn_layer = '<rgn file here, without extension>'
### To use a shape file from a different directory (other than git-annex/globalprep/spatial/v2015/data), add an ogr_location argument.
### To run a different type of analysis, add an ohi_type argument with global (default), HS, or AQ

sum_by_rgn <- process_means_per_rgn(sum_by_loiczid, rgn_cell_lookup, rgn_note = 'gl')
### This returns dataframe with variables:
### rgn_id | rgn_mean_cat | rgn_mean_trend | status

### Create final outputs:
if(!exists('sum_by_rgn')) 
  sum_by_rgn <- read.csv(file.path(dir_git, scenario, 'summary/rgn_summary_gl.csv'))
spp_status <- sum_by_rgn %>%
  dplyr::select(rgn_id, score = status)
spp_trend <- sum_by_rgn %>%
  dplyr::select(rgn_id, score = rgn_mean_trend)
write_csv(spp_status, file.path(dir_git, scenario, 'output/spp_status_global.csv'))
write_csv(spp_trend,  file.path(dir_git, scenario, 'output/spp_trend_global.csv'))
```

`r DT::datatable(sum_by_rgn, caption = 'Region summary: mean category, trend, and status')`


``` {r SPP 3nm}

### Summarize mean category and trend per region within 3 nm of shore -----
### create final outputs for 3nm zone:
### This version is for the 3 nm coastal zone cells...
rgn_cell_lookup_3nm <- extract_cell_id_per_region(reload = FALSE, 
                                                  ogr_location = file.path(dir_M, 'git-annex/Global/NCEAS-Regions_v2014/data'),
                                                  rgn_layer = 'rgn_offshore3nm_gcs')
### | sp_id | loiczid | proportionArea | csq | cell_area
### saves lookup table to git-annex/globalprep/spp_ico/rgns/cellID_rgn_offshore3nm_gcs_global.csv

sum_by_rgn_3nm <- process_means_per_rgn(sum_by_loiczid, rgn_cell_lookup_3nm, rgn_note = '3nm')

if(!exists('sum_by_rgn_3nm')) 
  sum_by_rgn_3nm <- read.csv(file.path(dir_git, scenario, 'summary/rgn_summary_3nm.csv'))
spp_status_3nm <- sum_by_rgn_3nm %>%
  dplyr::select(rgn_id, score = status)
spp_trend_3nm <- sum_by_rgn_3nm %>%
  dplyr::select(rgn_id, score = rgn_mean_trend)
write_csv(spp_status_3nm, file.path(dir_git, scenario, 'output/spp_status_3nm.csv'))
write_csv(spp_trend_3nm,  file.path(dir_git, scenario, 'output/spp_trend_3nm.csv'))

```

``` {R SPP HS and Antarctic}

###Summarize mean category and trend per region -----
##############################################################################=
### create final outputs for HS zone:
### This version is for the high seas cells...
rgn_cell_lookup_hs <- extract_cell_id_per_region(reload = FALSE, rgn_layer = 'regions_gcs', ohi_type = 'hs')
### | sp_id | loiczid | proportionArea | csq | cell_area
### saves lookup table to git-annex/globalprep/spp_ico/rgns/cellID_region_gcs_HS.csv

sum_by_rgn_hs <- process_means_per_rgn(sum_by_loiczid, rgn_cell_lookup_hs, rgn_note = 'hs')

if(!exists('sum_by_rgn_hs')) 
  sum_by_rgn_hs <- read.csv(file.path(dir_git, scenario, 'summary/rgn_summary_hs.csv'))
spp_status_hs <- sum_by_rgn_hs %>%
  dplyr::select(rgn_id, score = status)
spp_trend_hs <- sum_by_rgn_hs %>%
  dplyr::select(rgn_id, score = rgn_mean_trend)
write_csv(spp_status_hs, file.path(dir_git, scenario, 'output/spp_status_hs.csv'))
write_csv(spp_trend_hs,  file.path(dir_git, scenario, 'output/spp_trend_hs.csv'))


### create final outputs for AQ zone:
### This version is for the Antarctic cells...
rgn_cell_lookup_aq <- extract_cell_id_per_region(reload = FALSE, rgn_layer = 'regions_gcs', ohi_type = 'aq')
### | sp_id | loiczid | proportionArea | csq | cell_area
### saves lookup table to git-annex/globalprep/spp_ico/rgns/cellID_region_gcs_AQ.csv

sum_by_rgn_aq <- process_means_per_rgn(sum_by_loiczid, rgn_cell_lookup_aq, rgn_note = 'aq')

if(!exists('sum_by_rgn_aq')) 
  sum_by_rgn_aq <- read.csv(file.path(dir_git, scenario, 'summary/rgn_summary_aq.csv'))
spp_status_aq <- sum_by_rgn_aq %>%
  dplyr::select(rgn_id, score = status)
spp_trend_aq <- sum_by_rgn_aq %>%
  dplyr::select(rgn_id, score = rgn_mean_trend)
write_csv(spp_status_aq, file.path(dir_git, scenario, 'output/spp_status_aq.csv'))
write_csv(spp_trend_aq,  file.path(dir_git, scenario, 'output/spp_trend_aq.csv'))
```

``` {r SPP swapping priority on AM and IUCN spatial source, eval = FALSE}

### for AquaMaps/IUCN paper, using OHI case study
scenario <- 'vAM_IUCN'

pref_flag_vec <- c('_IUCNpref', '_AMpref')
prob_filt_vec <-   c(0.40, 0)

for (pref_flag in pref_flag_vec) {   ### pref_flag  <- pref_flag_vec[1]
  for (prob_filt in prob_filt_vec) { ### prob_filt <- prob_filt_vec[1]
    message('Processing parameters: ', pref_flag, ', prob = ', prob_filt)
    spp_all <- create_spp_master_lookup(source_pref = ifelse(pref_flag == '_AMpref', 'am', 'iucn'), 
                                        fn_tag = pref_flag, 
                                        reload = TRUE)
    # only affected by am/iucn preference
    spp_all <- spp_all %>%
      dplyr::select(-iucn_subpop) %>% ### need to figure out how to use this info
      dplyr::select(-sciname) %>%     ### get rid of this to eliminate duped names
      filter(!is.na(spatial_source)) %>% 
      unique()
        
    am_cells_spp_sum <- process_am_summary_per_cell(fn_tag = sprintf('%s_prob%s', pref_flag, prob_filt), 
                                                    prob_filter = prob_filt, reload = TRUE)
    # affected by prob filter and preference flag
    
    iucn_cells_spp_sum <- process_iucn_summary_per_cell(fn_tag = sprintf('_pref%s', pref_flag), reload = TRUE)
    # affected by preference flag
    
    sum_by_loiczid <- process_means_per_cell(am_cells_spp_sum, iucn_cells_spp_sum, 
                                             fn_tag = sprintf('%s_%s', pref_flag, prob_filt))
    
    rgn_cell_lookup <- extract_cell_id_per_region(reload = FALSE)
    
    sum_by_rgn     <- process_means_per_rgn(sum_by_loiczid, rgn_cell_lookup, 
                                            rgn_note = sprintf('%s_%s', pref_flag, prob_filt))
    ### This returns dataframe with variables:
    ### sp_id | rgn_mean_cat | rgn_mean_trend | status
    
    ### Create final outputs:
    spp_status <- sum_by_rgn %>%
      dplyr::select(rgn_id, score = status)
    spp_trend <- sum_by_rgn %>%
      dplyr::select(rgn_id, score = rgn_mean_trend)
    write_csv(spp_status, 
              file.path(dir_git, scenario, 
                        sprintf('output/spp_status_global%s_prob%s.csv', pref_flag, prob_filt)))
    write_csv(spp_trend,  
              file.path(dir_git, scenario, 
                        sprintf('output/spp_trend_global%s_prob%s.csv', pref_flag, prob_filt)))
  }
}

```

## Determine species by region

The `calc_rgn_spp()` function takes in lookup tables of species by cell (for both IUCN and AM), a cell-to-region lookup, and a species info lookup.  From this it generates a list of which species occur in which regions, including species information.

``` {r species_by_region}

iucn_cells_spp <- read_csv(file.path(dir_anx, scenario, 'int/iucn_cells_spp.csv'),
                           col_types = 'cddddc')
am_cells_spp   <- read_csv(file.path(dir_anx, scenario, 'int/am_cells_spp_prob0.csv'))

ics_keyed     <- data.table(iucn_cells_spp, key = "loiczid") 
acs_keyed     <- data.table(am_cells_spp,   key = "loiczid") 

rgn_keyed     <- data.table(rgn_cell_lookup,     key = "loiczid")
rgn_keyed_3nm <- data.table(rgn_cell_lookup_3nm, key = "loiczid")
rgn_keyed_aq  <- data.table(rgn_cell_lookup_aq,  key = "loiczid")
rgn_keyed_hs  <- data.table(rgn_cell_lookup_hs,  key = "loiczid")

rgn_spp_gl  <- calc_rgn_spp(ics_keyed, acs_keyed, rgn_keyed,     spp_all)
rgn_spp_3nm <- calc_rgn_spp(ics_keyed, acs_keyed, rgn_keyed_3nm, spp_all)
rgn_spp_aq  <- calc_rgn_spp(ics_keyed, acs_keyed, rgn_keyed_aq,  spp_all)
rgn_spp_hs  <- calc_rgn_spp(ics_keyed, acs_keyed, rgn_keyed_hs,  spp_all)

write_csv(rgn_spp_gl,  file.path(dir_git, scenario, 'output/rgn_spp_gl.csv'))
write_csv(rgn_spp_3nm, file.path(dir_git, scenario, 'output/rgn_spp_3nm.csv'))
write_csv(rgn_spp_aq,  file.path(dir_git, scenario, 'output/rgn_spp_aq.csv'))
write_csv(rgn_spp_hs,  file.path(dir_git, scenario, 'output/rgn_spp_hs.csv'))
```

`r knitr::kable(head(rgn_spp_gl), caption = 'Species by region - first few rows')`

## Comparing scenarios and data

The following plots compare the status scores generated for the 2015 assessment to those generated for 2016.  

* The first compares the released scores for 2015 to the proposed scores for 2016.  This changes both the datasets and the methodology between the two axes.
    * 2015 scores used data from 2014 (d2014) for both AquaMaps and IUCN, and scripts from the v2015 scenario.
    * 2016 scores use data from 2015 (d2015) for both AquaMaps and IUCN, and scripts from the v2016 scenario.
* The first compares regenerated scores for 2015 to the proposed scores for 2016. This keeps the methodology largely the same between the two axes (except as noted below), and compares the changes in data.
    * 2015 scores used data from 2014 (d2014) for both AquaMaps and IUCN.
    * 2016 scores use data from 2015 (d2015) for both AquaMaps and IUCN.
    * Both scores use scripts from the v2016 scenario.  Methodology exceptions: 
        * AquaMaps for d2014 is processed with a 40% threshold (as in v2015) while d2015 is processed with a 0% threshold.
        * IUCN species-cell lookup for d2014 was generated using v2015 scripts, which relied on the `sciname` field for joining.  I've added a step to add `iucn_sid` values according to `sciname`, to allow the species-cell lookup to work with the v2016 scripts.

``` {r plot_scores_vs_2015, eval = FALSE}

spp_all_d14v16 <- read_csv(file.path(dir_anx, 'v2015', 'int/spp_all.csv')) %>%
  rename(cat_score = category_score)

### AquaMaps: take original cells-to-species file (fields: SpeciesID, 
###   CsquareCode, probability), merge with csquarecode-to-loiczid using
###   data.table x[y] merging; then ditch csquarecode.  Resulting
###   fields: am_sid, loiczid, probability
am_cells <- read_csv(file.path(dir_data_am, '../d2014/tables/hcaf.csv')) %>%
  setnames(tolower(names(.))) %>%
  select(csquarecode, loiczid) %>%
  data.table(key = 'csquarecode')
acs_d14_raw <- read_csv(file.path(dir_data_am, '../d2014/tables/ohi_hcaf_species_native.csv'),
                    col_types = '_ccd__') %>%
  setnames(tolower(names(.))) %>%
  rename(am_sid = speciesid)
acs_d14 <- data.table(am_cells, key = 'csquarecode')[data.table(acs_d14_raw, key = 'csquarecode')] %>%
  select(-csquarecode) %>%
  as.data.frame()
  
rm('am_cells', 'acs_d14_raw') ### clean house a bit

### IUCN: v2015/int/iucn_cells_spp.csv is the bound species-to-cells, with 
###   id_no NA for many species (no id_no from polygons).
###   Bind this with iucn_sid-to-sciname matches from the species info list.
###   Resulting fields: iucn_sid, sciname, loiczid, presence, propArea
ics_d14_raw <- read_csv(file.path(dir_anx, 'v2015', 'int/iucn_cells_spp.csv'), col_types = 'cdddd') %>%
  rename(iucn_sid = id_no)
### the raw data contains many species with no id_no value from the polygons.  Fix that!: 
ics_d14 <- ics_d14_raw %>%
  # select(iucn_sid, sciname) %>%
  filter(!is.na(iucn_sid)) %>%      ### This first subset contains id_no (now iucn_sid) from polygons
  bind_rows(ics_d14_raw %>%         ### This next subset is missing id_no (iucn_sid); join to iucn_sid by sciname
              filter(is.na(iucn_sid)) %>%
              select(-iucn_sid) %>%
              left_join(spp_all_d14v16 %>%
                          select(iucn_sid, sciname),
                        by = 'sciname')) %>%
  unique()
rm('ics_d14_raw') ### cleaning house


acs_sum_d14v16 <- process_am_summary_per_cell(spp_all_d14v16,
                                              spp_cells = acs_d14,
                                              fn_tag = '_d14v16', 
                                              prob_filter = 0.4, reload = FALSE)

ics_sum_d14v16 <- process_iucn_summary_per_cell(spp_all_d14v16, 
                                                spp_cells = ics_d14,
                                                fn_tag = '_d14v16', reload = FALSE)

sum_cell_d14v16 <- process_means_per_cell(acs_sum_d14v16, ics_sum_d14v16, fn_tag = '_d14v16')

sum_rgn_d14v16  <- process_means_per_rgn(sum_cell_d14v16, rgn_cell_lookup, rgn_note = 'gl_d14v16')

### Create final outputs:
spp_status_d14v16 <- sum_rgn_d14v16 %>%
  dplyr::select(rgn_id, score = status)
spp_trend_d14v16 <- sum_rgn_d14v16 %>%
  dplyr::select(rgn_id, score = rgn_mean_trend)
write_csv(spp_status_d14v16, file.path(dir_git, scenario, 'output/spp_status_global_d14v16.csv'))
write_csv(spp_trend_d14v16,  file.path(dir_git, scenario, 'output/spp_trend_global_d14v16.csv'))


library(ggplot2)
status_2016   <- read_csv(file.path(dir_git, 'v2016', 'output/spp_status_global.csv'))
trend_2016    <- read_csv(file.path(dir_git, 'v2016', 'output/spp_trend_global.csv'))
status_2015   <- read_csv(file.path(dir_git, 'v2015', 'data/spp_status_global.csv'))
trend_2015    <- read_csv(file.path(dir_git, 'v2015', 'data/spp_trend_global.csv'))
status_d14v16 <- read_csv(file.path(dir_git, 'v2016', 'output/spp_status_global_d14v16.csv'))
trend_d14v16  <- read_csv(file.path(dir_git, 'v2016', 'output/spp_trend_global_d14v16.csv'))

rgn_cell_lookup <- extract_cell_id_per_region(reload = FALSE)

status_df <- status_2016 %>% rename(st_2016 = score) %>%
  full_join(status_2015 %>% rename(st_2015 = score), by = 'rgn_id') %>%
  full_join(trend_2016 %>% rename(tr_2016 = score), by = 'rgn_id') %>%
  full_join(trend_2015 %>% rename(tr_2015 = score), by = 'rgn_id') %>%
  full_join(status_d14v16 %>% rename(st_d14v16 = score), by = 'rgn_id') %>%
  full_join(trend_d14v16 %>% rename(tr_d14v16 = score), by = 'rgn_id') %>%
  full_join(rgn_cell_lookup %>% select(rgn_id, rgn_name) %>% unique(), by = 'rgn_id')

st_2016_2015 <- ggplot(status_df %>%
                         select(rgn_id, rgn_name, st_2016, v2015 = st_2015, v2016 = st_d14v16) %>%
                         gather(method, st_2015, c(v2016, v2015)), 
                       aes(x = st_2015, y = st_2016, color = method, key = rgn_name)) + 
  geom_point(alpha = .5) +
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  scale_x_continuous(limits = c(.5, 1)) +
  scale_y_continuous(limits = c(.5, 1)) +
  labs(x = 'Status: old data (d2014)',
       y = 'Status: new data (d2015)',
       title = 'SPP Status: v2015 vs v2016',
       color = 'Method for old data')
library(plotly)
ggplotly(st_2016_2015)
# print(st_2016_2015)
ggsave(file.path(dir_git, scenario, 'output/scatterplot_spp_status_global_d15v16_d14v15.png'),
       plot = st_2016_2015)

z <- read_csv(file.path(dir_git, scenario, 'output/rgn_spp_gl.csv')) %>%
  filter(pop_cat != 'DD' & !is.na(spatial_source)) %>%
  select(-sciname, -pop_cat, -spatial_source) %>%
  distinct() %>%
  group_by(rgn_id) %>%
  mutate(n_spp_rgn_valid = n())
zz <- z %>% 
  select(rgn_id, n_spp = n_spp_rgn_valid) %>%
  distinct()

st_d14v15_d14v16 <- ggplot(status_df %>%
                             left_join(zz, by = 'rgn_id'), 
                           aes(x = st_d14v16, y = st_2015, key = rgn_name, color = n_spp)) + 
  geom_point(alpha = .5) +
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  scale_x_continuous(limits = c(.5, 1)) +
  scale_y_continuous(limits = c(.5, 1)) +
  scale_color_gradient2(low = 'red', mid = 'blue', high = 'blue', midpoint = 2500) +
  labs(x = 'Status: old data, new method',
       y = 'Status: old data, new method',
       title = 'Status: old (d2014) data by new (v2016) and old (v2015) methods')
# library(plotly)
ggplotly(st_d14v15_d14v16)
# print(st_2016_d14v16)
ggsave(file.path(dir_git, scenario, 'output/scatterplot_spp_status_global_d14v15_d14v16.png'),
       plot = st_d14v15_d14v16)

# tr_2016v2015 <- ggplot(status_df, aes(x = tr_2015, y = tr_2016)) + 
#   geom_point() +
#   geom_abline(slope = 1, intercept = 0, color = 'red') +
#   scale_x_continuous(limits = c(-1, 1)) +
#   scale_y_continuous(limits = c(-1, 1)) +
#   labs(title = 'Trend: 2016 vs 2015')

# print(tr_2016v2015)

x <- read_csv(file.path(dir_git, scenario, 'debug/spp_gp_ranges.csv')) %>%
  mutate(label = ifelse(spp_group == 'BOTW', 'BIRDS!', 'not birds'))

y <- ggplot(x, aes(x = total_cells, y = gp_area_cat, color = label)) + 
  geom_point(size = 3, alpha = .8) + 
  labs(x = 'total area (# of cells)', 
       y = 'mean IUCN risk (0 = LC, 100 = EX)', 
       color = 'SPP group')
y

ggsave(file.path(dir_git, scenario, 'debug/birds_health.png'))

```