---
title: 'OHI: Species subgoal'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohiprep/src/templates/ohi_hdr.html'
  pdf_document:
    toc: true
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = FALSE, message = FALSE, warning = FALSE)

source('~/github/ohiprep/src/R/common.R')
library(readr)

goal      <- 'globalprep/spp_ico'
scenario  <- 'v2016'
dir_anx   <- file.path(dir_M, 'git-annex', goal) 
dir_goal  <- file.path('~/github/ohiprep', goal)

### set up provenance tracking for this script:
# source(file.path('~/github/ohibc', 'src/R/prov.R'))          

source(file.path(dir_goal, scenario, 'ico_fxn.R'))

if(!file.exists(file.path(dir_goal, 'README.md'))) {
  warning(sprintf('No README detected in %s', dir_git))
}
if(!file.exists(file.path(dir_goal, scenario, 'README.md'))) {
  warning(sprintf('No README detected in %s', file.path(dir_git, scenario)))
}
# SPP-specific and ICO-specific functions
```

# Summary:  Iconic Species Subgoal (Sense of Place)

This script prepares scores (status and trend) for Iconic Species in 
British Columbia's coastal regions.  Spatial data from IUCN and Aquamaps is
combined with extinction risk information from IUCN and conservation rank
info based on NatureServe categories.

The Iconic Species sub-goal model calculates a region's status based upon an unweighted average of species health for all 'iconic' species found within each BC reporting region.

From Halpern et al (2012):

> Iconic species are those that are relevant to local cultural identity through a species’ relationship to one or more of the following: 1) traditional activities such as fishing, hunting or commerce; 2) local ethnic or religious practices; 3) existence value; and 4) locally-recognized aesthetic value (e.g., touristic attractions/common subjects for art such as whales). Habitat-forming species are not included in this definition of iconic species, nor are species that are harvested solely for economic or utilitarian purposes (even though they may be iconic to a sector or individual). ...

> Ultimately, almost any species can be iconic to someone, and so the intent with this goal was to focus on those species widely seen as iconic within a country, and iconic from a cultural or existence value (rather than for a livelihoods or extractive reason). ...

> The reference point is to have the risk status of all assessed species as Least Concern (i.e., a goal score = 1.0)

The Status of this sub-goal (X~ICO~) is then the % of iconic species in each threat category (as defined by the IUCN Red List), such that:

$$X_{ICO} = \frac{\displaystyle\sum_{category}S_{cat}*w_{cat}}{\displaystyle\sum_{category}S_{cat}}$$

where for each IUCN threat category:

* *S~cat~* is the number of assessed species in the category
* *w~cat~* is the status weight assigned for that category (note, these are the inverse of the risk value used in the SPP calculations):
    * 'LC' = 1.0, 'NT' = 0.8, 'VU' = 0.6, 'EN' = 0.4, 'CR' = 0.2, 'EX' = 0.0

ICO trend is calculated in a similar manner, but weightings are assigned according to IUCN population trend: 'Decreasing' = -0.5, 'Stable' = 0.0, 'Increasing' = +0.5.  

# Updates from previous assessment

Changes since 2015 ICO subgoal for global OHI:

* The 2016 assessment now pulls data from the new IUCN API (http://apiv3.iucnredlist.org/api/v3/docs) instead of the old API.
* The trend calculations for 2016 are based on a ten-year linear trend of species risk status based on past assessment information, rather than population trend ('decreasing', 'stable', 'increasing').

***

# Data Sources

**List of iconic species:**

**Species native country information:**

* __Reference__: 
    * IUCN 2015. The IUCN Red List of Threatened Species. Version 2015-4. <http://www.iucnredlist.org>.
        * Shapefiles available from: http://www.iucnredlist.org/technical-documents/spatial-data
        * __Downloaded__: 21 December 2015.
* __Native data resolution__: NA
* __Time range__: NA
* __Format__:  Shapefile

**Species extinction risk information:**

* __Reference__: 
    * IUCN 2015. The IUCN Red List of Threatened Species. Version 2015-4. <http://www.iucnredlist.org>.
        * Shapefiles available from: http://www.iucnredlist.org/technical-documents/spatial-data
        * __Downloaded__: 21 December 2015.
* __Native data resolution__: NA
* __Time range__: NA
* __Format__:  Shapefile

***
  
# Methods

## get master list of Iconic Species

``` {r get_ico_spp_list}
ico_list_raw <- get_ico_list()

#                   comname                  sciname ico_gl ico_rgn_id
# 1      Peruvian anchoveta        Engraulis ringens  FALSE        138
# 2 Mediterranean monk seal        Monachus monachus  FALSE        183
# 3     Smooth-coated Otter  Lutrogale perspicillata  FALSE        203
# 4            Ganges Shark       Glyphis gangeticus  FALSE        203
# 5       Pygmy right whale        Caperea marginata  FALSE         16
# 6   Antarctic minke whale Balaenoptera bonaerensis  FALSE         16
### * ico_rgn_id: rgn_id in which species is iconic by regional/national
###   lists; if globally iconic, ico_rgn_id <- NA
```

`r DT::datatable(ico_list_raw, caption = 'Iconic species list')`

***

``` {r setup functions}

get_from_api <- function(url, param) {
    api_info <- fromJSON(sprintf(url, param, api_key)) %>%
      data.frame()
}

mc_get_from_api <- function(url, param_vec) {
  numcores = ifelse(Sys.info()[['nodename']] == 'mazu', 12, 1)
  out_df <- parallel::mclapply(param_vec, 
                          function(x) get_from_api(url, x),
                          mc.cores = numcores) %>% 
    rbind_all() 
  out_df <- out_df %>%
    setNames(names(.) %>%
               str_replace('result.', ''))
}
```

``` {r get_spp_info}
### Get all pages and bind into total species list

library(jsonlite)
api_key <- 'fb71ae836f415f04f41176f6d30c4a9e4cea620d46b9e5021bf2fb142ea51bf5'

spp_npage_url <- sprintf('http://apiv3.iucnredlist.org/api/v3/speciescount?token=%s', api_key)
n_spp <- fromJSON(spp_npage_url) %>%
  .$count %>% as.integer()
n_pages <- ceiling(n_spp/10000)

spp_page_url <- 'http://apiv3.iucnredlist.org/api/v3/species/page/%s?token=%s'
spp_df_all <- mc_get_from_api(spp_page_url, c(0:(n_pages - 1)))

spp_df_all <- spp_df_all %>%
  dplyr::select(-infra_rank, -infra_name, -count, -page) %>%
  rename(iucn_sid = taxonid, sciname = scientific_name) %>%
  setNames(names(.) %>%
             str_replace('_name', ''))

write_csv(spp_df_all, file.path(dir_goal, scenario, 'int/spp_list_from_api.csv'))

```

``` {r combine_iucn_spp_info_with_ico_list}

spp_df_all <- read_csv(file.path(dir_goal, scenario, 'int/spp_list_from_api.csv'))

spp_ico <- spp_df_all %>% 
  filter(sciname %in% ico_list_raw$sciname) 

spp_missing <- ico_list_raw %>% 
  filter(!sciname %in% spp_ico$sciname)
# none! :)

ico_list <- ico_list_raw %>%
  left_join(spp_ico %>% 
              select(iucn_sid, sciname, subpop = population, cat = category),
            by = 'sciname')
            
```

``` {r get_country_list}
### for each species ID, get country list
ico_country_url <- 'http://apiv3.iucnredlist.org/api/v3/species/countries/id/%s?token=%s'

ico_spp_countries <- mc_get_from_api(ico_country_url, ico_list$iucn_sid) 

ico_spp_countries <- ico_spp_countries %>% 
  rename(iucn_sid = name1) %>%
  mutate(iucn_sid = as.integer(iucn_sid),
         country  = str_trim(country)) %>% 
  left_join(ico_list %>% 
              select(iucn_sid, sciname) %>%
              distinct(),
            by = 'iucn_sid')

write_csv(ico_spp_countries, file.path(dir_goal, scenario, 'int/ico_spp_countries_raw.csv'))

```

``` {r clean_up_country_list}

ico_spp_rgn <- read_csv(file.path(dir_goal, scenario, 'int/ico_spp_countries_raw.csv'))

rgn_iucn2ohi <- read_csv(file.path(dir_goal, scenario, 'data/rgns_iucn2ohi.csv'))

ico_spp_rgn <- ico_spp_rgn %>%
  rename(iucn_rgn_name = country) %>%
  left_join(rgn_iucn2ohi,
            by = 'iucn_rgn_name') %>%
  rename(rgn_name = ohi_rgn_name) %>%
  select(-iucn_rgn_name, -count, -code)

write_csv(ico_spp_rgn, file.path(dir_goal, scenario, 'int/ico_spp_rgn.csv'))

```

***

``` {r get_category_list}
### for each species ID, get past assessments
ico_past_assess_url <- 'http://apiv3.iucnredlist.org/api/v3/species/history/id/%s?token=%s'

ico_assess_raw <- mc_get_from_api(ico_past_assess_url, ico_list$iucn_sid) 

ico_assess_raw <- ico_assess_raw %>% 
  rename(iucn_sid = name) %>%
  mutate(iucn_sid = as.integer(iucn_sid),
         year     = as.integer(year)) %>% 
  left_join(ico_list %>% 
              select(iucn_sid, sciname) %>%
              distinct(),
            by = 'iucn_sid')

write_csv(ico_assess_raw, file.path(dir_goal, scenario, 'int/ico_assessments_raw.csv'))

```

***

``` {r clean_up_category_list}
### Clean up the time series
### iucn_sid | year | code | category | sciname

ico_assess_raw <- read_csv(file.path(dir_goal, scenario, 'int/ico_assessments_raw.csv'))

ico_assess <- ico_assess_raw %>%
  rename(cat = code, cat_txt = category) %>%
  mutate(cat = toupper(cat),
         cat = str_replace(cat, 'LR/', ''),
         cat = ifelse(cat %in% c('K', 'I'), 'DD', cat),
         cat = ifelse(cat == 'NR', 'NE', cat),
         cat = ifelse(str_detect(toupper(cat_txt), 'VERY RARE'), 'CR', cat),
         cat = ifelse(str_detect(toupper(cat_txt), 'LESS RARE'), 'T', cat),
         cat = ifelse(str_detect(toupper(cat_txt), 'STATUS INADEQUATELY KNOWN'), 'DD', cat),
         cat = ifelse(cat == 'V', 'VU', cat), 
         cat = ifelse(cat == 'E', 'EN', cat))

### reclassifications:
#  LC <- "LOWER RISK/LEAST CONCERN (LR/LC)"                         
#  NT <- "LOWER RISK/NEAR THREATENED (LR/NT)"                       
#  T  <- "THREATENED (T)" treat as "EN"
#  VU <- "VULNERABLE (V)"                                           
#  EN <- "ENDANGERED (E)"                                           
#  LR/CD <- "LOWER RISK/CONSERVATION DEPENDENT (LR/CD)" treat as between VU and NT
#  CR <- "VERY RARE AND BELIEVED TO BE DECREASING IN NUMBERS"       
#  T  <- "LESS RARE BUT BELIEVED TO BE THREATENED-REQUIRES WATCHING"
#  DD <- "INSUFFICIENTLY KNOWN (K)"                                 
#  DD <- "INDETERMINATE (I)"                                        
#  DD <- "STATUS INADEQUATELY KNOWN-SURVEY REQUIRED OR DATA SOUGHT" 
#  NE <- "NOT RECOGNIZED (NR)"         
pop_cat <- data.frame(cat       = c("LC", "NT", "VU", "EN", "CR", "EX", "T", "CD", "NE", "DD"), 
                      cat_score = c(   0,  0.2,  0.4,  0.6,  0.8,  1.0, 0.6,  0.3,   NA,  NA))
  
ico_assess <- ico_assess %>% 
  left_join(pop_cat, by = 'cat') %>%
  filter(!is.na(cat_score) & !is.na(year)) %>%
  arrange(iucn_sid, year)

write_csv(ico_assess, file.path(dir_goal, scenario, 'int/ico_assess_clean.csv'))

```

***

``` {r fill_out_category_time_series}

ico_assess <- read_csv(file.path(dir_goal, scenario, 'int/ico_assess_clean.csv'))

ico_assess_full <- ico_assess %>%
  select(-sciname) %>%
  arrange(iucn_sid, year) %>%
  complete(year = full_seq(year, 1), nesting(iucn_sid)) %>%
  group_by(iucn_sid) %>%
  fill(cat, cat_txt, cat_score) %>% ### fills all the way to latest year (2015)
  ungroup()

ico_spp_cat <- ico_list %>% 
  rename(cat_2016 = cat) %>%
  left_join(ico_assess_full, by = c('iucn_sid'))

### if no time series available, time series years will be NA.  Assign a list to
### those NAs, then unnest it to create observations for those years.
ico_spp_cat <- ico_spp_cat %>%
  mutate(year = ifelse(is.na(year), 
                       list(c(min(year, na.rm = TRUE):max(year, na.rm = TRUE))), 
                       year)) %>% 
  unnest(year)

### NAs will be filled backward in time by starting from the most recent non-NA.
### To do this, we'll swap any current-year NAs with the cat_score (meaning no
### time series fill), and fill upwards instead of downwards.
ico_spp_cat <- ico_spp_cat %>%
  left_join(pop_cat %>% 
              rename(cat_2016 = cat, cat_2016_score = cat_score), 
            by = 'cat_2016') %>%
  mutate(cat_score = ifelse(year == max(year, na.rm = TRUE) & is.na(cat), 
                            cat_2016_score, 
                            cat_score)) %>%
  arrange(iucn_sid, year) %>%
  group_by(iucn_sid) %>%
  fill(cat, cat_score, cat_txt, .direction = 'up') %>%
  ungroup()
  
write_csv(ico_spp_cat, file.path(dir_goal, scenario, 'int/ico_spp_cat.csv'))

```

***

``` {r combine_ico_spp_cat_with_countries}

ico_list_ts_abbr <- read_csv(file.path(dir_goal, scenario, 'int/ico_spp_cat.csv')) %>%
  select(iucn_sid, sciname, year, cat, cat_score, ico_gl, ico_rgn_id) %>%
  filter(year >= 2000)

ico_spp_rgn <- read_csv(file.path(dir_goal, scenario, 'int/ico_spp_rgn.csv'))

ico_spp_rgn_cat <- ico_list_ts_abbr %>% 
  full_join(ico_spp_rgn, by = c('iucn_sid', 'sciname'))

# ico_2015 <- ico_spp_rgn_cat %>%
#   filter(year == 2015)
# 
# ex <- ico_spp_rgn_cat %>%
#   filter(str_detect(tolower(distribution_code), 'extinct')) %>%
#   filter(year == 2015)
# 
# ex2 <- ico_spp_rgn_cat %>%
#   filter(sciname %in% ex$sciname) %>%
#   filter(rgn_name %in% ex$rgn_name) %>%
#   filter(year == 2015)
### How to deal with "extinct" locally?  when did species go extinct? 
### But we're only really looking at the last ten-fifteen years, so
### maybe not important - just set all years to extinct for that region

ico_spp_rgn_cat <- ico_spp_rgn_cat %>%
  mutate(cat = ifelse(str_detect(presence, '^Extinct'), 'EX', cat), ### ^ indicates start of string
         cat_score = ifelse(cat == 'EX', 1, cat_score)) %>%
  filter(ico_gl | ico_rgn_id == rgn_id) ### Keep (all globally iconic) and (regionally iconic in region only)

write_csv(ico_spp_rgn_cat, file.path(dir_goal, scenario, 'int/ico_spp_rgn_cat.csv'))

```

***
``` {r calc_status_and_trend}
### Report and summarize regional iconic species status

ico_spp_rgn_cat <- read_csv(file.path(dir_goal, scenario, 'int/ico_spp_rgn_cat.csv'))

# Report out for toolbox format (rgn_id | sciname | category or popn_trend for each species within a region).
# Note: in toolbox, group_by(rgn_id, sciname) and then summarize(category = mean(category)) to
#   average any parent/subpop species listings before aggregating to overall average per region.

ico_status_raw <- ico_spp_rgn_cat %>%
  select(rgn_id, rgn_name, sciname, iucn_sid, cat, cat_score, year) %>%
  arrange(rgn_id, desc(year), sciname)

ico_status_calc <- ico_status_raw %>%
  group_by(rgn_id, rgn_name, year) %>%
  summarize(mean_cat = round(mean(cat_score, na.rm = TRUE), 5), ### since DDs are still in, use na.rm = TRUE
            ico_status = 1 - mean_cat,
            n_spp = n()) 


ico_trend <- data.frame()
for (i in 2010:max(ico_status_calc$year, na.rm = TRUE)) { # i <- 2013
  tmp_status <- ico_status_calc %>%
    filter(year <= i & year > (i - 10)) ### trend based on 10-year average since assessments are sporadic
  tmp_trend <- tmp_status %>%
    group_by(rgn_id) %>%
    do(trend_lm = lm(ico_status ~ year, data = .)$coefficients[2]) %>%
    mutate(year  = i,
           trend_lm  = as.numeric(trend_lm),
           ico_trend = round(trend_lm * 5, 5))
  ico_trend <- ico_trend %>%
    bind_rows(tmp_trend)
}

ico_sum <- ico_status_raw %>%
  left_join(ico_status_calc, by = c('rgn_id', 'rgn_name', 'year')) %>%
  left_join(ico_trend, by = c('rgn_id', 'year'))

write_csv(ico_sum, file.path(dir_goal, scenario, 'summary/ico_summary.csv'))
# Report out for finalized status and trend values per region.

write_csv(ico_status_raw,  file.path(dir_goal, scenario, 'output/ico_status_raw.csv'))
write_csv(ico_status_calc, file.path(dir_goal, scenario, 'output/ico_status_calc.csv'))
write_csv(ico_trend,       file.path(dir_goal, scenario, 'output/ico_trend.csv'))

```

### Iconic Species full list (year == 2015)

`r DT::datatable(ico_status_raw %>% filter(year == 2015))`

### Iconic Species processed status and trend by region (year == 2015)

`r DT::datatable(ico_status_calc %>% filter(year == 2015), caption = 'ICO status')`

`r DT::datatable(ico_trend %>% filter(year == 2015) %>% select(-trend_lm), caption = 'ICO trend')`

***

### Plot scores time series

``` {r spp_plot_scores_over_time, fig.height = 4, fig.width = 6, fig.align = 'center'}
library(ggplot2)
library(plotly)

status_ts_plot <- ggplot(ico_sum %>%
                           filter(!is.na(rgn_id)),
                         aes(x = year, y = ico_status, color = rgn_id, group = rgn_id)) +
#  ggtheme_plot +
  geom_line(size = 2) +
#  scale_colour_brewer(palette = 'PRGn') +
  labs(x = 'year',
       y = 'ICO status',
       title = 'ICO status over time',
       color = 'Region')

ggplotly(status_ts_plot)

trend_ts_plot <- ggplot(ico_sum %>%
                           filter(!is.na(rgn_id) &!is.na(ico_trend)),
                         aes(x = year, y = ico_trend, color = rgn_id, group = rgn_id)) +
#  ggtheme_plot +
  geom_line(size = 2) +
#  scale_colour_brewer(palette = 'PRGn') +
  labs(x = 'year',
       y = 'ICO trend',
       title = 'ICO trend over time',
       color = 'Region')

ggplotly(trend_ts_plot)
```

***

### Compare scores OHI 2016 vs OHI 2015 

``` {r spp_plot_scores_vs_2015, fig.height = 4, fig.width = 6, fig.align = 'center'}

ico_2015 <- read_csv('~/github/ohi-global/eez2013/scores.csv') %>%
  filter(goal == 'ICO' & dimension %in% c('status', 'trend')) %>%
  spread(key = dimension, value = score) %>%
  select(-goal)

ico_15_16 <- ico_sum %>%
  filter(year == max(year)) %>% 
  select(rgn_id, rgn_name, st_16 = ico_status, tr_16 = ico_trend, n_spp) %>%
  distinct() %>%
  full_join(ico_2015 %>%
              rename(rgn_id = region_id, st_15 = status, tr_15 = trend),
            by = 'rgn_id') %>%
  mutate(st_16 = st_16 * 100)
  
st_15_16_plot <- ggplot(ico_15_16 %>%
                           filter(!is.na(rgn_id)),
                         aes(x = st_15, y = st_16, color = n_spp, key = rgn_name, key2 = rgn_id)) +
  geom_point(alpha = .6) +
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  labs(x = 'ICO status v2015',
       y = 'ICO status v2016',
       title = 'ICO status comparison')

ggplotly(st_15_16_plot)
ggsave(file.path(dir_goal, scenario, 'int/scatterplot_status_15v16.png'))

tr_15_16_plot <- ggplot(ico_15_16 %>%
                           filter(!is.na(rgn_id)),
                         aes(x = tr_15, y = tr_16, color = n_spp, key = rgn_name, key2 = rgn_id)) +
  geom_point(alpha = .6) +
  geom_abline(slope = 1, intercept = 0, color = 'red') +
  labs(x = 'ICO trend v2015',
       y = 'ICO trend v2016',
       title = 'ICO trend comparison')

ggplotly(tr_15_16_plot)
ggsave(file.path(dir_goal, scenario, 'int/scatterplot_trend_15v16.png'))

```

``` {r eval = FALSE}

# ICO outlier check:
# 
#   1 Cocos Island 60 -> 75
#  59 Belgium      63 -> 60
# 164 Belize       60 -> 73
# 208 Singapore    60 -> 77
# 210 Japan        53 -> 63

rgns <- c(1, 59, 164, 208, 210)
x15st_tr <- read_csv('~/github/ohi-global/eez2013/layers/ico_spp_extinction_status.csv') %>%
  filter(rgn_id %in% rgns) %>%
  full_join(read_csv('~/github/ohi-global/eez2013/layers/ico_spp_popn_trend.csv') %>%
              filter(rgn_id %in% rgns),
            by = c('rgn_id', 'sciname'))
x16st_tr <- ico_sum %>%
  filter(year == 2015) %>%
  filter(rgn_id %in% rgns) %>%
  select(-year, -trend_lm)

write_csv(x15st_tr, file.path(dir_goal, 'tmp_15.csv'))
write_csv(x16st_tr, file.path(dir_goal, 'tmp_16.csv'))

```

***