---
  title: 'OHI 2016: Sea Level Rise Pressure layers prep'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
  toc: true
number_sections: true
theme: cerulean
highlight: haddock
includes: 
  in_header: '~/github/ohiprep/src/templates/ohi_hdr.html'
pdf_document:
  toc: true
---
  
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = FALSE, message = FALSE, warning = FALSE)

library(sp)        # the classes and methods that make up spatial ops in R
library(rgdal)
library(raster)
library(rasterVis)
library(ncdf4)
library(RColorBrewer)
library(doParallel)
library(foreach)
library(parallel)


dir_git <- '~/github/ohiprep'
source('~/github/ohiprep/src/R/common.R')

dir_anx <- file.path(dir_M, 'git-annex/globalprep')

### set up proj4string options: WGS84
p4s_wgs84 <- '+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'

### Define spectral color scheme for plotting maps
cols      = rev(colorRampPalette(brewer.pal(9, 'Spectral'))(255)) # rainbow color scheme


# define mollweide projection

mollCRS <- CRS('+proj=moll')

```

# Summary

There are two parts to creating this layer:
  
  1. Data prep to get raw data into the correct format:
  * If necessary, read .nc.gz files from aviso: ftp://ftp.aviso.altimetry.fr/global/delayed-time/grids/climatology/monthly_mean/
  * unzip .gz files and then delete .gz 
* Read monthly mean sea level data as rasters; aggregate to annual means
2. Creating the pressure layers for OHI Global:
  * Determine baseline using five-year average from 1993-1997.
* For each study year (2000 - 2015) calculate difference between study year and baseline
* Set to zero all negative values, indicating decreases in mean sea level
* Resample data from the native cell resolution (0.25 degrees) to ~ 1km
* Set reference point as the 99.99th quantile of the data distribution to rescale all values from 0 to 1
* Use thin plate spline interpolation to fill in NA values
* Clip results to 1 km offshore region, and determine average SLR pressure per region.

This process is completed entirely within this script.

# Data

Reference: The altimeter products were produced and distributed by Aviso (http://www.aviso.altimetry.fr/), as part of the Ssalto ground processing segment. [AVISO MSLA heights, monthly means](http://www.aviso.altimetry.fr/en/data/products/sea-surface-height-products/global/msla-mean-climatology.html)

* Downloaded: March 18, 2016
* Description: Monthly mean sea level anomaly
* Native data resolution: 0.25 degree grid cells
* Time range: January 1993 - August 2015
* Format: NetCDF

# Methods

## Data Prep

### Download files

``` {r download_files_and_unzip}

reload <- TRUE

dir_anx_aviso <- file.path(dir_M, 'git-annex/globalprep/_raw_data/AVISO_slr/d2016')

if(reload) {
  ### Get filenames from AVISO FTP URL:
  library(RCurl)
  url <- "ftp://ftp.aviso.altimetry.fr/global/delayed-time/grids/climatology/monthly_mean/"
  userpwd <- "ucsb_afflerbach:sioped54j"
  filenames <- getURL(url,
                      userpwd = userpwd,
                      ftp.use.epsv = FALSE,
                      dirlistonly = TRUE) %>%
    str_split('\n') %>%
    unlist()
  
  filenames <- filenames[!str_detect(filenames, '.png')] %>%
    sort()
  
  
  ### Set up loop to download each file and save to git-annex:
  
  ftp_dir <- 'ftp://ucsb_afflerbach:sioped54j@ftp.aviso.altimetry.fr/global/delayed-time/grids/climatology/monthly_mean'
  for (i in 274:length(filenames)) { # nc_file <- filenames[2]
    print(i)
    nc_file <- filenames[i]
    
    download.file(url      = file.path(ftp_dir, nc_file),
                  destfile = file.path(dir_anx_aviso, 'msla_monthly_mean', nc_file),
                  mode     = 'wb')
  }
  
  zipfiles <- list.files(file.path(dir_anx_aviso, 'msla_monthly_mean'),
                         full.names = TRUE)
  zipfiles <- zipfiles[str_detect(zipfiles, '.gz')]
  for (zipfile in zipfiles) { # zipfile <- zipfiles[1]
    message('Unzipping file: ', zipfile)
    R.utils::gunzip(zipfile, remove = TRUE, skip = TRUE)
  }
}
```


### Clip data to coastal cells

All NetCDF files for each month over the 22 years is rasterized and all non-coastal bordering cells are removed. To do this we use `raster::boundaries()` with the `type = 'inner'` which turns all coastal cells bordering land to 1 and the rest of the cells to 0.


```{r clip}

#list all netCDF files
nc_files <- list.files(file.path(dir_anx_aviso, 'msla_monthly_mean'),
                       full.names = TRUE,pattern='.nc')

ocean_big <- aggregate(ocean,c(25100, 30900))

#--------------------------------------------------------------------
## Testing boundaries() to get only coastal cells

    r <- raster(nc_files[1])%>%
      writeRaster(filename=file.path(dir_anx,'prs_slr/v2016/test/r.tif'))
    
    r_in <- boundaries(r,type='inner',asNA=T)%>%
            rotate()%>%
            writeRaster(filename=file.path(dir_anx,'prs_slr/v2016/test/r_inner_boundaries.tif'),overwrite=T)
    #r_out <- boundaries(r,type='outer',asNA=T,filename=file.path(dir_anx,'prs_slr/v2016/test/r_outer_boundaries.tif'))

## Using type='inner' we get all coastal cells bordering land. This is what we want to use for this pressure layer


#--------------------------------------------------------------------

## Function that rotates each monthly file, sets the long/lat projection, and keeps only coastal cells - saved to GitHub
months_coast <- function(x){
  
  m_yr <- substr(x,108,115)
    
  #read in month raster
  r <- raster(x)%>%rotate()
  
  #define projection of the raster before reprojecting
  projection(r) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
  
  r_mask <- mask(r,r_in,progress='text')

  
  writeRaster(r_mask,filename=paste0(dir_git,'/globalprep/prs_slr/v2016/int/msla_monthly_coast/msla_monthly_coast_',m_yr,'.tif'),overwrite=T)

}

#apply the clipping function to all 272 files
mclapply(nc_files,months_coast,mc.cores=14)

```

```{r years}

#years of interest

years <- c(2010:2014)

months <- c(1:12)

```

### Explore the data

#### 1. How does mean sea level vary over months?

```{r explore}

#Looking at means and sd over time for each month

files <- list.files(file.path(dir_git,
                              'globalprep/prs_slr/v2016/int/msla_monthly_coast'),full.names=T)

for (i in years){
  
  print(i)
  
  #get all files for the year i
    yr_files <- files[which(substr(files,100,103)==i)]%>%stack()

    names(yr_files) <- months
    #plot & save
    
    png(paste0('figs/monthly_boxplot_',i,'.png'), width = 6, height = 4, units = 'in', res = 300 )
    
    print(bwplot(yr_files,xlab = 'Months',ylab = 'Mean Sea Level Anomaly (meters)'))

    dev.off()
}

img <- png::readPNG('figs/month-boxplots-2011-2014.png')



```

#### 2. What is the distribution of mean and sd mean sea level from January 2010 - December 2014

```{r}


  #get all months for all years 2010-2014
    yr_m_files <- files[which(substr(files,100,103) %in% years)]

  #get unique names for each layer
    layer_names <- unique(basename(yr_m_files) %>% 
                      str_replace('msla_monthly_coast_', '') %>%
                      substr(1, 8)) %>%
                      sort()
  #stack into a raster stack (nlayers = 60) and assign names
    s <- stack(yr_m_files)
    
    names(s) <- layer_names
    
    #calculate the mean
    
    m <- calc(s,fun=mean,na.rm=T)
    
    #calculate the sd
    
    std <- calc(s,fun=function(x){sd(x,na.rm=T)})
    
    #plot + hist
    
    #plot & save
    
    png('figs/mean_sd_msla_2010_2014.png', width = 10, height = 4, units = 'in', res = 300 )
    par(mfrow=c(1,2),
         mai = c(1, 0.5, 1, 1))
    
    plot(m,axes=F,col=cols,box=F,colNA = 'dimgray',main='Mean SLA 2010-2014')
    plot(std,axes=F,col=cols,box=F,colNA = 'dimgray',main = 'Standard Deviation SLA 2010-2014')
    
    dev.off()
    
    h <- stack(m,std)
    names(h)<-c("Mean SL 2010-2014","St. Dev. 2010-2014")
    
    png('figs/mean_sd_msla_hist_2010_2014.png',width = 8, height = 4, units = 'in', res = 300)
    histogram(h)

    dev.off()

```



### Read raw data as monthly means and convert to annual means

Raw data in monthly mean sea level anomalies, at .25 degree grid globally, is processed into annual mean sea level anomalies.  These files are saved in git-annex as GeoTIFFs for later use.

``` {r msla_monthly_to_annual}

all_yrs <- c(1993:2014)

for (j in 1:length(all_yrs)) {
  

  msla_yr <- files[str_detect(files, as.character(all_yrs[j]))]
  ### set up an empty list to store rasters for this year
  rast_list_yr <- vector('list', length(msla_yr))
  
  message('Generating annual MSLA raster for ', all_yrs[j])
  
  ### Loop over all files for this year, read raster and assign to list for this year
  for (i in 1:length(msla_yr)) { ### i = 1
    rast_list_yr[i] <- raster(msla_yr[i], nogit = TRUE)
  }
  
  ### brick all rasters for this year, and calc annual mean, then write as raster
  rbrick_yr <- brick(rast_list_yr)
  
  rast_annual_mean <- calc(rbrick_yr, mean,na.rm=T)
  #rast_annual_sd <- calc(rbrick_yr,fun=function(x){sd(x,na.rm=T)})
  
  rast_file <- file.path(dir_git, 'globalprep/prs_slr/v2016/int/msla_annual_mean', 
                         sprintf('rast_msla_annual_%s.tif', all_yrs[j]))
  #rast_file_sd <- file.path(dir_git, 'globalprep/prs_slr/v2016/int/msla_annual_sd', 
                         #sprintf('rast_msla_annual_sd_%s.tif', years[j]))
  
  message('Writing raster to ', rast_file)
  writeRaster(rast_annual_mean,
              rast_file,
              overwrite = TRUE,
              nogit = TRUE)

}

```

### Get reference point to rescale each annual raster layer

We want to use the reference period, 1993-2012, since this time period was used to calculate the raw data.

First calculate mean anomaly over this time period per cell, then get the 99.99th quantile.

```{r ref}

#stack all annual mean rasters from 1993 - 2012

ref_list <- list.files(file.path(dir_git,'globalprep/prs_slr/v2016/int/msla_annual_mean'),pattern = '*.tif',full.names=T)[1:20]

ref_mean <- stack(ref_list)%>%
              calc(fun=function(x){mean(x,na.rm=T)})
  
histogram(ref_mean)
ref <- quantile(ref_mean,prob=0.9999,na.rm=T)


```

#### Compare mean to aggregate dataset from Aviso

```{r}

ref_list_2 <- list.files(file.path(dir_git,'globalprep/prs_slr/v2016/int/msla_annual_mean'),pattern = '*.tif',full.names=T)[1:22]

all_yr_mean <- stack(ref_list_2)%>%
              calc(fun=function(x){(mean(x,na.rm=T))*0.0001})

agg <- raster(file.path(dir_anx_aviso,'MSL_Map_MERGED_Global_IB_RWT_NoGIA_Adjust.nc'))%>%
        rotate() #divide by 23 years to get mean mm per year

agg_mask <- mask(agg,all_yr_mean)

compare <- agg - all_yr_mean


```

The reference point is `r ref*100` mm per year (~2 inches).

### Rescale each cell according to this reference point.
```{r rescale}

rescale_slr <- function(x){
  
  r <- raster(x)
  
  yr = substr(names(r)[1],18,21)
  
  out <- calc(r,fun=function(x){ifelse(x>ref,1,x/ref)})

}

l <- list.files(file.path(dir_git,'globalprep/prs_slr/v2016/int/msla_annual_mean'),pattern = '*.tif',full.names=T)
mclapply(l,rescale_slr,mc.cores=14)

```


### Reproject to Mollweide and resample to 1km

```{r reproj_resamp}

# grab all annual means from 2006 - 2014
annual <- list.files(file.path(dir_git,'globalprep/prs_slr/v2016/int/msla_annual_mean'),full.names = T,pattern='.tif')[1:22]


# Read in ocean raster with cells at 1km. Use this as a template for resampling

ocean = raster(file.path(dir_M,'git-annex/globalprep/spatial/ocean.tif'))


registerDoParallel(10) 

#utilize parallelization to rescale & reproject all rasters

foreach(i = 1993:2010) %dopar% {
  
  yrs <- c(i,i+1,i+2,i+3,i+4)

  r <- annual[which(substr(annual,96,99)%in%yrs)]%>%
        stack()%>%
          mean(na.rm=T)
  
  r[r<0]<-0
  
  #rescale & save to file
  
  r_resc <- calc(r,fun=function(x){ifelse(x>ref,1,x/ref)},filename = 
                   paste0(file.path(dir_git),'/globalprep/prs_slr/v2016/int/msla_annual_rescale/msla_annual_rescaled_',yrs[1],'-',yrs[5],'.tif'),overwrite=T)
  
  #define projection of the raster before reprojecting
  projection(r_resc) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"

  #reproject raster to mollweide, resample to 1km
  r <- projectRaster(r_resc,crs = mollCRS,over=T)%>%
          resample(ocean, method = 'ngb', 
                   filename = paste0(file.path(
                      dir_anx),'/prs_slr/v2016/output/slr_moll_1km_rescaled_',yrs[1],'-',yrs[5],'.tif'),overwrite=T)
}


```

### Plot

```{r plot}

resc <- list.files(file.path(dir_git,'globalprep/prs_slr/v2016/int/msla_annual_rescale'),full.names = T)%>%stack()

plot(resc,col=cols)

names <-substr(names(resc),22,30)

names(resc)<-names

histogram(resc)

#look at non-rescaled data dist

a <- stack(annual)
histogram(a)


out <- 
  
plot(ocean,col='cornsilk2',axes=F,box=F,main='Sea Level Rise Pressure 2010-2014',legend=F)
plot(out[[5]],col=cols,axes=F,box=F,add=T)
```










## Rescale values as pressures

Baseline sea level anomaly is based upon the count within 1993 to 1997.

Change in mean sea level is determined by simple subtraction of the 1998-2014 rasters from the 1993-1997 mean raster.

``` {r calc_delta_sl_anomalies}

base_yrs <- 5
names(rbrick_sla)[1:base_yrs]
sla_baseline <-rbrick_sla[[1:base_yrs]] %>%
  calc(mean,na.rm=T,filename = file.path(dir_M,'git-annex/globalprep/prs_slr/v2016/int/sla_baseline.tif'),overwrite=T)


sla_study <- rbrick_sla[[(base_yrs + 1):nlayers(rbrick_sla)]]

sla_delta <- sla_study - sla_baseline
names(sla_delta) <- names(sla_study)

#sla_delta_file <- file.path(dir_M, 'git-annex/globalprep/prs_slr/v2016/int/rast_sla_delta_raw.tif')
#writeRaster(sla_delta, sla_delta_file, bandorder='BIL', overwrite = TRUE) ### Saving a brick to file which includes years 1998 - 2015

plot(sla_delta[[1]], 
          main = sprintf('Change in SL anomaly %s', names(sla_delta)[1]), 
          col=cols,box=F,axes=F)
plot(sla_delta[[nlayers(sla_delta)]], 
          main = sprintf('Change in SL anomaly %s', names(sla_delta)[nlayers(sla_delta)]), 
          col=cols,axes=F,box=F)

```


## Reproject to molleweide & Resample to 1km2 cells

```{r reproject_annual_changes, dependson=calc_delta_sl_anomalies, eval=F}

# Read in ocean raster with cells at 1km. Use this as a template for resampling

ocean = raster(file.path(dir_M,'git-annex/globalprep/spatial/ocean.tif'))

## Multiprocessing cores
# best to leave empty arguments; by default, the number of cores used for parallel 
# execution is 1/2 the number of detected cores (if number is unspecified)
# You can also detect the number of Cores: ?detectCores
registerDoParallel(16) 

#do some parallelization to reproject all rasters
foreach(i = 1:nlayers(sla_delta)) %dopar% {
  
  r <- sla_delta[[i]]
  
  projection(r) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"

  r <- projectRaster(r,crs = mollCRS,over=T)%>%
          resample(ocean,filename = paste0(file.path(dir_anx, 'prs_slr/v2016/int/sla_delta_moll_1km/sla_delta_moll_1km_'),
                                              names(sla_delta[[i]]),'.tif'),overwrite=T)
}

```

Values below zero (drop in sea level) are set to zero.

This difference value is rescaled from zero (no sea level rise) to 1 (at 99.99%ile of local slr).

Pull in reference point from csv
``` {r drop_negative_values}

slr_moll <- list.files(file.path(dir_M,'git-annex/globalprep/prs_slr/v2016/int/sla_delta_moll_1km'),full.names=T)

slr_ref <- read.csv('../reference_points_pressures.csv',stringsAsFactors=F)%>%
            filter(pressure == 'Sea Level Rise')%>%
              dplyr::select(ref_point)

slr_ref = as.numeric(slr_ref$ref_point[[1]])/1000 #turn mm into m

resc <- function(x){
  
  yr <- substr(x,nchar(x)-7,nchar(x)-4)
  r <- raster(x)
  
  r[r<0]<-0

  slr_resc <- calc(r, fun = function(x) ifelse(x > slr_ref, 1, x/slr_ref))
  
  writeRaster(slr_resc,filename = paste0(file.path(dir_anx),'/prs_slr/v2016/int/sla_resc/slr_rescaled_',yr,'.tif')) #add year
}

mclapply(slr_moll,resc,mc.cores=14)

```


Interpolate change in sea level values, to fill gaps.
``` {r interpolate_bc_sla}

plot_rast(bc_slr_int[[1]], 
          title = sprintf('delta SL anomaly %s', names(bc_slr_int)[1]), 
          scale_label = 'meters',
          rev_scale = TRUE)
plot_rast(bc_slr_int[[nlayers(bc_slr_int)]], 
          title = sprintf('delta SL anomaly %s', names(bc_slr_int)[nlayers(bc_slr_int)]), 
          scale_label = 'meters',
          rev_scale = TRUE)

```


## Calculate mean pressure by region by year

Using 1000 m region ID raster to run zonal statistics on the pressures layers, calculate the regional average pressures for the study period, and save as a .csv.

``` {r calc_mean_pressures_by_year}

slr_rgn_mean <- zonal(bc_slr_resc, rast_rgn, fun = 'mean') %>%
  as.data.frame() %>%
  rename(pressure = mean, rgn_id = zone) %>%
  left_join(rgn_poly@data %>% select(rgn_id, rgn_name), by = 'rgn_id')

knitr::kable(slr_rgn_mean)

write_csv(slr_rgn_mean, file.path(dir_goal, 'output/slr/slr_rgn_pressures.csv'))

```

