---
  title: 'OHIBC: Sea Level Rise Pressure layers prep'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
  toc: true
number_sections: true
theme: cerulean
highlight: haddock
includes: 
  in_header: '~/github/ohibc/src/templates/ohibc_hdr1.html'
pdf_document:
  toc: true
---
  
  ``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = FALSE, message = FALSE, warning = FALSE)

library(sp)        # the classes and methods that make up spatial ops in R
library(rgdal)
library(raster)

dir_git <- '~/github/ohibc'
source(file.path(dir_git, 'src/R/common.R'))
### an OHIBC specific version of common.R
dir_anx <- file.path(dir_neptune_data, 'git-annex/bcprep')
dir_rgn <- file.path(dir_git, 'prep/regions')  ### github: general buffer region shapefiles

### goal specific folders and info
goal      <- 'pressures'
scenario  <- 'v2016'
dir_goal  <- file.path(dir_git, 'prep', goal, scenario)
dir_goal_anx <- file.path(dir_anx, goal, scenario)

### provenance tracking
source('~/github/ohibc/src/R/prov.R')      
### Provenance tracking functions: must source at start to initialize prov_track
dir_prov <- file.path(dir_goal, 'prov') 
### set a provenance folder for this script
this_script_file <- file.path(dir_goal, 'pressures_slr_prep.Rmd') 
### can't recognize the current script name on its own :(
prov_run_tag <- 'standard run'

### goal-specific source scripts
source(file.path(dir_goal, 'pressures_lyr_fxns.R'))

### other support functions
source(file.path(dir_git, 'src/R/rast_tools.R'))

### set up proj4string options: BC Albers and WGS84
p4s_wgs84 <- '+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0'
p4s_bcalb <- '+proj=aea +lat_1=50 +lat_2=58.5 +lat_0=45 +lon_0=-126 +x_0=1000000 +y_0=0 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0'

```

# Summary

There are two parts to creating this layer:
  
  1. Data prep to get raw data into the correct format:
  * If necessary, read .nc.gz files from aviso: ftp://ftp.aviso.altimetry.fr/global/delayed-time/grids/climatology/monthly_mean/
  * unzip .gz files and then delete .gz 
* Read monthly mean sea level data as rasters; aggregate to annual means
* Clip rasters to extents of OHIBC region, and reproject into BC Albers projection.
2. Creating the pressure layers for OHIBC:
  * Determine baseline using five-year average from 1993-1997.
* For each study year (2000 - 2015) calculate difference between study year and baseline
* Set to zero all negative values, indicating decreases in mean sea level
* Resample data from the native cell resolution (0.25 degrees) to ~ 1km
* Set reference point as the 99.99th quantile of the data distribution to rescale all values from 0 to 1
* Use thin plate spline interpolation to fill in NA values
* Clip results to 1 km offshore region, and determine average SLR pressure per region.

This process is completed entirely within this script.

# Data

Reference: The altimeter products were produced and distributed by Aviso (http://www.aviso.altimetry.fr/), as part of the Ssalto ground processing segment. [AVISO MSLA heights, monthly means](http://www.aviso.altimetry.fr/en/data/products/sea-surface-height-products/global/msla-mean-climatology.html)

* Downloaded: March 18, 2016
* Description: Monthly mean sea level anomaly
* Native data resolution: 0.25 degree grid cells
* Time range: January 1993 - August 2015
* Format: NetCDF

# Methods

## Data Prep

### Read raw data as monthly means and convert to annual means

Raw data in monthly mean sea level anomalies, at .25 degree grid globally, is processed into annual mean sea level anomalies.  These files are saved in git-annex as GeoTIFFs for later use.

``` {r download_files_and_unzip}

reload <- FALSE

dir_anx_aviso <- file.path(dir_neptune_data, 'git-annex/globalprep/_raw_data/AVISO_slr')

if(reload) {
  ### Get filenames from AVISO FTP URL:
  library(RCurl)
  url <- "ftp://ftp.aviso.altimetry.fr/global/delayed-time/grids/climatology/monthly_mean/"
  userpwd <- "nceas_ohara:diors54jh"
  filenames <- getURL(url,
                      userpwd = userpwd,
                      ftp.use.epsv = FALSE,
                      dirlistonly = TRUE) %>%
    str_split('\n') %>%
    unlist()
  
  filenames <- filenames[!str_detect(filenames, '.png')] %>%
    sort()
  
  
  ### Set up loop to download each file and save to git-annex:
  
  ftp_dir <- 'ftp://nceas_ohara:diors54jh@ftp.aviso.altimetry.fr/global/delayed-time/grids/climatology/monthly_mean'
  for (nc_file in filenames) { # nc_file <- filenames[2]
    download.file(url      = file.path(ftp_dir, nc_file),
                  destfile = file.path(dir_anx_aviso, 'msla_monthly_mean', nc_file),
                  mode     = 'wb')
  }
  
  zipfiles <- list.files(file.path(dir_anx_aviso, 'msla_monthly_mean'),
                         full.names = TRUE)
  zipfiles <- zipfiles[str_detect(zipfiles, '.gz')]
  for (zipfile in zipfiles) { # zipfile <- zipfiles[1]
    message('Unzipping file: ', zipfile)
    R.utils::gunzip(zipfile, remove = TRUE, skip = TRUE)
  }
}
```

``` {r msla_monthly_to_annual}

library(ncdf4)
nc_files <- list.files(file.path(dir_anx_aviso, 'msla_monthly_mean'),
                       full.names = TRUE)
nc_files <- nc_files[str_detect(nc_files, '.nc') & !str_detect(nc_files, '.gz')]

years <- unique(basename(nc_files) %>% 
                  str_replace('dt_global_allsat_msla_h_y', '') %>%
                  substr(1, 4)) %>%
  sort()

for (j in 1:length(years)) { ### j <- 1
  ### get the list of netCDF files associated with the appropriate year
  msla_yr <- nc_files[str_detect(nc_files, years[j])]
  ### set up an empty list to store rasters for this year
  rast_list_yr <- vector('list', length(msla_yr))
  
  message('Generating annual MSLA raster for ', years[j])
  
  ### Loop over all files for this year, read raster and assign to list for this year
  for (i in 1:length(msla_yr)) { ### i = 1
    rast_list_yr[i] <- raster(msla_yr[i], nogit = TRUE) %>%
      raster::rotate()
  }
  
  ### brick all rasters for this year, and calc annual mean, then write as raster
  rbrick_yr <- brick(rast_list_yr)
  rast_annual_mean <- calc(rbrick_yr, mean)
  
  rast_file <- file.path(dir_anx_aviso, 'msla_annual_mean', 
                         sprintf('rast_msla_annual_%s.tif', years[j]))
  message('Writing raster to ', rast_file)
  writeRaster(rast_annual_mean, 
              rast_file,
              overwrite = TRUE,
              nogit = TRUE)
}

```

For analysis, convert the annual means into a raster brick with each layer representing a year.

``` {r }
rast_files <- file.path(dir_anx_aviso, 'msla_annual_mean', 
                        sprintf('rast_msla_annual_%s.tif', years))

rast_list <- vector('list', length(rast_files))
for (i in 1:length(rast_files)) {
  message('Reading raster from ', rast_files[i])
  rast_list[[i]] <- raster(rast_files[i])
}

rbrick_sla <- brick(rast_list)
names(rbrick_sla) <- years

range_sla <- range(values(rbrick_sla), na.rm = TRUE)
```

Range of annual mean sea level anomalies for global dataset: `r paste(range_sla, collapse = ' - ')` meters.


### Convert global data to OHIBC region data

The global data, in WGS84 coordinate ref system, is clipped down to a bounding box around the OHIBC region, and reprojected into BC Albers projection.  This data is then resampled to 1 km resolution using nearest neighbor method.

The resulting raster layers include annual mean sea level anomaly for each coarse cell (approx .25 degree, from original resolution) for each year (1993-2015).

``` {r load_rgn, echo = FALSE}

rgn_lyr <- 'ohibc_rgn'

message(sprintf('Reading OHIBC regions shapefile...\n  %s/%s.shp', dir_rgn, rgn_lyr))
rgn_poly <- readOGR(dsn = path.expand(dir_rgn), layer = rgn_lyr,
                    verbose = FALSE, stringsAsFactors = FALSE)

rgn_poly_wgs84 <- readOGR(dsn = path.expand(dir_rgn), 
                          layer = paste(rgn_lyr, '_wgs84', sep = ''),
                          verbose = FALSE, stringsAsFactors = FALSE)

### In case we need to resample to 1 km... use base raster from 
### regions folder (note: 500m base raster available too).  Used later in
### finding zonal statistics anyway.
rast_rgn <- raster(file.path(dir_rgn, 'ohibc_rgn_raster_1000m.tif'))

```

``` {r crop_sla_brick, echo = FALSE}
### clip rasters to extents of region (in WGS 84 CRS)
### then: reproject to BC Albers projection
bc_sla_file <- file.path(dir_goal, 'int/slr/rast_sla_raw.tif')

if(!file.exists(bc_sla_file)) {
  bc_sla <- raster::crop(rbrick_sla, rgn_poly_wgs84, snap = 'out') %>%
    raster::projectRaster(crs = p4s_bcalb)
  
  plot_rast(bc_sla[[1]], 
            title = sprintf('Mean SL anomaly %s', names(bc_sla)[1]), 
            scale_label = 'meters',
            rev_scale = TRUE)
  plot_rast(bc_sla[[nlayers(bc_sla)]], 
            title = sprintf('Mean SL anomaly %s', names(bc_sla)[nlayers(bc_sla)]), 
            scale_label = 'meters',
            rev_scale = TRUE)
  
  message('Resampling rasters to new resolution')
  bc_sla_resamp <- resample_brick(bc_sla, rast_rgn,
                                  filename = bc_sla_file,
                                  overwrite = TRUE)
} else {
  bc_sla_resamp <- brick(bc_sla_file)
  names(bc_sla_resamp) <- paste('y', years, sep = '')
}

plot_rast(bc_sla_resamp[[1]], 
          title = sprintf('Mean SL anomaly %s', names(bc_sla_resamp)[1]), 
          scale_label = 'meters',
          rev_scale = TRUE)
plot_rast(bc_sla_resamp[[nlayers(bc_sla_resamp)]], 
          title = sprintf('Mean SL anomaly %s', names(bc_sla_resamp)[nlayers(bc_sla_resamp)]), 
          scale_label = 'meters',
          rev_scale = TRUE)

```


## Rescale values as pressures

Baseline sea level anomaly is based upon the count within 1993 to 1997.

Change in mean sea level is determined by simple subtraction of the 1998-2014 rasters from the 1993-1997 mean raster.

``` {r calc_delta_sl_anomalies}

base_yrs <- 5
names(bc_sla_resamp)[1:base_yrs]
bc_sla_baseline <- bc_sla_resamp[[1:base_yrs]] %>%
  calc(mean)

bc_sla_study <- bc_sla_resamp[[(base_yrs + 1):nlayers(bc_sla_resamp)]]

bc_sla_delta <- bc_sla_study - bc_sla_baseline
names(bc_sla_delta) <- names(bc_sla_study)

bc_sla_delta_file <- file.path(dir_goal, 'int/slr/rast_sla_delta_raw.tif')
message('Writing BC SLA delta to: ', bc_sla_delta_file)
writeRaster(bc_sla_delta, bc_sla_delta_file, bandorder='BIL', overwrite = TRUE)

plot_rast(bc_sla_delta[[1]], 
          title = sprintf('delta SL anomaly %s', names(bc_sla_delta)[1]), 
          scale_label = 'meters',
          rev_scale = TRUE)
plot_rast(bc_sla_delta[[nlayers(bc_sla_delta)]], 
          title = sprintf('delta SL anomaly %s', names(bc_sla_delta)[nlayers(bc_sla_delta)]), 
          scale_label = 'meters',
          rev_scale = TRUE)

```

Interpolate change in sea level values, to fill gaps.  Use thin-plate spline method to interpolate.

``` {r interpolate_bc_sla}
### sea level rise = delta (sea level anomaly) so slr <- d(sla)
bc_slr_int_file <- file.path(dir_goal, 'int/slr/rast_slr_int.tif')

if(!file.exists(bc_slr_int_file)) {
  message('Interpolating BC SL rise data')
  bc_slr_int <- interpolate_brick(bc_sla_delta, subset_n = 2000)
  
  ### mask to regions
  values(bc_slr_int)[is.na(values(rast_rgn))] <- NA
  
  message('Writing interpolated BC SLA delta to: ', bc_slr_int_file)
  writeRaster(bc_slr_int, bc_slr_int_file, bandorder='BIL', overwrite = TRUE)
  
} else {
  message('Reading interpolated BC SLA delta from: ', bc_slr_int_file)
  
  bc_slr_int <- brick(bc_slr_int_file)
  names(bc_slr_int) <- names(bc_sla_delta)
}

plot_rast(bc_slr_int[[1]], 
          title = sprintf('delta SL anomaly %s', names(bc_slr_int)[1]), 
          scale_label = 'meters',
          rev_scale = TRUE)
plot_rast(bc_slr_int[[nlayers(bc_slr_int)]], 
          title = sprintf('delta SL anomaly %s', names(bc_slr_int)[nlayers(bc_slr_int)]), 
          scale_label = 'meters',
          rev_scale = TRUE)

```

Values below zero (drop in sea level) are set to zero.

This difference value is rescaled from zero (no sea level rise) to 1 (at 99.99%ile of local slr).


``` {r rescale_uv_pressures}

### If SLR velocity is negative, set to zero
bc_slr_pos <- bc_slr_int
values(bc_slr_pos)[values(bc_slr_pos) < 0] <- 0

plot_rast(bc_slr_pos, 
          title = 'SLR pos velocity', 
          scale_label = 'mm/yr',
          rev_scale = TRUE)

ref_pt = quantile(bc_slr_pos, prob=0.9999)

bc_slr_resc <- calc(bc_slr_pos, fun = function(x) ifelse(x > ref_pt, 1, x/ref_pt))
plot_rast(bc_slr_resc, 
          title = 'rescaled SLR velocity', 
          scale_label = 'scaled pressure',
          rev_scale = TRUE)

writeRaster(bc_slr_resc, file.path(dir_goal, 'output/slr/slr_rescaled.tif'), 
            overwrite = TRUE)

```


## Calculate mean pressure by region by year

Using 1000 m region ID raster to run zonal statistics on the pressures layers, calculate the regional average pressures for the study period, and save as a .csv.

``` {r calc_mean_pressures_by_year}

slr_rgn_mean <- zonal(bc_slr_resc, rast_rgn, fun = 'mean') %>%
  as.data.frame() %>%
  rename(pressure = mean, rgn_id = zone) %>%
  left_join(rgn_poly@data %>% select(rgn_id, rgn_name), by = 'rgn_id')

knitr::kable(slr_rgn_mean)

write_csv(slr_rgn_mean, file.path(dir_goal, 'output/slr/slr_rgn_pressures.csv'))

```

-----
  
  ``` {r child = file.path(dir_git, 'src/templates/ohibc_prov_ftr.Rmd')}
```