---
title: "OHI 2017 - Artisanal Opportunities: Preparing need data"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '../../../src/templates/ohi_hdr.html'
pdf_document:
  toc: true
---


# Summary

This script generates the "need" layer for the artisanal opportunities goal.  The "access" layer, which is not updated due to a lack of a data source, is located here: globalprep/res_mora_ao/v2013/data/r_mora_s4_2013a.csv.


## Updates from previous assessment
None

***
#Data Source 

**Downloaded**: 8/17/2017

**Description**:  
GDP adjusted per capita by PPP (ppppcgdp)
http://data.worldbank.org/indicator/NY.GDP.PCAP.PP.KD
Reported at country scale.

GDP per capita based on purchasing power parity (PPP). PPP GDP is gross domestic product converted to international dollars using purchasing power parity rates. An international dollar has the same purchasing power over GDP as the U.S. dollar has in the United States. GDP at purchaser's prices is the sum of gross value added by all resident producers in the economy plus any product taxes and minus any subsidies not included in the value of the products. It is calculated without making deductions for depreciation of fabricated assets or for depletion and degradation of natural resources. Data are in constant international dollars based on the 2011 ICP round.

**Time range**: 1990-2016


***
# Methods

## Setup

``` {r setup, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/', message = FALSE, warning = FALSE)

#setting up provenance
# devtools::install_github('oharac/provRmd')
# library(provRmd)
# prov_setup()

library(ohicore) # devtools::install_github('ohi-science/ohicore@dev')
library(dplyr)
library(tidyr)
library(WDI) # install.packages('WDI')
library(stringr)
library(readr)

# comment out when knitting:
# setwd("globalprep/ao/v2017")

# directory paths
source('../../../src/R/common.R')

```

## Download and save data

```{r load_data,eval=F}

# check website to see what years are available
yr_start = 1990
yr_end   = 2016


## get description of variables (NOTE: these descriptions appear out of date, they aren't in sync with the definitions of the World Bank):
indicators <-  data.frame(WDI_data[[1]])
#indicators[grep("NY.GDP.PCAP.PP.CD", indicators$indicator), ]  # current dollars (influenced by inflation, not being used)
indicators[grep("NY.GDP.PCAP.PP.KD", indicators$indicator), ]  # constant dollars 

# download the data
gdppcppp = WDI(country = "all",
               indicator = "NY.GDP.PCAP.PP.KD", 
               start = yr_start, end=yr_end)

write.csv(gdppcppp, 'raw/raw_gdppcppp.csv', row.names=FALSE)

```


## Within country gapfilling

When a country has some values, a within country regression model is used to predict the missing values.

```{r}


d <- read.csv('raw/raw_gdppcppp.csv') %>%
  select(country, value=NY.GDP.PCAP.PP.KD, year) %>%
  filter(year >= 2005) %>%
  spread(year, value) %>%  ## fill in potentially missing values with NA
  data.frame() %>%
  gather(year, value, starts_with("X")) %>%
  mutate(year = gsub("X", "", year)) %>%
  mutate(year = as.numeric(year))

## for the first case, if there is only one value use this value for all years
## this is not ideal, but likely better than other forms of gapfilling

d <- d %>%
  group_by(country) %>%
  mutate(value_num = sum(!is.na(value))) %>%
  filter(value_num > 0) %>%    # filter out the countries with no data between 2005 and 2015   
  mutate(value_num_gf = ifelse(value_num==1, mean(value, na.rm=TRUE), NA)) %>%
  ungroup() %>%
  mutate(value = ifelse(is.na(value), value_num_gf, value)) %>%
  select(country, year, value, value_num)

d_gf <- d %>%
  group_by(country) %>%
  do({
    mod <- lm(value ~ year, data=.)
    value_pred <- predict(mod, newdata=.[c('year')])
    data.frame(., value_pred)
  })

d_gf <- d_gf %>%
  ungroup() %>%
  mutate(gapfilled = ifelse(is.na(value), 1, 0)) %>%
  mutate(gapfilled = ifelse(value_num == 1, 1, gapfilled)) %>%
  mutate(value = ifelse(is.na(value), value_pred, value)) %>%
  mutate(method = ifelse(gapfilled==1, paste("lm based on N years data:", value_num, sep=" "), NA)) %>%
  mutate(method = ifelse(value_num == 1, "gapfilled using one year of data", method))
  
```

## Calculate standardized values
This is performed by taking the natural log of each value and then dividing by the 95th quantile of values across all years (from 2005 to 2016). 

```{r}

d_stand <- d_gf %>%
  mutate(quantile_95 = quantile(value, probs=0.95)) %>%
  mutate(value_stand = value/quantile_95) %>%
  mutate(value_stand = ifelse(value_stand > 1, 1, value_stand))

d_stand <- d_stand %>%
  select(country, year, value, score=value_stand, gapfilled, method)

plotData <- d_stand %>%
  mutate(need = 1-score) %>%
  dplyr::select(country, year, need)

# library(googleVis)
#
# Motion = gvisMotionChart(plotData,
# idvar="country", timevar="year")
# 
#  plot(Motion)
# 
#  print(Motion, file = file.path('ao_need_95q.html'))

```


# Convert country names to ohi regions

```{r}

### Function to convert to OHI region ID
d_stand_rgn <- name_2_rgn(df_in = d_stand, 
                       fld_name='country', 
                       flds_unique=c('year'))


### Combine the duplicate regions (we report these at lower resolution)
### In this case, we take the average score weighted by population
population_weights <- read.csv('../../../../ohiprep/src/LookupTables/Pop_weight_ChinaSAR_USVIslPRico.csv')

d_stand_rgn <- d_stand_rgn %>%
  left_join(population_weights, by="country") %>%
  mutate(population = ifelse(is.na(population), 1, population)) %>%
  group_by(rgn_id, year) %>%
  summarize(score = weighted.mean(score, population),
            value = weighted.mean(value, population),
            gapfilled = weighted.mean(gapfilled, population),
            method = method[1]) %>%
  ungroup() 

d_stand_rgn <- d_stand_rgn %>%
  filter(rgn_id <= 250)

summary(d_stand_rgn)

## save the cleaned gdppcppp for other goals
gdppcppp_data <- d_stand_rgn %>%
  select(rgn_id, year, value)

write_csv(gdppcppp_data, "intermediate/gdppcppp_ohi.csv")

```

# Gapfilling: part 2
In this case, we gapfill regions with no data using UN geopolitical means.

```{r}

d_stand_gf <- data.frame(year=min(d_stand_rgn$year):max(d_stand_rgn$year)) %>% 
  merge(georegions, by=NULL) 

d_stand_gf <- d_stand_gf %>%  
  left_join(d_stand_rgn, by = c("rgn_id", "year")) %>%
  group_by(r2, year) %>%
  mutate(r2_value = mean(score, na.rm=TRUE)) %>%
  ungroup() %>%
  group_by(r1, year) %>%
  mutate(r1_value = mean(score, na.rm=TRUE)) %>%
  ungroup() %>%
  group_by(r0, year) %>%
  mutate(r0_value = mean(score, na.rm=TRUE)) %>%
  ungroup()

d_stand_gf <- d_stand_gf %>%
  mutate(gapfilled = ifelse(is.na(score) & !is.na(r2_value), "1", gapfilled)) %>%
  mutate(method = ifelse(is.na(score) & !is.na(r2_value), "UN_geopolitical region avg, r2", method)) %>%
  mutate(score = ifelse(is.na(score), r2_value, score)) %>%
  mutate(gapfilled = ifelse(is.na(score) & !is.na(r1_value), "1", gapfilled)) %>%
  mutate(method = ifelse(is.na(score) & !is.na(r1_value), "UN_geopolitical region avg, r1", method)) %>%
  mutate(score = ifelse(is.na(score), r1_value, score)) %>%
  mutate(gapfilled = ifelse(is.na(score) & !is.na(r0_value), "1", gapfilled)) %>%
  mutate(method = ifelse(is.na(score) & !is.na(r0_value), "UN_geopolitical region avg, r0", method)) %>%
  mutate(score = ifelse(is.na(score), r0_value, score))

```


# Save the data

```{r}

final <- d_stand_gf %>%
  select(rgn_id, year, value=score)

write.csv(final, "output/wb_gdppcppp_rescaled.csv", row.names=FALSE)

final_gf <- d_stand_gf %>%
  select(rgn_id, year, gapfilled, method)

write.csv(final_gf, "output/wb_gdppcppp_rescaled_gf.csv", row.names=FALSE)


```
