SPP: IUCN Species 2014 vs 2013 Report
========================================================

Versions:
* **2013**: _**2013.1**_ (as of Jul 1,  2013)
* **2014**: _**2013.2**_ (as of Apr 20, 2014)


```{r setup, echo=FALSE}
suppressPackageStartupMessages(library(dplyr))
library(knitr)
options(knitr.table.format='markdown')

# because [kable chokes on table with NA](https://github.com/yihui/knitr/issues/720)
# library(devtools)
# install_github('yihui/knitr')

setwd('~/ohiprep') # specific to host, eg neptune

# directories
source('src/R/common.R')
# working dir in local github repo
nm = 'Global/NCEAS-SpeciesDiversity_v2014'
wd = file.path(getwd(), nm)
# data dir
dd    = file.path(dir_neptune_data, 'git-annex', nm)
dd.13 = file.path(dir_neptune_data, 'model/GL-NCEAS-SpeciesDiversity_v2013a')

# data
spp_iucn_all       = read.csv(file.path(dd   , 'cache', 'spp_iucn_all.csv'   ), na.strings='')
spp_iucn_marine    = read.csv(file.path(dd   , 'cache', 'spp_iucn_marine.csv'), na.strings='')
spp_iucn_all.13    = read.csv(file.path(dd.13, 'tmp'  , 'spp_iucn_all.csv'   ), na.strings='')
spp_iucn_marine.13 = read.csv(file.path(dd.13, 'tmp'  , 'spp_iucn_marine.csv'), na.strings='')
```


```{r sid dupes, echo=FALSE, eval=FALSE}
## Check for Duplicates
# Duplicates detected for 2014. Checking if existed in outputs from 2013.

names(spp_iucn_all)
table(duplicated(spp_iucn_all$Red.List.Species.ID))

names(spp_iucn_all.13)
table(duplicated(spp_iucn_all.13$Red.List.Species.ID))

dim(spp_iucn_marine)
dim(spp_iucn_marine.13)

# Ok, confirmed that use of `Primary` to identify non-redundant rows (presumably used for species synonyms) is new to IUCN 2014 data.
```

## extinction categories

Totals compared between years.
```{r cat_, results='asis', echo=FALSE, dependson='setup'}
d = spp_iucn_marine.13 %.%
  group_by(category) %.%
  summarize(freq_2013=n()) %.%
  merge(spp_iucn_marine %.%
          group_by(category) %.%
          summarise(freq_2014 = n()),
        by='category') %.%
  mutate(freq_dif = freq_2014 - freq_2013)
# add marginal totals row
d = suppressWarnings(rbind(d, cbind(category='Total', as.data.frame(t(colSums(d[,-1]))))))
kable(format(d), row.names=F)
```


## species

### species new to 2014 by species id (sid)
Looking first at a summary of species counts by `order` and `modified.year` new to 2014 without matching field `sid` in 2013.
```{r cat_sid, results='asis', echo=FALSE, dependson='setup'}
d = spp_iucn_marine %.%
  anti_join(spp_iucn_marine.13, by='sid') %.%
  select(sid, sciname, category, criteria, modified.year, class, order, family) %.%
  group_by(class, order, modified.year) %.%
  summarize(count=n())
# add marginal totals row
d = suppressWarnings(rbind_list(d, data.frame(class='Total', count=sum(d$count))))
kable(format(d), row.names=F)
```

### species new to 2014 by scientific name (sciname)
But the existence of a new species id (`sid`) seems to most likely be attribution to a newly identified subpopulation, since when looking for unmatched scientific names (`sciname`) present in 2014 and missing in 2013, there are the only a few new species to 2014.
```{r anti_14not13_sciname, results='asis', echo=FALSE, dependson='setup'}
d = spp_iucn_marine %.%
  anti_join(spp_iucn_marine.13, by='sciname') %.%
  select(sid, sciname, class, order, family, modified.year, category)
kable(d, row.names=F)
```

### species missing in 2014 but present in 2013, by scientific name (sciname)
Conversely, here are the species present in 2013 but missing from 2014 by matching field `sciname`.
```{r anti_13not14, results='asis', echo=FALSE, dependson='setup'}
d = spp_iucn_marine.13 %.%
  anti_join(spp_iucn_marine, by='sciname') %.%
  select(sid, sciname, class, order, family, modified.year, category)
kable(d, row.names=F)
```

Of those 2013 species missing in 2014, here are the same species ids (field `sid`) still found in 2014. So the `sciname` changed.
```{r anti_13not14_lu, results='asis', echo=FALSE, dependson='anti_13not14'}
dd = spp_iucn_marine %.%
  filter(sid %in% d$sid) %.%
  select(sid, sciname, class, order, family, modified.year, category)
kable(d, row.names=F)
```

### species with extinction changed from 2013 to 2014, by species id (sid)

Species with extinction category changed by matching `sid`.
```{r cat-chg_by-sid, results='asis', echo=FALSE, dependson='setup'}
d = spp_iucn_marine %.%
  inner_join(
    spp_iucn_marine.13 %.%
      select(sid                = sid,
             sciname_2013       = sciname,
             modified.year_2013 = modified.year, 
             category_2013      = category, 
             criteria_2013      = criteria),
    by='sid') %.%
  select(sid, sciname, sciname_2013, category, category_2013, criteria, criteria_2013, modified.year, modified.year_2013, class, order, family) %.%
  filter(category!=category_2013)
kable(d, row.names=F)
```

### species with extinction changed from 2013 to 2014, by scientific name (sciname)

```{r ex_cat, results='asis', echo=FALSE, dependson='setup'}
d = spp_iucn_marine %.%
  inner_join(
    spp_iucn_marine.13 %.%
      select(sciname            = sciname,
             sid_2013           = sid,             
             modified.year_2013 = modified.year, 
             category_2013      = category, 
             criteria_2013      = criteria),
    by='sciname') %.%
  select(sid, sid_2013, sciname, category, category_2013, 
         criteria, criteria_2013, modified.year, modified.year_2013, 
         class, order, family) %.%
  filter(category!=category_2013)
cat(sprintf('Number of rows: %d\n\nFirst 6 rows...\n', nrow(d)))
kable(format(head(d)), row.names=F)
```

So scientific name (`sciname`) can repeat for a given `sid`, indicating a subpopulation, in which case matching by `sciname` alone is insufficient.