---
title: 'Exploring new OA data: OHI 2016'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: ../../../src/templates/ohi_hdr.html 
  pdf_document:
    toc: true
---

For OHI 2015 we used data from [Feely et al. (2009)](http://www.tos.org/oceanography/archive/22-4_feely.pdf) but updated to 2014 and provided by WHOI.

In 2015 a [new paper](https://www.researchgate.net/publication/283938912_Climatological_distribution_of_aragonite_saturation_state_in_the_global_oceans) came out with global aragonite saturation state at multiple depths. Data is publicly available in NetCDF format. This data claims to be an improvement on Feely et al. 2009. 

Here is an initial exploration to see if it should be used instead of the 2014 data.

```{r setup}
source('src/R/common.R')

library(raster)

data = stack(file.path(dir_neptune_data,'git-annex/globalprep/_raw_data/NOAA_OA/Aragonite.nc'),varname="Aragonite")
data
```

There are 9 layers in the file. One for each depth (0,50,100,200, 500, 1000, 2000, 3000, and 4000 meters).

**Surface Aragonite Saturation State**

```{r surface}

sur = data[[1]]%>%
        t(.) #transpose 

```

Off the bat I don't think there is any temporal information for this data....

